---
title: "01: Intro Stats Review"
---


```{r}
#| echo: false


library(tidyverse)
```

Researchers are interested in whether how much time young adults spend sitting each day (reading, watching TV, using a computer, commuting) is related to symptoms of depression. To study this, they used data from the National Health and Nutrition Examination Survey (NHANES), an ongoing survey run by the CDC that collects health and behavioral data from a nationally representative, randomly selected sample of the U.S. population.

Using data from the NHANES survey, participants aged 18-25 were classified into two groups: those who reported **more** sitting time than average ("High"), and those who reported **less** ("Low"). Their depression symptoms were measured with the PHQ-9, a standard 9-item screening scale (0-27, higher = more symptoms).

```{r}
#| echo: false 
data <- read_csv(here::here("data/nhanes-sitting-depression.csv"))
median_hrs <- median(data$sedentary_hours)

day1_data <- data |>
  mutate(sitting_group = if_else(sedentary_hours > median_hrs, "High", "Low")) |>
  select(sitting_group, phq9)

day1_data |>
  group_by(sitting_group) |>
  summarize(n = n(),
            xbar = mean(phq9),
            sd = sd(phq9)) |>
  knitr::kable(digits = 2)
```


1.  What is the null hypothesis for this test?

::: {.content-visible when-format="html"}

::: {.callout-tip collapse="true"}

### Solution

$H_0: \mu_1 = \mu_2$ (or $H_0: \mu_1 - \mu_2 = 0$)

:::

:::


\vspace{1in}

2.  What is the alternative hypothesis for this test?


::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$H_A: \mu_1 \neq \mu_2$ (two-sided)

:::
:::

\vspace{1in}

3.  The researchers conduct a two-sample t-test and report a test statistic of $t=1.26$. Below is a plot of the appropriate t-distribution for this test. Sketch how you would calculate the p-value for this situation.

```{r}
#| echo: false
#| fig.height: 2
#| fig.width: 4
#| fig.align: center
#| fig-alt: "A t-distribution density curve, with t on the x-axis and density on the y-axis."
data.frame(x = seq(-4, 4, length = 100)) |>
  ggplot(aes(x = x)) +
  stat_function(fun = dt, args = list(df = 20), linewidth = 1) +
  labs(y = "Density", x = "t") +
  theme_classic() +
  theme(text = element_text(size=12))
```


::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

The p-value is equal to the area in the two tails beyond $t=1.26$ and $-t=1.26$.

```{r}
#| echo: false
#| fig.height: 3
#| fig.width: 6
#| fig.align: center
#| fig-alt: "A t-distribution density curve, with the two tail regions  shaded."
data.frame(x = seq(-4, 4, length = 100)) |>
  ggplot(aes(x = x)) +
  stat_function(fun = dt, args = list(df = 20), linewidth = 1) +
  stat_function(fun = dt, args = list(df = 20),
                xlim = c(-4,-1.26),
                geom = "area", fill = "blue", alpha = 0.6) +
stat_function(fun = dt, args = list(df = 20),
                xlim = c(1.26, 4),
                geom = "area", fill = "blue", alpha = 0.6) +
  labs(y = "Density", x = "t") +
  theme_classic() +
  theme(text = element_text(size=12))
```

:::
:::


4.  How did you calculate this p-value in your last statistics course?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

Answers will vary based on where you took intro stats. You may have used StatKey or a similar online app/calculator, you may have used R's `pt()` command, or you may have used a t-distribution table (we won't use tables in this class).
:::
:::

\vspace{1in}


5.  If the p-value you calculated was 0.26, what would you conclude about the null hypothesis?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

A p-value of 0.26 does not provide evidence against the null hypothesis. In other words, there is no statistically discernible evidence of a difference in means between the two groups ($t=1.26$, p-value = 0.26).

:::
:::

\vspace{1in}

6.  Was this an observational study or an experiment? Based on your answer, can you conclude that spending more time sitting *causes* higher depression symptoms? Why or why not?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

This is an observational study — participants were not randomly assigned to a sitting-time group, they were classified based on their reported behavior. Because sitting time was not randomly assigned, other variables (e.g., physical activity, chronic illness, employment status) could differ between the two groups and explain the difference in depression scores. We cannot conclude that sitting time causes a difference in depression symptoms, only that the two are associated.

:::
:::

\vspace{1in}


7.  Can you generalize these results to all young adults in the U.S.? Why or why not?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

Yes, we can generalize these results to all young adults in the U.S. because NHANES is a nationally representative survey.

:::
:::
