---
title: "04: Prediction for SLR"
---

## Warm-up questions

Is education level associated with income? Researchers collected education level (in years) and income (in thousands of dollars) for a random sample of 32 employees working for the city of Riverview.

```{r}
#| include: false
city <- readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/riverview.csv")
lm.a = lm(income ~ 1 + education, data = city)
```

The researchers fit a simple linear regression of income on education and obtained the following output:

```{r}
#| echo: false
#| tbl-align: center
broom::tidy(lm.a) |>
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
knitr::kable(format = "pipe", digits = 3, booktabs = TRUE)
```

**Q1.** Write the fitted regression equation.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q2.** Interpret the slope coefficient in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q3.** Interpret the intercept in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

A new researcher joined the team and decided that education should be standardized (to have mean 0 and SD 1) before fitting the regression model. The output from this regression is shown below:

```{r}
#| echo: false
lm.b <- lm(income ~ 1 + scale(education), data = city)
broom::tidy(lm.b) |>
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
knitr::kable(format = "pipe", digits = 3, booktabs = TRUE)
```

**Q4.** Write the fitted regression equation for this new model.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q5.** Interpret the slope coefficient in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q6.** Interpret the intercept in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q7.** Compare the two models. What's the same? What's different?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

## Prediction and confidence intervals

::: {.callout-tip}
The R code for the following questions is found at <br>

<https://aluby.github.io/stat230/activities/04R-prediction>

The URL is also posted on Moodle.
:::

Now, we'll explore the `city` data using R.


**Q1.** Make a plot of `income` against `education`  and overlay the linear regression equation. Does it look like what you expected?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q2.** Make a plot of `income` against `scale(education)` and overlay the linear regression equation. Does it look like what you expected?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q3.** Find the linear regression equation from Q1. Confirm that it matches the table at the beginning of this worksheet.

**Q4.** The first person in this dataset has `education = 8`. Calculate the expected income of this person using the first regression equation.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q5.** If we want to predict the price of **this** person, should we use a confidence interval or a prediction interval?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q6.** Use R to construct the appropriate 89% interval for this person's income. Record this interval below.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q7.** Interpret the interval in context.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}

**Q8.** Run the code to produce a scatterplot, regression line, and both types of intervals. Which is the prediction interval and which is the confidence interval? How can you tell?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

:::
:::

\vspace{1in}
