---
title: "04: Prediction for SLR"
---

## Warm-up questions

Is education level associated with income? Researchers collected education level (in years) and income (in thousands of dollars) for a random sample of 32 employees working for the city of Riverview.

```{r}
#| include: false
city <- readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/riverview.csv")
lm.a = lm(income ~ 1 + education, data = city)
library(ggformula)
```

The researchers fit a simple linear regression of income on education and obtained the following output:

```{r}
#| echo: false
#| tbl-align: center
broom::tidy(lm.a) |>
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
knitr::kable(format = "pipe", digits = 3, booktabs = TRUE)
```

**Q1.** Write the fitted regression equation.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$\widehat{\text{income}} = 11.321 + 2.651 \times \text{education}$

:::
:::

\vspace{1in}

**Q2.** Interpret the slope coefficient in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

For each additional year of education, the model predicts an average increase of \$2,651 (2.651 thousand dollars) in income.

:::
:::

\vspace{1in}

**Q3.** Interpret the intercept in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

For an employee with 0 years of education, the model predicts an average income of \$11,321 (11.321 thousand dollars). Note that this is extrapolation -- education ranges from 8 to 24 years in this sample, so no one has 0 years of education, and this interpretation may not be meaningful in context.

:::
:::

\vspace{1in}

A new researcher joined the team and decided that education should be standardized (to have mean 0 and SD 1) before fitting the regression model. The output from this regression is shown below:

```{r}
#| echo: false
lm.b <- lm(income ~ 1 + scale(education), data = city)
broom::tidy(lm.b) |>
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
knitr::kable(format = "pipe", digits = 3, booktabs = TRUE)
```

**Q4.** Write the fitted regression equation for this new model.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$\widehat{\text{income}} = 53.742 + 11.567 \times \text{scale(education)}$

:::
:::

\vspace{1in}

**Q5.** Interpret the slope coefficient in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

For each one standard deviation increase in education (about 4.36 years), the model predicts an average increase of \$11,567 (11.567 thousand dollars) in income.

:::
:::

\vspace{1in}

**Q6.** Interpret the intercept in the context of the problem. (Don't forget to specify units.)

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

For an employee with average education (16 years, i.e. `scale(education) = 0`), the model predicts an average income of \$53,742 (53.742 thousand dollars). Unlike the intercept in the original model, this is a meaningful value since it corresponds to an employee actually near the center of the observed data.

:::
:::

\vspace{1in}

**Q7.** Compare the two models. What's the same? What's different?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

The two models describe exactly the same relationship -- the p-values for the slope are identical. What differs is the scale of the x-variable, which changes the numeric values of the coefficients and what they mean: the slope goes from "per 1 year of education" (2.651) to "per 1 SD of education" (11.567), and the intercept goes from "predicted income at 0 years of education" (an extrapolation) to "predicted income at average education" (a value within the data).

:::
:::

\vspace{1in}

## Prediction and confidence intervals

::: {.callout-tip}
The R code for the following questions is found at <br>

<https://aluby.github.io/stat230/activities/04R-prediction>

The URL is also posted on Moodle.
:::

Now, we'll explore the `city` data using R.


**Q1.** Make a plot of `income` against `education`  and overlay the linear regression equation. Does it look like what you expected?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

```{r}
#| fig-width: 4
#| fig-height: 3
gf_point(income ~ education, data = city) |>
    gf_lm()
```

Yes -- the scatterplot shows a positive, linear, moderately strong relationship between education and income, consistent with the positive slope from the regression output above.

:::
:::

\vspace{1in}

**Q2.** Make a plot of `income` against `scale(education)` and overlay the linear regression equation. Does it look like what you expected?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

```{r}
#| fig-width: 4
#| fig-height: 3
gf_point(income ~ scale(education), data = city) |>
    gf_lm()
```

Yes -- the plot looks identical in shape to Q1, just with the x-axis relabeled in standard deviations from the mean instead of raw years. Standardizing doesn't change the strength or direction of the relationship, only the scale of the x-axis.

:::
:::

\vspace{1in}

**Q3.** Find the linear regression equation from Q1. Confirm that it matches the table at the beginning of this worksheet.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

```{r}
#| fig-width: 4
#| fig-height: 3

library(broom)
lm.a = lm(income ~ 1 + education, data = city)
tidy(lm.a)
```

:::
:::

**Q4.** The first person in this dataset has `education = 8`. Calculate the expected income of this person using the first regression equation.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$\widehat{\text{income}} = 11.321 + 2.651(8) = 32.531$, i.e. \$32,531.

```{r}
predict(lm.a, newdata = data.frame(education = 8))
```

:::
:::

\vspace{1in}

**Q5.** If we want to predict the income of **this** person, should we use a confidence interval or a prediction interval?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

A prediction interva. We're predicting an individual outcome (this specific person's income), not the average income of all employees with 8 years of education. 

:::
:::

\vspace{1in}

**Q6.** Use R to construct the appropriate 89% interval for this person's income. Record this interval below.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

```{r}
predict(lm.a, newdata = data.frame(education = 8),
 interval = "prediction", level = .89)
```

89% prediction interval: (16.746, 48.318), i.e. (\$16,746, \$48,318).

:::
:::

\vspace{1in}

**Q7.** Interpret the interval in context.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

We are 89% confident that a randomly selected City of Riverview employee with 8 years of education will have an income between \$16,746 and \$48,318.

:::
:::

\vspace{1in}

**Q8.** Run the code to produce a scatterplot, regression line, and both types of intervals. Which is the prediction interval and which is the confidence interval? How can you tell?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

The prediction interval is the wider band; the confidence interval is the narrower band hugging the regression line. The confidence interval only captures uncertainty in the *average* income at a given education level, while the prediction interval also accounts for individual-level variability around that average, which makes it wider.

:::
:::

\vspace{1in}
