---
title: "04R: Prediction for SLR"
editor: source
---

::: callout-note
You can access the .qmd for this activity by clicking the "code" button above and then opening in maize/Rstudio. If you prefer to follow the tutorial version in the browser, you can use <https://aluby.github.io/stat230/activities/04-prediction-tutorial>
:::

## Loading packages

The code chunk below loads the R packages that contain functions we'll use. Make sure to run it before the rest of the code!

```{r}
library(tidyverse)
library(ggformula)
```

## Loading data

The data set is loaded by running the following code chunk:

```{r}
city <- readr::read_csv(url("https://raw.githubusercontent.com/zief0002/modeling/main/data/riverview.csv"))
```

The columns of interest are `education` and `income`.

## Making plots

**Q1.** Make a plot of `income` against `education`  and overlay the linear regression equation. Does it look like what you expected? (I've started the code for you; you will need to overlay the linear regression equation)

```{r}
gf_point(income ~ education, data = city)
```

**Q2.** Make a plot of `income` against `scale(education)` and overlay the linear regression equation. Does it look like what you expected?

```{r}
# Write your graph code here
```


## Fitting a simple linear regression model

**Q3.** Use the `lm()` command to fit the simple linear regression
model where `education` is used to predict `income`. 
Make sure your output matches the output on the worksheet.

```{r}
# Write your lm() command here
```



## Making predictions

The first person in this dataset has `education = 8`. 

**Q4.** Calculate the expected income of this person using the first regression equation. (You may use R or do this by hand. There's a code chunk below for your convenience)

```{r}
# Write your code here
```

**Q5.** If we want to predict the income of **this** person, should we use a confidence interval or a prediction interval?

**Q6.** Use R to construct the appropriate 89% interval for this person's income. 

```{r}
# Write your code here
```

**Q7.** Interpret the interval in context.

## Plotting our intervals

It can be useful to plot your confidence/prediction intervals to communicate your findings to a broader audience. The code below creates a scatterplot of the data, adds the fitted regression line, and then adds 89% prediction intervals.

```{r}
gf_point(income ~ education, data = city) |> 
  gf_lm(interval = "prediction", level = .89) |>
  gf_lm(interval = "confidence", alpha = 0.6, level = .89)
```

**Q8.** Run the code to produce a scatterplot, regression line, and both types of intervals. Which is the prediction interval and which is the confidence interval? How can you tell?

***

## Function quick reference

The following table summarizes the functions we learned today:

| Function | Purpose |
|----|-----------|
| `lm(formula, data)` | Fit a linear model |
| `predict(model, newdata, interval, level, se.fit)` | Make predictions and calculate intervals |
| `gf_point(formula, data) |> gf_lm(interval = "confidence")` | Plot data and add confidence intervals for the mean response|
| `gf_point(formula, data) |> gf_lm(interval = "prediction")` | Plot data and add prediction intervals for the mean response|



