---
title: "05: Conditions"
---

::: callout-note
This is the quarto version of the R activity, designed for opening in maize/RStudio. If you prefer to use the tutorial version, you can use <https://aluby.github.io/stat230/activities/05-conditions-tutorial>
:::

```{r}
library(tidyverse)
library(ggformula)
theme_set(theme_minimal())
```

# Example 1: Making residual plots in R 

Recall the `movies` dataset from the R activity we did on Day 02. (`head()` prints out the first few rows)

```{r}
movies <- read.csv("https://aloy.github.io/stat230-materials/data/movie_scores.csv")
head(movies)
```

On that day, we made the following scatterplot: 

```{r}
gf_point(audience ~ critics, data = movies) |>
  gf_lm()
```

and fit this regression model: 

```{r}
movie_lm <- lm(audience ~ critics, data = movies)
movie_lm
```

## Finding residuals 

In order to make residual plots, we first need to find the residuals. While we can do this by hand for a couple of points, it would be extremely tedious to do it for *every single data point*! Instead, we'll use the `augment()` function from the {broom} R package to created an augmented movie dataset. This takes the original X and Y variables and adds the $\hat{y}$ values (`.fitted`) and residuals (`.resid`) (along with some other columns that we'll use later) 

Run the following code chunk and take a peek at the `movie_aug` dataset. Can you find the: 

- original X variable?
- original Y variable?
- Predictions?
- Residuals?

```{r}
library(broom)
movie_aug <- augment(movie_lm)
movie_aug
```


## Making residual plots

The code below makes a normal QQ plot and a histogram of the residuals. Do you think the normality assumption is satisfied?

```{r}
#| layout-ncol: 2


gf_qq(~.resid, data = movie_aug) |>
  gf_qqline()

gf_histogram(~.resid, data = movie_aug, col = "white")
```

### Checkpoint question

Use the `movie_aug` dataset to make a scatterplot with `.fitted` on the x-axis and `.resid` on the y-axis. List the SLR conditions you can assess with this plot, and whether or not they appear to be satisfied. 

```{r}
# your code here
```

## A shortcut

We will make a **ton** of residual plots this term. A shortcut is to use the `resid_panel` function within the {ggResidpanel} package: 

```{r}
library(ggResidpanel)
resid_panel(movie_lm)
```


### Checkpoint question

Which panel(s) would you look at for each of the following SLR conditions. List as many panels as possible, along with how you would identify any violations:

 - Linearity: 
 
 - Independence: 
 
 - Normality: 
 
 - Equal variance: 

# Example 2: $R^2$ in R

### Checkpoint question

Find the $R^2$ value for the `movie_lm` model using `summary()`. Confirm that it is equal to $r^2$

```{r}
# your code here
```

## Summary

### Checkpoint question

Suppose that we fit a linear regression model to a sample of 1,000 observations and we find evidence in the normal Q-Q plot of a violation of the normality assumption, but no evidence in the residual plot of violations of the linearity and constant variance assumptions and no reason to suspect a violation of the independence assumption. Which of the following inferences are valid, and which are not?:

- p-value from the hypothesis test $H_0: \beta_1 = 0$ vs $H_A: \beta_1 \ne 0$ 
- Confidence interval for $\beta_1$
- Confidence interval for the mean of Y at some X value
- Prediction interval for Y at some X value

## Function quick reference

The following table summarizes the functions we learned today:

| Function | Purpose |
|----|-----------|
| `augment(model)` | Create augmented dataset that includes predictions and residuals |
| `gf_qq(~variable, data)` | Make a qq plot |
| `gf_qq_line()` | Overlay a straight line on the qq plot|
| `resid_panel(model)` | Shortcut for making a panel of residual plots|


