---
title: "18: Tips for Case Studies"
---

```{r}
library(tidyverse)
library(ggformula)
library(ggResidpanel)
library(bayesrules)
library(patchwork)
```

```{r}
mod <- lm(rides ~ temp_actual, data = bikes)
```


# Figure size

In the first few lines of your code chunks, you can specify figure output. 

Default:

```{r}
gf_point(rides ~ temp_actual, data = bikes)
```

with 

```
#| fig-width: 4
#| fig-height: 2

```

at top of chunk:

```{r}
#| fig-width: 4
#| fig-height: 2
gf_point(rides ~ temp_actual, data = bikes)
```

# Choosing a couple of residual plots

Sometimes, you might want to include some residual plots instead of the full panel (For example, you want to mention a weird trend in the index plot but all other residual plots are uninteresting). Specify using `plots` argument. 

```{r}
resid_panel(mod)
```

vs

```{r}
resid_panel(mod, plots = "resid")
```


# Combining graphics

Use the {patchwork} package to save individual plots and put them side by side

```{r}
#| fig-width: 8
#| fig-height: 3
p1 <- gf_point(rides ~ temp_actual, data = bikes)
p2 <- resid_panel(mod, plots = "resid")

p1 + p2
```




# Customize theme

Oh no! The plots above look terrible together because they use different `theme`s. *Note:* setting a theme is different with "regular plots" compared to resid panels

```{r}
p1 <- gf_point(rides ~ temp_actual, data = bikes) + theme_bw()
p2 <- resid_panel(mod, plots = "resid", theme = "bw", title.opt = FALSE) 

p1 + p2
```

Classic theme: 

```{r}
p1 <- gf_point(rides ~ temp_actual, data = bikes) + theme_classic()
p2 <- resid_panel(mod, plots = "resid", theme = "classic", title.opt = FALSE) 

p1 + p2
```

*Note:* I know they show up as different sizes. I am not aware of a workaround besides using `augment()` and making your own residual plots. 

# Tables

The default summary output takes up too much space in a case study report:

```{r}
summary(mod)
```

Instead, use `broom::tidy()` for a more compact output. 

```{r}
broom::tidy(mod)
```

You can also specify confidence intervals: 

```{r}
broom::tidy(mod, conf.int = TRUE)
```

And to make it look EVEN nicer, pass it into `knitr::kable()`:

```{r}
broom::tidy(mod, conf.int = TRUE) |>
  knitr::kable()
```

which has a `digits` option:

```{r}
broom::tidy(mod, conf.int = TRUE) |>
  knitr::kable(digits = 2)
```

# Equations

If you are doing your report in .qmd, you might want to include some math! To include an inline equation, use single `$`: $y = \beta_0 + \beta_1 x_1 + \varepsilon, \varepsilon \sim N(0, \sigma^2)$. 

To give an equation its own line, use double `$$`

$$y = \beta_0 + \beta_1 x_1 + \varepsilon, \varepsilon \sim N(0, \sigma^2)$$

(this typsetting language is called LaTeX and there are a bunch of resources available online) 

- https://latexeditor.lagrida.com/ 
- https://webdemo.myscript.com/views/math/index.html# 

