---
title: "24-case-studies"
format: html
---


```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

```

## Common issues on case studies

- Correct things in an incorrect order
- What goes in an EDA? 
- What types of plots go in the main text? 
- Report appendix vs quarto appendix

## General order

- Intro: 
  - What is the research question framed in general? What background information should the reader have to understand your study? How does this research question translate to a question we can answer with a statistical model? What is the response and (general) predictor variables?
  

- EDA/Data: 
  - What dataset are you using? How many observations/variables? What is the distribution of the response variable? How are the predictors related to the response variable? Is there collinearity or skewness that should be addressed before fitting any models? 

- Results: 
    (a) Model building: what were your candidate models? what were your criteria for choosing a model? (You can report *theoretical* model here)
    (b) Conditions and diagnostics: does everything look OK? Did you need to do any further investigation? 
    (c) Now that you have verified the validity of your model, you can report the *fitted* model equation or the coefficient table
    (d) Interpretations of coefficients (with uncertainty!) or results from tests. Answer your research question here
    
- Discussion: 
  - Reiterate your main result without any technical results
  - Are there any limitations to your analysis? 
  - If you had another week, what could you investigate? If you had another dataset, what could you investigate? 
  
- Report appendix

  - These are graphs that you *directly refer to in the text* but that *aren't central for understanding the results*. 
      - "We investigated scatterplots of all predictors against each other to assess collinearity. Variable $x_1$ and $x_2$ were highly correlated ($r=.99$), and so decided to drop $x_2$ from the model." -- the reader does not need to see the pairs plot in the main text to understand what you did, but may want to skim through to see what you saw. 
  - You may also want to include model statements for models you tested but don't necessarily want to talk about: 
      - "We tested 10 different models (see appendix for model statements) and chose the following based on Adjusted $R^2$ and BIC:"

- Quarto appendix

  - This is your code for all analyses (audience is me)
  - Do not assume the reader has access to this appendix

## Presenting equation

```{r}
#| echo: false


moths <- read.csv("https://aluby.github.io/stat230/data/moths.csv") |>
  mutate(notremoved = Placed - Removed, 
               log_odds_removed = log(Removed / notremoved),
               prop_removed = Removed / Placed, 
               dark = ifelse(Morph=="dark",1,0), 
               distance = Distance, 
               morph = Morph) 


moth_glm <- glm(Removed/Placed ~ morph*distance, weights = Placed,
                data = moths, 
                family = "binomial")
```


My recommendation is to report the *theoretical* equation first, then include a coefficient table: 


After determining that no transformations were needed, the binomial model with an interaction term is given by:

`r equatiomatic::extract_eq(moth_glm, intercept = "beta")`

Based on binned residual plots and a ad-hoc test of overdispersion ($\psi \approx 1$), we determine that binomial logistic regression conditions are satisfied. The estimated coefficients with 90% confidence intervals are given in Table 1. 

```{r}
#| label: tbl-moth-glm
#| tbl-cap: "Coefficient from moth binomial regression model."


broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  select(-c("statistic", "p.value")) |>
  knitr::kable(digits = 2)
```


## Coefficient graph

Alternatively, put coefficient table in the report appendix and include a coefficent graph

```{r}
broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  gf_pointrange(estimate + conf.low + conf.high ~ term) |>
  gf_hline(yintercept = 0, linetype = "dashed", col = "gray") 
```

I prefer to flip the coordinates here so that the x axis labels don't overlap:

```{r}
broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  gf_pointrange(estimate + conf.low + conf.high ~ term) |>
  gf_hline(yintercept = 0, linetype = "dashed", col = "gray") +
  coord_flip()
```

Removing intercept can often make the remaining terms easier to read:

```{r}
broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  filter(term != "(Intercept)") |>
  gf_pointrange(estimate + conf.low + conf.high ~ term) |>
  gf_hline(yintercept = 0, linetype = "dashed", col = "gray") +
  coord_flip()
```


## Examples: 

- [Multiple linear regression example](https://www.causeweb.org/usproc/sites/default/files/usclap/2024-2/usclap%203418%20-%20community%20factors%20on%20violent%20crime%20rates.pdf))

- [Logistic regression example](https://causeweb.org/usproc/sites/default/files/usclap/2025-2/usclap%203657%20-%20predicting%20health%20insurance%20coverage.pdf)

- [Advanced method example](https://causeweb.org/usproc/sites/default/files/usclap/2025-2/usclap%203675%20-%20modeling%20sockeye%20salmon%20populations.pdf)

- [Experiment example](https://causeweb.org/usproc/sites/default/files/usclap/2025-1/usclap%203471.pdf)

- [A logistic regression example](https://causeweb.org/usproc/sites/default/files/usclap/2025-1/usclap%203502.pdf)
