---
title: "Homework 01"
webr:
  packages:
    - dplyr
    - ggformula
    - ggplot2
    - Sleuth3
format: 
  live-html:
    toc: true
engine: knitr
editor: source
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(ggformula)
library(here)
library(broom)
```

Due: Friday of Week 1 at start of class

::: callout-important
## Important! 

I will add problems after each class day. Be sure to check back here after each class meeting for new problems.
:::


# Problem 1: Intro Stat review

A 2021 article in *Pediatrics* ("Baby’s First Years: Design of a Randomized Controlled Trial of Poverty Reduction in the United States") describes the design of a study to measure the impact of poverty reduction on children's early development: 

> A total of 1000 low-income mothers of newborns were enrolled in the study and began receiving a monthly unconditional cash gift for the first several years of their children’s lives. Mothers were randomly assigned to receive either a large monthly cash gift ($333) or a nominal monthly cash gift ($20). All monthly gifts are administered via debit card and can be freely spent with no restrictions. Baby’s First Years aims to answer whether poverty reduction in early childhood (1) improves children’s developmental outcomes and promotes healthier brain functioning, and (2) improves family functioning and better enables parents to support child development. 

   (a) Why was it important to randomize amount of the cash gift in the study?

   (b) Suppose the outcome variable $Y$ in a statistical analysis is a quantitative measure of child brain functioning. We then calculate the mean of the large gift group ($\bar{Y}_{large}$) and the small gift group ($\bar{Y}_{small}$). What would an appropriate hypothesis test be? (You only need to set up the hypotheses statements)

  (c) If the researchers fail to reject the null hypothesis from (b), what type of error might they have made?
  
  (d) How could the researchers decrease the probability that they make the type of error in (c)?
  
# Problem 2: Regression

One factor affecting water availability in Southern California is stream runoff from snowfall.
If runoff could be predicted, engineers, planners, and policy makers could do their jobs more
effectively because they would have an estimate as to how much water is entering the area.

We have access to a data set compares the stream runoff (in acre-feet) of a river near Bishop,
California (due east of San Jose) with snowfall (in inches) at a site in the Sierra Nevada
mountains. For each of the following questions, assume that your audience is city water
planners with moderate statistical training.

```{r}
#| echo: false
runoff <- read.csv(here(here(), "data/runoff.csv"))
```

(a) Below is a scatterplot of stream runoff against snowfall. Comment on the apparent relationship (strength, shape, direction).

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 3


ggplot(runoff, aes(x = Precip, y = Runoff)) + 
  geom_point() + 
  theme_minimal() +
  labs(
    x = "Snowfall (inches)",
    y = "Runoff (acre-feet)"
  )
```


(b) Below is the summary table produced when I fit the regression of the expected runoff given snowfall. Write the fitted regression equation. Be sure to use proper notation.

```{r}
lm(Runoff ~ Precip, data = runoff) |> tidy()
```


(c)  Interpret the slope coefficient in the context of the problem. (Don’t forget to specify
units.)

(d) Interpret the intercept in the context of the problem. (Don’t forget to specify units.)

(e) If one winter the site received 10 inches of snowfall, how much runoff is expected? (Don’t
forget to specify units.)


# Problem 3: Grizzlies

The data below are length and girth measurements (in inches) for a sample of Grizzly bears. 

```{webr}
grizzly <- read.csv("https://aluby.github.io/stat230/data/grizzly.csv")
grizzly
```

(a) Plot the bear's girth (response variable) against its length (predictor variable) and show the linear regression line. Describe the relationship.


```{webr}
#| min-lines: 5
```


(b) Compute and report the correlation coefficient


```{webr}
#| min-lines: 5
```

(c) Fit the linear regression model and report the fitted regression equation

```{webr}
#| min-lines: 5
```

(d) What is the interpretation of the intercept coefficient, in context? Does it make sense to perform this interpretation?

```{webr}
#| min-lines: 5
```

(e) A new bear was observed with a length of 148 inches. What would the model predict for its girth?

```{webr}
#| min-lines: 5
```

# Problem 4: The Hubble Constant

> Edwin Hubble used the power of the Mount Wilson Observatory telescopes to meaure features of nebulae outside the Milky WAy. He was surprised to find a relationship between a nebula's distance from earth and the velocity with which it was going away from the earth.... The apparent statistical relationship between distance and velocity led scientists to consider how such a relatinoship could arise. It was proposed that the universe came into being with a Big Bang, a long time ago. If this were true, then $Y = TX$, where $Y$ is Distance and $X$ is velocity, and $T$ gives the time elapsed since the Big Bang (that is, the age of the universe). 

The `ex0725` data points below are measured distances and recession velocities for 10 clusters of nebulae, much farther from earth than Hubble's data. 

```{webr}
library(Sleuth3)
ex0725
```

(a) Make a scatterplot with `Velocity` as the response and `Distance` as the predictor and describe the relationship

```{webr}
#| min-lines: 5
```

(b) Are the data consistent with Hubble's theory (that the intercept is zero?)

```{webr}
#| min-lines: 5
```

(c) Regardless of your answer to the previous question, fit a simple linear regression model through the origin (i.e., forcing $\beta_0 = 0$) by running the following code. Use this model to provide an estimate of the Hubble constant (the age of the universe)

```{webr}
hub_mod <- lm(Velocity ~ 0 + Distance, data = ex0725)
```

(d) Find a 95% confidence interval for this estimate

```{webr}
#| min-lines: 5
```
