---
title: "Homework 03"
format: html
engine: knitr
editor: source
navbar: false
execute: 
  message: false
  warning: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(ggformula)
library(here)
library(broom)
library(tidyverse)
```

Due: Friday of Week 3 at start of class

There is a [homework template](../files/hw-template.qmd) available if you'd like to use that as a starter file, and copy and paste the code as needed. 

::: callout-important
## Important! 

I will add problems after each class day. Be sure to check back here after each class meeting for new problems.
:::


# Problem 1: Fri

The `child_speaks`^[Source: *Stat2*] data are from a study about whether there is a relationship between the age at which a child first speaks (in months) and his/her score on the Gesell aptitude test taken later in childhood. Higher scores on this test indicate age appropriate levels of motor skills, language, social, and cognitive abilities; while lower scores identify children who may need extra support. 

```{r}
child_speaks <- read_csv("https://aluby.github.io/stat230/data/ChildSpeaks.csv")
```

(a) *Before looking at the data*, do you expect to see a positive, negative, or no relationship between these variables? Justify your answer in 1 sentence. 
(b) Produce a scatterplot of these data, and comment on whether age of first speaking appears to be a good explainer of Gesell aptitude score

(c) Fit and report the linear regression equation and the value of $R^2$ (with an interpretation). 

(d) Which child has the largest (in absolute value) residual? Explain what is unusual about that child

(e) Remove the child you identified in (d) and repeat parts (b) and (c). Comment on how removing the one child has affected the line and the value of $R^2$
  
(f) Now, go back to the original dataset. Instead of removing the large residual child, remove the child who took 42 months to speak. Repeat parts (b) and (c) and comment on how removing the one child has affected the line and the value of $R^2$. 

::: callout-note

Here is some example code for creating a filtered dataset. This removes the 10th row from the dataset. You'll want to edit it to filter out the child you identified in (d).

```{r}
library(tidyverse)
child_speaks_filtered <- child_speaks |> slice(-10)
```
:::

# Problem 2: Fri

Return to Problem 8's data on metabolic rate from HW 2:

> The following data comes from *Statistical Sleuth* Chapter 8. The data are the average mass, metabolic rate, and lifespan for 95 species of mammals. [Kleiber’s law](https://en.wikipedia.org/wiki/Kleiber%27s_law) states that the metabolic rate of an animal species, on average, is proportional to its mass raised to the power of 3/4.

```{r}
animals <- Sleuth3::ex0826
head(animals)
```

(a) Review the original SLR model for metabolic rate against mass you found in HW2, and find the $R^2$ value. 

(b) Fit the SLR model for log(metabolic rate) against log(mass). Report the model equation and the $R^2$ value. 

(c) Check the SLR conditions for the model in (b) 

(d) Which model is better: the original model from HW 2 or the log-log model from part (b)? 

# Problem 3: Wed

Statistics students completed a class project where they designed an experiment to explore the relationship between the wing length (in cm) and flight time (in seconds) of paper helicopters. One group collected data from six different wing lengths.

```{r}
flights <- read.csv(here::here("data/winglengths2.csv"))
```

(a) Create a scatterplot of flight time against wing length. Describe what you learn from this plot.

(b) Fit a simple linear regression model of flight time against wing length. Create residual plots to explore the adequacy of your model. Report your results.

(c) Try several transformations of either the response or the explanatory variable to see if you can alleviate the problems you observed in part (b). Report the “best” transformation and whether it remedies the issues.

(d) It is likely that you cannot successfully find a transformation to solve the nonlinearity problem. A closer look at the residuals helps to explain why. The residuals from the regression follow a somewhat sinusoidal pattern: down, then up, then down again. This is a pattern that is seen in a typical third-degree polynomial. Create a scatterplot of flight time against wing length with a third degree polynomial overlayed. Does this look like a reasonable description of the data?

(e) Fit a third degree polynomial regression model and report the fitted model equation.

(f) Create residual plots to explore the adequacy of your model. Report your results.

(g) Using visual inspection, what wing length will lead to the longest flight time?


# Problem 4/5: Case Study 

*Note:* This problem is worth 16 points (4 points each) and will be graded individually but mostly on a good-faith effort. This is to help ensure you've made progress on the case study before the (group) due date on Sunday. 

(a) EDA: include 1 graph
(b) Fit and report an SLR model
(c) One type of inference for a regression coefficient and/or prediction
(d) At least one model evaluation (residual plots and/or $R^2$) with an interpretation
