---
title: "Homework 09"
format: html
engine: knitr
editor: source
navbar: false
execute: 
  message: false
  warning: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(ggformula)
library(here)
library(broom)
library(tidyverse)

theme_set(theme_minimal())
```

Due: Monday of Week 10 at start of class

There is a [homework template](../files/hw-template.qmd) available if you'd like to use that as a starter file, and copy and paste the code as needed.


# Problem 1: `possums`

```{r}
possums <- read.csv("https://aluby.github.io/stat230/data/possums.csv", header = TRUE)
```

**1.** Make a plot of `y` vs. `Bark` and argue that there is a need for a transformation of **at least one of the response and/or predictor**. 

::: callout-note
I recommend that you jitter the value of `y` so that you can more easily see the density of points. Replace `gf_point` with `gf_jitter`. You can vary the amount of horizontal "jitter" with `width = .25` and the amount of vertical jitter with `height = .25` (and replace .25 with whatever number you want!) 
:::

**2.** The variable `y` can take the value 0, so we cannot take its logarithm. One possibility for transformation is to add 1 to the value before taking the logarithm. Plot `log(y+1)` vs. `Bark` and fit the corresponding log-normal linear model. Interpret the estimated slope coefficient.


**3.** Now estimate the Poisson regression model, using `y` as the response variable and `Bark` as the predictor (and the log link function). Interpret the estimated slope coefficient.


**4.** Let's visualize how well our two models capture the data on its original count scale. Fill in the missing pieces of the predict() functions in the starter code below to generate predictions across a range of bark indices from 0 to 30, then complete the plot.

Compare and contrast the two fitted curves. Explain why the Poisson model predictions are bigger than the Gaussian model predictions. (Hint: the Poisson model gives estimates of the mean on the original scale.)

::: callout-note

Here is some starter code that first makes a `preds` dataset showing the predicted values under the two models for all `Bark` values between 0 and 30. Then, it creates a jittered scatterplot of the original data, and overlays the two fits. 

- For the Log-Normal OLS model (`possum_lognormal`): Remember that this model predicts $\log(y+1)$. To map these predictions back to the original count scale ($y$), you will need to mathematically undo both operations.
- For the Poisson GLM model (`possum_glm`): By default, `predict()` returns values on the log-link scale ($\eta$). Look at the help documentation (`?predict.glm`) to find the specific character string needed for the type argument to return predictions on the original response scale ($\mu$).
- Fill in both blanks in the code with a version of `predict(model, newdata = data.frame(Bark = Bark))`. I've given you most of the `possum_glm` code. 

```{r}
#| eval: false
#| fig-width: 4
#| fig-height: 3

preds <- tibble(
  Bark = 0:30,
  y_ols = _______,
  y_glm = predict(possum_glm, newdata = data.frame(Bark = Bark), type = "____")
)

gf_jitter(y ~ Bark, data = possums, width = .25, height = .25) |>
  gf_line(y_ols ~ Bark, data = preds, col = "darkgreen") |>
  gf_line(y_glm ~ Bark, data = preds, col = "darkblue") 
```
:::

**5.** Calculate a 95% Wald-based confidence interval for the slope coefficient ($\beta_1$) of `Bark` using the Poisson model. Do this (1) by hand using the formula and (2) with R using `confint` or `tidy`.


**6.** Suppose a newly surveyed forest site has a bark quality index of `Bark = 15`. Use your Poisson regression model to calculate the predicted mean number of possum species ($\hat{\mu}$) for this site. Be careful with your scales!

# Problem 2: More `possums`

**1.** As discussed last class, the variable `y` can take the value 0, so we cannot take its logarithm. One possibility for transformation is to add 1 to the value before taking the logarithm. Plot `log(y+1)` vs. each of the predictors and describe what you see. (Starter code below to save you time)

```{r}
#| fig-width: 8
#| fig-height: 6

library(patchwork) 

p1 <- gf_jitter(y ~ Acacia, data = possums, width = .15, height = .15)
p2 <- gf_jitter(y ~ Bark, data = possums, width = .15, height = .15)
p3 <- gf_jitter(y ~ Habitat, data = possums, width = .15, height = .15)
p4 <- gf_jitter(y ~ Shrubs, data = possums, width = .15, height = .15)
p5 <- gf_jitter(y ~ Stags, data = possums, width = .15, height = .15)
p6 <- gf_jitter(y ~ Stumps, data = possums, width = .15, height = .15)

p1 + p2 + p3 + p4 + p5 + p6
```

**2.** Fit the quasipoisson regression model using `y` as the response variable and include all the other predictors as main effects. The estimated dispersion parameter is less than one; argue that using the "regular" Poisson model will give more a conservative (cautious) assessment of uncertainty.

**3.** Refit the model above, now using the standard Poisson model (setting $\phi = 1$). Interpret the estimated effect of the number of hollow trees (`Stags`).

**4.** Give a 95% confidence interval for the effect of the number of hollow trees on the mean number of species.

**5.** Take a look at the fitted values (the estimated means). Should we expect the Pearson residuals to follow a normal distribution in this setting? Can we trust the Deviance Goodness of Fit test?

**6.** Drop `Acacia`, `Shrubs`, and `Stumps` from your model. On the basis of the likelihood ratio test, is there evidence against this reduced model?

**7.** Using the reduced model from part 6, give an estimate of the mean number of species when the bark index is 8, the habitat score is 4, and the number of stumps is 10. Give a 95% confidence interval for this estimate by treating the $\log(\hat{\mu})$ as approximately normally distributed. Why is this justified?

# Problem 3: Reflections

(a) Start by reviewing either HW2 or Case Study 1. (Look at your submission on gradescope) What was something you found difficult initially, but now seems "easy"? 

(b) Write 2-3 sentences describing how you have grown this term. (Some ideas to get you started: overcoming a technical hurdle; managing group work; incorporating R and math; study skills; time management; etc.)

(c) Write 2-3 sentences describing the "main point" of this class. 
