---
title: Evaluating SLR models
subtitle: Day 05
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)
```

## Topics for Exam I


## Last Time

1. Inference for predictions

    - $\widehat{\mu}(Y|X)$
    - $\text{Pred}(Y|X)$
    
### Today: how "good" is our model?

## Two definitions for "good"

<br> 

1. The model is appropriate and our inferences are valid

<br> 

2. The X variable does a good job at explaining the Y variable

## Notation

<br> 

Theoretical model: 

$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br> 

Fitted model: 

$\widehat{y}_i = \widehat\beta_0 + \widehat\beta_1 x_i$

$y_i = \widehat{y}_i + \widehat{\varepsilon}_i$
 

## Bike Data (small sample) 

```{r}
#| fig-width: 4
#| fig-height: 2.5
#| fig-align: 'left'


set.seed(040726)
bike_sample <- bikes |> slice_sample(n = 50)

gf_point(rides ~ temp_actual, data = bike_sample) |>
  gf_lm() + 
  geom_abline(intercept = -2168.77, slope = 89.23, col = "darkorange", linetype = "dashed", linewidth = 1.25)

bike_sm_lm = lm(rides ~ temp_actual, data = bike_sample)
```


```{r}
lm(rides ~ temp_actual, data = bike_sample)
```

## Conditions for inference 

Our model must be valid in order for our p-values or confidence intervals to be valid: 

$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br>

Conditions to check:

- **L**inear relationship is appropriate
- **I**ndependent and identically distributed errors (iid)
- **N**ormally distributed errors
- Errors have **E**qual variance


## Residuals


**Definition**: $e_i =\widehat{\varepsilon}_i = y_i - \widehat{y}_i$

<br>


**Properties:**

-  sum to zero $\Longrightarrow$ mean is 0

-  **uncorrelated** with $x$ and $\widehat{y}$

- normally distributed 
 
- $SD(e_i) = \widehat{\sigma} \sqrt{1 - \dfrac{1}{n} - \dfrac{(x_i - \overline{x})^2}{\sum (x_i - \overline{x})^2}}$


## What is a residual plot?


```{r}
p1 <- ggplot(broom::augment(bike_sm_lm), aes(x = temp_actual, y = rides)) + 
  geom_segment(aes(xend = temp_actual, yend = .fitted), col = "darkorange") + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5)

p2 <- ggplot(broom::augment(bike_sm_lm), aes(x = temp_actual, y = .resid)) + 
  geom_segment(aes(xend = temp_actual, yend = 0), col = "darkorange") + 
  geom_point() + 
  geom_abline(slope = 0, intercept = 0, linewidth = 1.5, col = "blue")

p1 + p2
```

## Put $\widehat{y}$ on the y axis instead of $x$


```{r}
p1 <- ggplot(broom::augment(bike_sm_lm), aes(x = temp_actual, y = .resid)) + 
  geom_segment(aes(xend = temp_actual, yend = 0), col = "darkorange") + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, linewidth = 1.5, col = "blue")

p2 <- ggplot(broom::augment(bike_sm_lm), aes(x = .fitted, y = .resid)) + 
  geom_segment(aes(xend = .fitted, yend = 0), col = "darkorange") + 
  geom_point() + 
  geom_abline(slope = 0, intercept = 0, linewidth = 1.5, col = "blue")

p1 + p2
```


## A "good" residual plot

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| layout-ncol: 2
set.seed(72384)
x <- rnorm(100, mean = 10, sd = 3.5)
y <- 1 + 1*x + rnorm(100, mean = 0, sd = 10)
good_mod <- lm(y ~ x)

gf_point(residuals(good_mod) ~ fitted(good_mod), 
         ylab = "Residual", xlab = "Fitted values") |>
  gf_hline(yintercept = 0, color = "blue", linetype = 2) 

gf_point(residuals(good_mod) ~ x, 
         ylab = "Residual", xlab = "x") |>
  gf_hline(yintercept = 0, color = "blue", linetype = 2) 
```

## Your turn

::: {.task}
- Work in groups

- On the whiteboards, sketch a plot of $y$ vs. $x$ and a corresponding residual plot that would indicate a violation of the

  1. linearity condition

  2. equal variance condition
    
:::

## Assessing Linearity 

## Assessing Equal Variance

## Assessing Normality

- histogram of residuals
- normal Q-Q plot of residuals

**Examples of "good" plots:**

```{r}
#| echo: false
#| layout-ncol: 4
#| fig.height: 3.5
#| fig.width: 5
x2 <- rnorm(30, mean = 10, sd = 3.5)
y2a <- 1 + 1*x2 + rnorm(30, mean = 0, sd = 10)
y2b <- 1 + 1*x2 + rnorm(30, mean = 0, sd = 10)
good_mod2a <- lm(y2a ~ x2)
good_mod2b <- lm(y2b ~ x2)

gf_histogram(~residuals(good_mod2a), color = "white", bins = 12, xlab = "Residual", ylab = "Count") 

gf_histogram(~residuals(good_mod2b), color = "white", bins = 12, xlab = "Residual", ylab = "Count")

gf_qq(~residuals( good_mod2a), ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline() 

gf_qq(~residuals( good_mod2b), ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline()
```

## Harder to gauge at small sample sizes

All of these are random samples from normal distributions

```{r}
#| echo: false
#| layout-ncol: 4
#| fig.height: 3.5
#| fig.width: 5


x1 <- rnorm(10, mean = 10, sd = 3.5)

x2 <- rnorm(50, mean = 10, sd = 3.5)

x3 <- rnorm(100, mean = 10, sd = 3.5)

x4 <- rnorm(500, mean = 10, sd = 3.5)


gf_histogram(~x1, color = "white", bins = 10, xlab = "Residual", ylab = "Count", title = "n = 10") 

gf_histogram(~x2, color = "white", xlab = "Residual", ylab = "Count", title = "n = 50") 

gf_histogram(~x3, color = "white", xlab = "Residual", ylab = "Count", title = "n = 100") 

gf_histogram(~x4, color = "white", xlab = "Residual", ylab = "Count", title = "n = 500") 

gf_qq(~x1, ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline() 

gf_qq(~x2, ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline() 

gf_qq(~x3, ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline() 

gf_qq(~x4, ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline() 
```

## Assessing independence

- plot residuals vs. variable inducing dependence (e.g. time, location, subject ID)

**Examples of "good" plots:**

```{r}
#| echo: false
#| layout-ncol: 2
#| fig.height: 3.5
#| fig.width: 5


gf_point(rstandard(good_mod2a) ~ 1:30, ylab = "Standardized Residual", xlab = "Order") |>
  gf_line() |>
  gf_hline(yintercept = 0)

gf_point(rstandard(good_mod2b) ~ 1:30, ylab = "Standardized Residual", xlab = "Order") |>
  gf_line() |>
  gf_hline(yintercept = 0) 
```

# Your turn

Work through Example 1 on the activity


## What happens the conditions aren't valid? 


-   **Linearity:** if nonlinear, everything breaks!

-   **Independence:** estimates are still unbiased (i.e. we fit the right line) but measures of the accuracy of those estimates (the SEs) are typically too small

-   **Normality:** estimates are still unbiased (i.e. we fit the right line), SEs are correct BUT confidence/prediction intervals are wrong (we can't use t-distribution)

-   **Equal variance:** estimates are still unbiased but standard errors are wrong (and we don't know how wrong)


## What do we do if our assumptions are violated?

1. Change our assumptions (hard, need more stats)

2. Transform $y$, $x$, or both

3. Change the type of inference (remember the bootstrap?)

# Is the X variable a "good" predictor of the Y variable? 

## 

```{r}
#| fig-width: 8
#| fig-height: 4
#| 
p1 <- gf_point(rides ~ temp_actual, data = bike_sample) |>
  gf_lm() + 
  xlim(c(40, 85)) + 
  ylim(c(0, 7000))

p2 <- tibble(
  x = rnorm(50, sd = 6) + 60, 
  y = -2940 + 101*x + rnorm(50, sd = 3000)
) |>
  gf_point(y ~ x) |>
  gf_lm() + 
  xlim(c(40, 85)) + 
  ylim(c(0, 7000))

p3 <- tibble(
  x = rnorm(50, sd = 6) + 60, 
  y = -2940 + 101*x + rnorm(50, mean = 50, sd = 100)
) |>
  gf_point(y ~ x) |>
  gf_lm() + 
  xlim(c(40, 85)) + 
  ylim(c(0, 7000))

p1 + p2 + p3
```

## $R^2$

\begin{align*}
\sum(Y - \bar{Y})^2 &= \sum (\hat{Y} -\bar{Y})^2  \hspace{.9cm} + \sum e_i^2 \\
\text{total variability} &= \text{variability of } \hat{Y}\text{'s} + \text{variability of residuals}\\
\end{align*}

Then: 

$$R^2 = \frac{\sum (\hat{Y} - \bar{Y})^2}{\sum (Y - \bar{Y})} = \frac{\text{variability of } \hat{Y}}{\text{variability of }Y} $$


::: callout-note
## $R^2$

The $R^2$ value gives the proportion of total variability in Y explained by the SLR model
:::

## SSNBATNT

## Finding $R^2$ in R {.smaller}

```{r}
#| echo: true
summary(bike_sm_lm)
```

## Properties of $R^2$

- Between 0 and 1
- Big when $\sigma$ is small relative to the overall variability
    - big proportion of variability explained by the model
- For SLR, $R^2 = r^2$ 
    - ($r$ = correlation coefficient between X and Y)

## $R^2$ dangers


```{r}
#| echo: false

set.seed(040726)

fake_data <- tibble(
  x = c(runif(50, 0, 200), 400),
  y = c(.5*x^2 + c(rnorm(50, 0, 50)))
)

fake_data[51, 2] = 30000

p1 <- ggplot(fake_data, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(
    title = expression(R^2 == .968)
  )

fake_lm <- lm(y ~ x, data = fake_data)

p2 <- ggplot(broom::augment(fake_lm), aes(x = .fitted, y = .resid)) +
  geom_point() 

p1 + p2

# cor(fake_data$x, fake_data$y)
```


