---
title: Inference for Logistic Regression
subtitle: Day 20
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

```

## Running Example

This dataset has information on 150 news articles, some that are real and some that are fake. Our goal is to build a statistical model to explain `type` based on features of the article

- `type`: "real" or "fake" / `type_num`: 0 (real) or 1 (fake)
- `title` and `url` of article
- `title_words`: how many words in the title
- `title_has_excl`: whether title has exclamation
- `title_caps_percent`: percent of title in capital letters
- `negative`: Percent of words that have negative sentiment
- `positive`: Percent of words that have positive sentiment

## 

::: small

```{r}
news <- bayesrules::fake_news |>
  select(title, url, type, title_words, title_has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

write_csv(news, here("data/fake_news_bayesrules.csv"))

knitr::kable(head(news))
```

:::

## Last Time

- Intro to logistic regression
- Moving between log odds &harr; odds &harr; probability scales
- Interpreting coefficients

### Today

- Generalized linear model
- Inference for a single coefficient

## Fitted regression model: 3 scales


:::: {.columns}
::: {.column width="33%" .small}
$\log(\frac{\widehat p}{1-\widehat p}) = \widehat\beta_0 + \widehat\beta_1 x$
:::

::: {.column width="33%" .small}
$\text{odds}(x) = e^{\widehat\beta_0 + \widehat\beta_1 x}$
:::

::: {.column width="33%" .small}
$P(Y = 1 | x) = \frac{e^{\widehat\beta_0 + \widehat\beta_1 x}}{1 + e^{\widehat\beta_0 + \widehat\beta_1 x}}$
:::
::::

```{r}
#| fig-width: 12
#| fig-height: 4
tibble(
  x = seq(0, 60, by = .5),
  log_odds = -0.9318 + 0.09772 * x,
  odds = exp(log_odds),
  prob = exp(log_odds)/(1 + exp(log_odds))
) |>
  pivot_longer(-x) |>
  ggplot(aes(x = x, y = value)) + 
  geom_line(color = "#003300", linewidth = 1.5) + 
  facet_wrap(vars(name), scales = "free_y") + 
  labs(y = "")
```

## Logistic regression: linear predictor with a logit link function

:::: {.columns}
::: {.column width="33%" .small}
$\widehat \eta = \log(\frac{\widehat p}{1-\widehat p}) = \widehat\beta_0 + \widehat\beta_1 x$
:::

::: {.column width="33%" .small}
Link function: logit

$\widehat \eta \to P(Y=1)$
:::

::: {.column width="33%" .small}
$P(Y = 1 | x) = \frac{e^{\widehat\beta_0 + \widehat\beta_1 x}}{1 + e^{\widehat\beta_0 + \widehat\beta_1 x}}$
:::
::::

```{r}
#| fig-width: 12
#| fig-height: 4
tibble(
  x = seq(0, 60, by = .5),
  log_odds = -0.9318 + 0.09772 * x,
  #odds = exp(log_odds),
  prob = exp(log_odds)/(1 + exp(log_odds))
) |>
  pivot_longer(-x) |>
  ggplot(aes(x = x, y = value)) + 
  geom_line(color = "#003300", linewidth = 1.5) + 
  facet_wrap(vars(name), scales = "free_y") + 
  labs(y = "")
```

## Linear regression: linear predictor with an identity link function

:::: {.columns}
::: {.column width="33%" .small}
$\widehat \eta = \widehat y = \widehat\beta_0 + \widehat\beta_1 x$
:::

::: {.column width="33%" .small}
Link function: identity

$f(\eta) = \eta$
:::

::: {.column width="33%" .small}
$\widehat y= \widehat \beta_0 + \widehat \beta_1 x$
:::
::::

```{r}
#| fig-width: 12
#| fig-height: 4
tibble(
  x = seq(0, 60, by = .5),
  eta = -0.9318 + 0.09772 * x,
  y_hat = eta,
  #odds = exp(log_odds),
  #prob = exp(log_odds)/(1 + exp(log_odds))
) |>
  pivot_longer(-x) |>
  ggplot(aes(x = x, y = value)) + 
  geom_line(color = "#003300", linewidth = 1.5) + 
  facet_wrap(vars(name), scales = "free_y") + 
  labs(y = "")

```


## Generalized linear model

<br>

**Linear Predictor**

$\eta = \beta_0 + \beta_1 x_1 + ... + \beta_p x_p$

<br>

**Link Function**

$\theta = g(\eta)$

<br>

**Response distribution**

$Y$ is a random variable with mean $\theta$

## Binary logistic regression

If $Y$ follows a Bernoulli distribution

$${\rm E}(Y | X) = \pi(X)$$

We link this mean function to the explanatory variables using the logit link

\begin{align*}
\eta &= {\rm logit}(\pi(X))\\ 
&= \log \left( \dfrac{\pi(X)}{1 - \pi(X)} \right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p
\end{align*}

## Fake news model (from activity)


```{r}
news_mod <- glm(type_num ~ title_words + title_has_excl + title_caps_percent, 
                data = news, family = "binomial")
```

::: small

```{r}
equatiomatic::extract_eq(news_mod, intercept = "beta", terms_per_line = 2,
                         swap_var_names = c("type_num" = "type", "title_words" = "words",
                                            "title_has_excl" = "excl", "title_caps_percent" = "pct_caps"))
```

:::

```{r}
tidy(news_mod)
```

# Wald-based inference 

## Maximum likelihood (ML) estimation {.smaller}

The coefficients in logistic regression are estimated by finding the $\widehat{\beta}_0, \ldots, \widehat{\beta}_p$ that maximize the probability of the observed outcomes



$$L(\beta) = P(Y_1 = y_1, \ldots Y_n = y_n | \beta, X) = \prod_{i=1}^{n} \pi(X)^{y_i} \Big[ 1 - \pi(X) \Big]^{1- y_i}$$


where

$$\pi(X) = \dfrac{e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}}{1 + e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}}$$

::: footer
For more, consider taking Stat250! 
:::



## Properties of ML estimators 

If the sample is large and the model is correct, the ML estimator $\hat\beta_i$ is: 

1. Essentially unbiased

1. SEs can be computed and are about as small as any other unbiased estimator

1. The sampling distributions for the estimators are approximately **normal**

::: footer
For more, consider taking Stat250! 
:::

## Wald test for a coefficient 

:::: {.columns}
::: {.column width="25%"}
**Hypotheses:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=1}
$H_0: \beta_i = 0 \quad$ vs. $\quad H_a: \beta_i \ne 0$
:::
:::

::::

:::: {.columns}
::: {.column width="25%"}
**Test statistic:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=2}
$z = \dfrac{\widehat{\beta}_i}{SE(\widehat{\beta}_i)}$
:::
:::

::::

<br>


:::: {.columns}
::: {.column width="25%"}
**Reference distribution:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=3}
$N(0, 1)$
:::
:::

::::




## CI for a coefficient 

A normal-based confidence interval for ${\beta}_i$

$$\widehat{\beta}_i \pm z^*_{1-\alpha/2} SE(\widehat{\beta}_i)$$

. . .

CI for the multiplicative effect on odds of success for a 1- unit change in $x$ (odds ratio of Success for $x+1$ vs. $x$):

$$e^{\widehat{\beta}_i - z^*_{1-\alpha/2} SE(\widehat{\beta}_i)} \quad \text{to} \quad e^{\widehat{\beta}_i + z^*_{1-\alpha/2} SE(\widehat{\beta}_i)}$$

. . .

... for a $C$-unit change in $x$

$$e^{C \cdot\widehat{\beta}_i - z^*_{1-\alpha/2} C \cdot SE(\widehat{\beta}_i)} \quad \text{to} \quad e^{C \cdot\widehat{\beta}_i + z^*_{1-\alpha/2} C \cdot SE(\widehat{\beta}_i)}$$


## fake news example

What impact does the *number of words in the title* have on the odds of being fake news?

```{r echo=FALSE}
tidy(news_mod, conf.int = TRUE) |> select(-statistic, -p.value)
```

::: {.fragment fragment-index=1}
Let's construct a 95% confidence interval for $\beta_1$: $\quad \widehat{\beta}_1 \pm z^*_{1-\alpha/2} SE(\widehat{\beta}_1)$
:::

::: {.fragment fragment-index=2}
$z^*_{1-0.05/2} = z^*_{0.975} =$ 0.975 quantile from $N(0, 1)$ 
:::

::::: {.fragment fragment-index=3}
:::: {.columns}
::: {.column width="35%"}
```{r}
qnorm(0.975)
```
:::
::::
:::::

::: {.fragment fragment-index=4}
$0.120 \pm 1.96 \cdot (0.0581) = (0.008,\ 0.237)$
:::


## fake news example

$0.120 \pm 1.96 \cdot (0.058) = (0.008,\ 0.237)$

<br>

::: {.fragment fragment-index=1}
Exponentiating the endpoints to get the CI for the odds ratio for a one-word increase in title length:

$$e^{0.008} = 1.008 \quad \text{to} \quad e^{0.0724} = 1.267$$
:::
<br>

::: {.fragment fragment-index=2}
We are 95% confident that a one-word increase in title length is associated with an increase in the odds of being fake news by between 0.8% (a 1.008 factor increase) and  26.7% (a 1.267 factor increase), holding all other variables constant. 
:::


## fake news example

::: small
How do the odds of being fake news change for an increase of 10 words in the title length?

```{r echo=FALSE}
tidy(news_mod, conf.int = TRUE) |> select(-statistic, -p.value)
```
:::

::: {.fragment fragment-index=1}
:::{style="font-size:0.9em"}
95% CI: $\quad C \cdot \left[\widehat{\beta}_1 \pm z^*_{1-\alpha/2} SE(\widehat{\beta}_1)\right]$
:::
:::


::: {.fragment fragment-index=2}
:::{style="font-size:0.9em"}
$\begin{split}
10 \left[0.120 \pm 1.96 \cdot (0.058) \right] &= (10\cdot(0.008),10\cdot(0.237))\\ &= (0.08, \ 2.37)\end{split}$
:::
:::

::: {.fragment fragment-index=3}
:::{style="font-size:0.9em"}
We are 95% confident that a 10-word increase in title length is associated with an increase in the odds of being fake news by between $e^{.08} = 1.08$ (an 8% increase) and a factor of $e^{2.37} = 10.69$ (a 1000% increase!), holding all other variables constant. 
:::
:::

# Pause for the most dramatic logistic regression story I've ever heard

# How can we tell if inference is correct?

## Recall

LINE conditions must be met for inference to be valid

- If Linearity fails, the line doesn't actually describe the relationship and everything breaks

- If INE fails, we can still interpret our coefficients (the geometry is still correct), but cannot trust inference
    - Confidence intervals too narrow
    - p-values too small

## What are the conditions for logistic regression model? 

1. **L**inear relationship between the log odds (logit) and the predictors

2. **I**ndependence: no pairs, clusters, etc.

3. **R**andomness: we have a random sample from the population or the data were collected from a randomized experiment

## "Residuals"

In logistic regression, we'll work with residuals on the response scale: 

$$e_i = y_i - \widehat \pi_i = \begin{cases} 1 - \widehat \pi_i & \text{if } y_i = 1 \\ - \widehat \pi_i & \text{if } y_i = 0 \end{cases}$$

What will residual plot look like? 

```{r}
#| fig-width: 12
#| fig-height: 4
p1 <- tibble(
  x = seq(0, 50, by = .5),
  log_odds = -0.9318 + 0.09772 * x,
  odds = exp(log_odds),
  prob = exp(log_odds)/(1 + exp(log_odds))
) |>
  ggplot(aes(x = x, y = prob)) + 
  geom_line(color = "#003300", linewidth = 1.5) +
  geom_jitter(data = news, aes(x = title_caps_percent, y = type_num), height = .05, width = .1) +
  labs(y = "Probability")

p2 <- tibble(
  x = seq(0, 50, by = .5),
  log_odds = -0.9318 + 0.09772 * x,
  odds = exp(log_odds),
  prob = exp(log_odds)/(1 + exp(log_odds))
) |>
  ggplot(aes(x = x, y = prob)) + 
  labs(y = "residual") +
  ylim(c(-1,1))

p1 + p2
```

## Residual plot

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 4
resid_panel(news_mod, plot = "resid", type = "response")
```

## Binned residuals

Idea: If we group data points based on predictor, we can compute the mean response residual within each group and end up with something that looks more like an OLS residual plot

::: incremental
1. Calculate response residual
2. Order observations either by the values of the predicted log odds/probabilities, or by quantitative predictor variable
3. Use the ordered data to create g bins of approximately equal size. (Default: $g = \sqrt{n}$)
4. Calculate average residual value in each bin
5. Plot average residuals vs. average predicted log odds, probability, or predictor value
:::

## Binned residual plot 

- Using 13 bins
- You will recreate this during the R activity today

```{r}
#| fig-width: 5
#| fig-height: 3
news_aug <- augment(news_mod, type.predict= "link", type.residuals = "pearson") |>
  mutate(.resp.resid = resid(news_mod, type= "response"))

binned_resids <- news_aug |> 
  mutate(bin = ntile(news_mod$linear.predictors, 13)) |>
  group_by(bin) |>
  summarize(mean_linear_pred = mean(.fitted),
            avg_resid = mean(.resp.resid),
            .groups = "drop")

gf_point(avg_resid ~ mean_linear_pred, data = binned_resids,
         xlab = "Predicted log odds", ylab = "Average response residual") |>
  gf_hline(yintercept = 0, linetype = 2) 
```

## Binned residual 

- Nonlinear trend may be indication that squared term or log transformation of predictor variable required
- If bins have large average residuals (in magnitude)
  - Look at averages of other predictor variables across bins
  - Interaction may be required if large magnitude residuals correspond to certain combinations of predictor variables

## Check binned residuals against X variables

```{r}
#| fig-width: 5
#| fig-height: 3
news_aug <- augment(news_mod, type.predict= "link", type.residuals = "pearson") |>
  mutate(.resp.resid = resid(news_mod, type= "response"))

p1 <- news_aug |> 
  mutate(bin = ntile(title_words, 13)) |>
  group_by(bin) |>
  summarize(mean_x = mean(title_words),
            avg_resid = mean(.resp.resid),
            .groups = "drop") |>
  gf_point(avg_resid ~ mean_x,
         xlab = "title words", ylab = "Average response residual") |>
  gf_hline(yintercept = 0, linetype = 2) 


p2 <- news_aug |> 
  mutate(bin = ntile(title_caps_percent, 13)) |>
  group_by(bin) |>
  summarize(mean_x = mean(title_caps_percent),
            avg_resid = mean(.resp.resid),
            .groups = "drop") |>
  gf_point(avg_resid ~ mean_x,
         xlab = "title words", ylab = "Average response residual") |>
  gf_hline(yintercept = 0, linetype = 2) 


p1 + p2
```

## Residuals vs categorical predictor

- Calculate average residual for each level of the predictor
- All of the means should be close to 0

```{r}
news_aug |>
  group_by(title_has_excl) |>
  summarize(`avg residual` = mean(.resp.resid))
```

## Pearson residual

Standardize the distance between the observation (0 or 1) and the expected value (probability)

$$r_{p_i} = \frac{y_i - \widehat \pi_i}{\sqrt{\widehat{\pi_i}(1-\widehat \pi_i)}}$$

## Pearson residuals in R 

- If the model is correct, then the expected (average) Pearson residual within each bin should be close to 0
- A smoother should be approximately horizontal line at 0

```{r}
#| echo: true
#| layout-ncol: 2
#| fig-width: 4
#| fig-height: 2.5
resid_panel(news_mod, plot = "resid", type = "pearson")
resid_panel(news_mod, plot = "resid", type = "pearson", smoother = TRUE)
```

## 

Smoothers can help draw attention to major flaws, but there will always be variation from the horizontal!

```{r}
#| cache: true
#| echo: false
#| fig-height: 6
#| fig-width: 8
#| fig-align: 'center'
#| message: false
library(nullabor) # lineup tools
library(tidyverse)

set.seed(051326)

N <- 1000
df <- tibble(
  x1 = rnorm(N),
  x2 = rnorm(N),
  prob_y = exp(-1 + x1 + x2 + 0.7 * x1^2) / 
    (1 + exp(-1 + x1 + x2 + 0.7 * x1^2)),
  y = rbinom(N, size = 1, prob = prob_y)
)

no_quad <- glm(y ~ x1 + x2, data = df, family = "binomial")
no_quad_aug <- augment(no_quad, type.residuals = "pearson")

sim_ys <- simulate(no_quad, nsim = 19)
sim_aug <- map_dfr(
  sim_ys, 
  ~augment(glm(.x ~ x1 + x2, data = df, family = binomial))
)
sim_aug$.n <- rep(1:19, each = nrow(df))

no_quad_raw_lineup <- lineup(true = no_quad_aug, n = 19, samples = sim_aug) %>%
    ggplot(aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange") +
  facet_wrap(~.sample, ncol = 5) +
  labs(x = "Linear predictor", y = "Pearson residuals") +
  theme_bw()

no_quad_raw_lineup
```

::: footer
Only one of these datasets was simulated from the "wrong" distribution
:::

# Checking linearity is hard! 

## Independence

- We conduct inference assuming we have $n$ pieces of independent information from the $n$ data points 
- If the data points are dependent, we have *less* pieces of information than $n$ and our inference is overoptimistic
- Best to assess based on description of the data: 
    - Are they grouped in some way? 
    - Measurements over time?
    - Spatial component?
- The reason we make an index plot is to see if our data is ordered in some way

## Independence: index plot

```{r}
#| echo: true
#| fig-width: 4
#| fig-height: 3
resid_panel(news_mod, plot = "index", type = "pearson")
```

## Randomness

> A dataset containing data behind the study "FakeNewsNet: A Data Repository with News Content, Social Context and Spatialtemporal Information for Studying Fake News on Social Media" https://arxiv.org/abs/1809.01286. The news articles in this dataset were posted to Facebook in September 2016, in the run-up to the U.S. presidential election.


