---
title: Inference for MLR
subtitle: Day 11
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

bikes <- bikes |> select(date, season, year, month, weekend, holiday, temp_actual, temp_feel, humidity, windspeed, rides)

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt", sep = "\t", header = TRUE)

bikes_slr <- lm(rides ~ temp_actual, data = bikes)


```

## Warm up: new groups!

::: task

- Introduce yourselves
- Group expectations worksheet

:::

## MLR model for `bikes` data {.smaller}

```{r}
bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)
bikes_mlr
```

::: task
Write out the theoretical and fitted regression model
:::

## Do we _need_ a different slopes model? 

```{r}
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes) |>
  gf_lm()

tidy(bikes_mlr) 
```



## Testing a single coefficient

::: {style="font-size: 0.9em;"}
**Hypotheses:** $H_0: \ \beta_j = \#$ vs. $H_a: \ \beta_j \underset{>}{\overset{<}{\ne}} \#$
:::

. . .

::: {style="font-size: 0.9em;"}
**Test statistic:** $t = \dfrac{\widehat{\beta}_j - \#}{SE(\beta_j)}$
:::
. . .

::: {style="font-size: 0.9em;"}
**Reference distribution:** $t$ distribution with d.f. = $n-(k+1)$


**p-value:** Area in the tail(s) specified by $H_a$
:::


```{r echo=FALSE, fig.height = 2, fig.width = 7.5, out.width=850}
p1 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] < 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p2 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] != 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p3 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] > 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

(p1 | p2 | p3)
```

## CIs for a single coefficient


$\widehat{\beta}_j \pm t^*_{n-(k+1)} \cdot SE(\widehat{\beta}_j )$


## Answer: Do we need a different slopes model? 


## Do we need `weekend` at all?


```{r}
tidy(bikes_mlr) 
```

::: callout-caution
## Caution!

<br>

<br>

<br>

<br>

:::

## Do we need `weekend` at all?

```{r}
bikes_mlr2 <- lm(rides ~ temp_actual + weekend, data = bikes)
tidy(bikes_mlr2)
```

## Explained Variability

```{r echo=FALSE, fig.height = 3.25, fig.width = 6, fig.align='center'}
slr_aug <- augment(bikes_slr)
slr_aug <- mutate(slr_aug, mean.y = mean(rides))

p1 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Total") %>%
  gf_segment(mean.y + rides ~ temp_actual + temp_actual, color = "gray70", alpha = .4) %>%
  gf_hline(yintercept = mean(slr_aug$rides), color = "#f15a31") 

p2 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Error") %>%
  gf_segment(.fitted + rides ~ temp_actual + temp_actual, color = "#003069", alpha = .4) %>%
  gf_lm(color = "#f15a31") 

p1 + p2
```

## Explained Variability


```{r}
ssdf <- tribble(
  ~Source, ~SS,
  "Error",      6,
  "Total",       10,
  "Regression", 4
)

sst_plot <- ggplot(data = ssdf |> filter(Source == "Total"), aes(x = "", y = SS)) +
  geom_bar(stat="identity", width=1) +
  theme_void() +
  labs(title = "SSTotal")


decomp_plot <- ggplot(data = ssdf |> filter(Source != "Total"), aes(x = "", y = SS, fill = Source)) +
  geom_bar(stat="identity", width=1) +
  geom_text(aes(label = paste0(Source)), position = position_stack(vjust = 0.5), color = "white") +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#f15a31", "#003069"))

ssdf2 <- tribble(
  ~Source, ~SS,
  "Error",      5,
  "Total",       10,
  "Regression", 5
)

sst_plot2 <- ggplot(data = ssdf2 |> filter(Source == "Total"), aes(x = "", y = SS)) +
  geom_bar(stat="identity", width=1) +
  theme_void() +
  labs(title = "SSTotal")


decomp_plot2 <- ggplot(data = ssdf2 |> filter(Source != "Total"), aes(x = "", y = SS, fill = Source)) +
  geom_bar(stat="identity", width=1) +
  geom_text(aes(label = paste0(Source)), position = position_stack(vjust = 0.5), color = "white") +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#f15a31", "#003069"))
```

::: columns
:::{.column width="30%"}
"Smaller" Model


```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
sst_plot + decomp_plot
```
:::
:::{.column width="30%"}
"Larger" Model


```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
sst_plot2 + decomp_plot2
```
:::

:::{.column width="30%" .fragment .center}
**Idea:** Is the proportion of variability explained by Model 2 significantly better than Model 1?

:::
:::

## Can we just use $R^2$? 

- About `r round(summary(bikes_slr)$r.squared, 4)*100`% of the variability in rides is explained by the regression model including temperature
- About `r round(summary(bikes_mlr2)$r.squared, 4)*100`% of the variability in rides is explained by the regression model including temperature and an indicator for weekend
- About `r round(summary(bikes_mlr)$r.squared, 4)*100`% of the variability in rides is explained by the regression model including temperature and an indicator for weekend

What trend do you notice? 

## $R^2$ vs adjusted $R^2$

::: columns
:::{.column width="50%"}

$$R^2 = 1 - \frac{SSE}{SST}$$

:::
:::{.column width="50%"}
$$R^2_{adj} = 1 - \frac{SSE/(n-k)}{SST/(n-1)}$$
$$R^2_{adj} = 1 - \frac{MSE}{MST}$$
:::

:::

- $R^2$ **always** increases when predictors are added to the model
- $R^2_{adj}$ **could** decrease if unnecessary predictors are added to the model
- Better to use $R^2_{adj}$ for multiple regression

## F Test for comparing models

:::: {.columns align=center}
::: {.column width="25%"}
Full model
:::

::: {.column width="70%"}
$\mu \lbrace Y | X \rbrace = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2$
:::

::: {.column width="25%"}
Reduced model
:::

::: {.column width="70%"}
$\mu \lbrace Y | X \rbrace = \beta_0 + \beta_1 x_1$
:::

::::

. . .

<br>


:::: {.columns align=center}
::: {.column width="25%"}
Hypotheses
:::

::: {.column width="70%"}
$H_0:\ \beta_{2} = \beta_3 = 0$

$H_a:$ at least one $\beta_j \ne 0$, $j=2,3$
:::

::::

## Comparing models with an F-test



Test statistic


$$
\begin{split}
F &= \frac{(R^2_{\text{full}} - R^2_{\text{reduced}}) / d}{(1 - R^2_{\text{full}}) / df_{\text{full}}}\\
&= \frac{(\text{SSR}_{\text{full}} - \text{SSR}_{\text{reduced}}) / d}{\text{MSE}_{\text{full}}}\\
&= \frac{(\text{SSE}_{\text{reduced}} - \text{SSE}_{\text{full}}) / d}{\text{MSE}_{\text{full}}}
\end{split}
$$

::: xsmall

where 

- $d = \text{df}_{full} - \text{df}_{reduced}$ = \# betas being tested
- $\text{df}_{i} = n - (p + 1)=$ error d.f. for model $i$
- $\text{MSE}_{\text{full}} = \dfrac{\text{SSE}_{\text{full}}}{\text{df}_{\text{full}}}$

:::


## F distribution


The F-statistics follows an $F$ distribution with $d$ and $n-p-1$ d.f.

```{r fig.height=3, fig.width=5, echo=FALSE, message=FALSE, fig.align='center'}
par(mar = c(2, .1, .1, .1))
plot(c(0, 5), c(0, df(.01, 1, 1)), type = "n", axes = FALSE)
at <- seq(0, 5, 1)
axis(1, at)
# axis(1, at, rep("", length(at)), tcl = -0.1)
X <- seq(0, 5, 0.01)
DF1 <- c(1, 2, 5, 10, 100)
DF2 <- c(1, 1, 2, 1, 100)
COL. <- RColorBrewer::brewer.pal(n = 7, name = "Blues")[-1]
for (i in 1:5) {
  Y <- df(X, as.numeric(DF1[i]), as.numeric(DF2[i]))
  lines(X, Y, col = COL.[i])
}
legend(2.5, 3.25,
       legend = c(paste('df1 = ', DF1[1:5], ', df2 = ', DF2[1:5], sep = '')),
       col = COL.,
       text.col = COL.,
       lty = rep(1, 5))

```

```{r include=FALSE}
dev.off
```

## Upper-tail p-values

```{r fig.height=3, fig.width=5, echo=FALSE, message=FALSE, fig.align='center'}
 ggplot(data = data.frame(x = c(.00001, 5.5)), aes(x)) +
  stat_function(fun = df, args = list(df1=4, df2=23), color = "navy", size = 1) +
  stat_function(fun = df, args = list(df1=4, df2=23), xlim = c(3, 5.5),
                  geom = "area", fill = "navy", alpha = .4) +
  labs(x = "F", y = "Density") +
  scale_y_continuous(limits = c(0, 0.8), expand = c(0,0)) +
   geom_curve(aes(x = 5, y = .22, xend = 4.25, yend = .03), arrow = arrow(length = unit(0.03, "npc")), curvature = .2, color = "navy") +
  annotate("text", x = 5.26, y = .25, label = "p-value", color = "navy", size = 6) 
```


To obtain upper-tail areas: `1 - pf(stat, df1, df2)`


## Putting it all together



```{r}
full_table <- glance(bikes_mlr) |> 
  select(r.squared, df, df.residual, nobs) |>
  mutate(model = "Full")

reduced_table <- glance(bikes_slr) |> 
  select(r.squared, df, df.residual, nobs) |>
  mutate(model = "Reduced")

bind_rows(full_table, reduced_table) |>
  select(model, everything()) |>
  knitr::kable(digits = 3)
```

. . .

$$
F = \frac{(R^2_{\text{full}} - R^2_{\text{reduced}}) / d}{(1 - R^2_{\text{full}}) / df_{\text{full}}} = \frac{(0.371 - 0.332) / 2}{(1 - 0.371) / 496} \approx 15.4
$$

. . .

```{r}
#| echo: true
1 - pf(15.4, 2, 496)
```




## Reading R output

```
lm(formula = rides ~ temp_actual * weekend, data = bikes)

Coefficients:
                         Estimate Std. Error t value Pr(>|t|)    
(Intercept)             -1843.867    415.700  -4.436 1.13e-05 ***
temp_actual                87.173      6.439  13.538  < 2e-16 ***
weekendTRUE              -714.938    795.698  -0.899    0.369    
temp_actual:weekendTRUE     0.411     12.539   0.033    0.974    

Residual standard error: 1253 on 496 degrees of freedom
Multiple R-squared:  0.3709,	Adjusted R-squared:  0.3671 
F-statistic: 97.47 on 3 and 496 DF,  p-value: < 2.2e-16
```


::: footer
Output was edited to fit on one slide
:::


## Extra sums of squares F-test in R

```{r}
#| echo: true
full <- lm(formula = rides ~ temp_actual * weekend, data = bikes)    #<1>
reduced <- lm(formula = rides ~ temp_actual, data = bikes) #<2>
anova(reduced, full) #<3>
```

1. Fit the full model
2. Fit the reduced model
3. Use `anova()` to compare the two models. The first argument should be the reduced model.







