---
title: Adding Categorical Predictors
subtitle: Day 10
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

bikes <- bikes |> select(date, season, year, month, weekend, holiday, temp_actual, temp_feel, humidity, windspeed, rides)

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt", sep = "\t", header = TRUE) |>
  mutate(Smoke = factor(Smoke, levels = c(0, 1), labels = c("Nonsmoker", "Smoker")))

mod_parallel <- lm(FEV ~ Age + Smoke, data = fev)
mod_diff_slope <- lm(FEV ~ Age + Age:Smoke, data = fev)
mod_separate <- lm(FEV ~ Age * Smoke, data = fev)

fev_fit_plot <- function(mod) {
  ggplot(augment(mod, data = fev), aes(x = Age, y = FEV, color = Smoke, shape = Smoke)) +
    geom_point(alpha = 0.5) +
    geom_line(aes(y = .fitted), linewidth = 1) +
    scale_color_manual(values = c("#f15a31", "#000036")) +
    labs(y = "FEV (in L/s)")
}

```


## Last time: `bikes` MLR

```{r}
bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)

equatiomatic::extract_eq(bikes_mlr, use_coefs = TRUE)
```

<br> $R^2 =$ `r round(summary(bikes_mlr)$r.squared, 3)`

<br>

. . . 

Both predictors so far have been *quantitative*. What if a predictor is *categorical*?

## New data example {.smaller}

:::: {.columns align=center}
::: {.column width="20%"}
**Goal:**

:::

::: {.column width="80%" .fragment}
investigate the association between smoking and lung capacity using data from 345 adolescents between the ages of 10 and 19
:::

::::


:::: {.columns align=center}
::: {.column width="20%"}
**Wrinkle:**
:::

::: {.column width="80%" .fragment}
Lung function is expected to increase during adolescence, but smoking may slow its progression

:::

::::

<br>

:::: {.columns align=center}
::: {.column width="20%"}
**Data:**
:::

::: {.column width="80%" .fragment}
Variable | Description
:---------|:---------------
`FEV`    | forced expiratory volume (in liters per second)
`Age`    | age in years
`Smoke`  | `Smoker` or `Nonsmoker`

:::

::::

:::{.footer}
Source: Rosner (2006) 
:::

## EDA

```{r echo=FALSE, fig.width = 5, fig.height = 3, fig.align='center', out.width = 800}
ggplot(data = fev) +
  geom_boxplot(aes(x = Smoke, y = FEV, fill = Smoke), alpha = 0.6) +
  scale_fill_manual(values = c("#f15a31", "#000036")) +
  theme(legend.position = "none") +
  coord_flip() +
  labs(x = "", y = "FEV (in L/s)")
```

## EDA

```{r echo=FALSE, fig.width = 5, fig.height = 3.5, fig.align='center', out.width = 800}
ggplot(data = fev) +
  geom_point(aes(x = Age, y = FEV, color = Smoke, shape = Smoke), alpha = 0.5) +
  scale_color_manual(values = c("#f15a31", "#000036")) +
  labs(y = "FEV (in L/s)")
```

. . . 

Smokers skew *older*, so any raw FEV difference is tangled up with age

## Indicator variables

Regression requires a numeric representation of all variables, so we create an **indicator** (0/1) variable

<br>

Create a `Smoker` indicator variable:

- `Smoker` = 1

- `Nonsmoker` = 0

::: task
## Example 1

For each regression model on the handout, sketch the fitted model on the whiteboard

- Each fitted model will have two lines: one for smokers and one for nonsmokers
- Work with your groups (last day together!) 
:::

## Parallel lines model

FEV increases at the same rate with age, but is shifted up or down for smokers

```{r echo=FALSE, fig.width = 6, fig.height = 3.5, fig.align='center', out.width = 800}
fev_fit_plot(mod_parallel)
```

## Different slopes model

FEV starts at the same baseline (age 0), but the relationship between age and FEV is different for smokers vs nonsmokers

```{r echo=FALSE, fig.width = 6, fig.height = 3.5, fig.align='center', out.width = 800}
fev_fit_plot(mod_diff_slope)
```

## Separate lines model

Smokers/nonsmokers have different baselines *and* different relationships between age and FEV

```{r echo=FALSE, fig.width = 6, fig.height = 3.5, fig.align='center', out.width = 800}
fev_fit_plot(mod_separate)
```

## Parallel lines model

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{age} + \beta_2 \texttt{smoker}$$

<br> 

- $\beta_0$: intercept for nonsmokers
- $\beta_0 + \beta_2$: intercept for smokers
- $\beta_1$: slope for both groups

## Different slopes model

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{age} + \beta_2 \texttt{age} \times \texttt{smoker}$$

<br> 

- $\beta_0$: intercept for both groups
- $\beta_1$: slope for nonsmokers
- $\beta_1 + \beta_2$: slope for smokers

## Separate lines model

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{age} + \beta_2 \texttt{smoker} + \beta_3 \texttt{age} \times \texttt{smoker}$$

<br> 

- $\beta_0$: intercept for nonsmokers
- $\beta_0 + \beta_2$: intercept for smokers
- $\beta_1$: slope for nonsmokers
- $\beta_1 + \beta_3$: slope for smokers

## Fitting the separate lines model in R

```{r}
#| echo: true
fev_sep <- lm(FEV ~ Age * Smoke, data = fev)
coef(fev_sep)
```

<br>

R names an indicator after the level that is coded **1**. The level that never appears in a coefficient name (here, `Nonsmoker`) is the **reference level**.

## Reading the coefficient names

| R output | Model term | Meaning |
|:--------------|:------|:-------------------------------------|
| `(Intercept)` | $\beta_0$ | intercept for nonsmokers (reference level) |
| `Age` | $\beta_1$ | slope for nonsmokers |
| `SmokeSmoker` | $\beta_2$ | **difference** in intercepts (smokers $-$ nonsmokers) |
| `Age:SmokeSmoker` | $\beta_3$ | **difference** in slopes (smokers $-$ nonsmokers) |

## Fitted lines:

Nonsmokers: $\widehat{\text{FEV}} = `r round(coef(fev_sep)[1], 2)` + `r round(coef(fev_sep)[2], 2)` \, \texttt{age}$

<br> 

Smokers: $\widehat{\text{FEV}} = (`r round(coef(fev_sep)[1], 2)` + `r round(coef(fev_sep)[3], 2)`) + (`r round(coef(fev_sep)[2], 2)` `r round(coef(fev_sep)[4], 2)`) \, \texttt{age} = `r round(coef(fev_sep)[1] + coef(fev_sep)[3], 2)` + `r round(coef(fev_sep)[2] + coef(fev_sep)[4], 2)` \, \texttt{age}$

## Same idea with a `TRUE`/`FALSE` predictor

`weekend` in the `bikes` data is `TRUE`/`FALSE`, so R treats `TRUE` as 1 and names the coefficients `weekendTRUE`

```{r}
#| echo: true
tidy(lm(rides ~ temp_actual * weekend, data = bikes))
```

We'll pick this model back up on Wednesday

# More than 2 categories

What if we want to incorporate `season`? (fall, spring, summer, winter)

## A possibly bad idea

Convert the column to numeric: 

- Fall $\to$ 1
- Winter $\to$ 2
- Spring $\to$ 3
- Summer $\to$ 4

$$\mu(\text{rides} | \text{season}) = \beta_0 + \beta_1 \text{season}$$

## A good idea

Create a set of indicator (dummy) variables to represent the four categories

:::: {.columns}

::: {.column width="25%"}
|Original|
|--------|
| winter |
| fall |
| spring  |
| spring |
| fall |
| summer |
:::

::: {.column width="20%" .fragment}
| winter |
|--------|
| 1|
| 0|
|0 |
| 0|
| 0|
| 0|
:::

::: {.column width="20%" .fragment}
| spring |
|--------|
| 0|
| 0|
| 1|
| 1|
| 0|
| 0|
:::

::: {.column width="20%" .fragment}
|summer|
|--------|
| 0|
| 0|
| 0|
| 0|
| 0|
| 1|
:::

::::

# Indicator variables

For a categorical variable with $k$ levels, create $k-1$ indicator variables to fully represent the original variable

## Example 2

::: task
Depression score vs. age, with education (college / some college / HS only) as a 3-level categorical predictor

- Define the indicator variables
- For each of the two sketched models on the handout, write the matching mean function (line type = education level)
- Work with your groups

:::
