---
title: Overdispersion and Inference for Poisson Regression
subtitle: Day 27
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```

## Motivating data: number of anti-discrimination laws

The `equality_index` data includes the observed number of anti-discrimination laws in each state,  $Y$ . We are interested in modeling  $Y$ by a state’s `percent_urban` and its `historical` presidential voting patterns. The `historical` variable has 3 levels:

- `dem` = the state typically votes for the Democratic candidate
- `gop` = the state typically votes for the Republican (GOP) candidate
- `swing` = the state often swings back and forth between Democratic and Republican candidates

::: footer
Source: the {bayesrules} R package
:::

## EDA


```{r}
#| echo: false 
#| fig-width: 8
#| fig-height: 3


data("equality_index")

p1 <- ggplot(equality_index, aes(x = laws)) + 
  geom_histogram(col = "white")

p2 <- ggplot(equality_index, aes(y = laws, x = percent_urban)) + 
  geom_point(size = .5)

p3 <- ggplot(equality_index, aes(y = laws, x = historical)) + 
  geom_boxplot(aes(fill = historical), alpha = .6) +
  geom_point(size = .5) +
  theme(legend.position = "none")

p2 + p1 + p3
```
- `laws` is skewed
- as percent urban increases, laws also increases
- `dem` and `swing` generally have more laws than `gop`

## Review: Poisson regression model

- If $Y_i =$ # of events in a fixed interval of time/space, then $Y_i \sim \text{Poisson}(\lambda)$ 


- This leads to the Poisson regression model for counts:

    $$\log(\lambda) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$
    

- or equivalently

$$\lambda = e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}$$


## Assumptions

1. **Linearity**: the log of the mean of Y, $\log(\lambda)$, must be a linear function of the predictors

2. **Mean = Variance**: the mean of a Poisson random variable must be equal to its variance

3. **Independence**: the observations must be independent

## Model fitting {.smaller}

```{r echo=TRUE}
full_glm <- glm(laws ~ percent_urban + historical,
            data = equality_index,
            family = poisson)
summary(full_glm)
```

## Pearson residuals

$$Pres_i = \frac{Y_i - \widehat{\text{mean}(Y_i)}}{\widehat{SD(Y_i)}}$$



```{r}
#| fig-width: 8
#| fig-height: 3
ggResidpanel::resid_xpanel(full_glm, type = "pearson")
```


## For Poisson model

$$Pres_i = \frac{Y_i - \widehat \mu_i}{\widehat{SD(Y_i)}}$$


- Mean = $\widehat \lambda$, SD = $\sqrt{\widehat \lambda}$
- Pearson residual: $\frac{Y_i - \widehat \lambda_i}{\sqrt{\widehat \lambda_i}}$
- If $\widehat\lambda$'s are "big" ($\ge 5$), Pearson residuals are $\approx N(0,1)$ if Poisson model is correct. Can use to diagnose problems with linearity
- If overdispersed, then SD of Pearson residuals is $>1$

## Goodness of Fit test

<br> 

**$H_0$:** The Poisson model fits the data adequately.

**$H_A$:** The Poisson model is deficient (a more complex, saturated model is required).

<br>

The **Residual Deviance** ($D$) serves as our test statistic:
$$D = 2 \sum_{i=1}^n \left[ y_i \log\left(\frac{y_i}{\widehat{\lambda}_i}\right) - (y_i - \widehat{\lambda}_i) \right]$$

::: small

Under $H_0$, $D \sim \chi^2$ distribution with $df = n - (p + 1)$.

:::

## Are means big enough for GoF test? 

```{r}
summary(predict(full_glm, type = "response"))
```


## Quasipoisson model {.smaller}

```{r}
full_quasi <- glm(laws ~ percent_urban + historical,
            data = equality_index,
            family = quasipoisson)
summary(full_quasi)
```

## Pearson resids still show overdispersion

Since they use the Poisson mean/sd

```{r}
#| fig-width: 8
#| fig-height: 3
ggResidpanel::resid_xpanel(full_quasi, type = "pearson")
```

## Standardized Pearson residuals

Use quasi versions of SD: $SD(Y_i) = \sqrt{\psi \lambda_i}$ (also adjusts for leverage)

```{r}
#| fig-width: 8
#| fig-height: 2.75
#| echo: true
ggResidpanel::resid_xpanel(full_quasi, type = "stand.pearson")
```

. . . 

&rarr; better! perhaps one outlier we'd want to investigate further

## Inference for coefficients

**Confidence interval** for $\beta$:

- For Poisson model: $\widehat \beta \pm z^* SE(\widehat\beta)$
- For Quasi model: $\widehat \beta \pm t_{n-(p+1)}^* SE_{quasi}(\widehat\beta)$

<br>

**Hypothesis testing** for $H_0: \beta = 0$ against $H_A: \beta \ne 0$

- Test stat: $\frac{\widehat \beta - 0}{SE}$
  - $\sim N(0,1)$ under Poisson model
  - $\sim t_{n-(p+1)}$ under Quasi model


## Confidence Intervals: Interpretation

<br>

We construct intervals on the **log (link) scale** first, then exponentiate:

$$\text{Log scale:} \quad \hat{\beta}_j \pm z^* \cdot SE(\hat{\beta}_j)$$
$$\text{Count/Response scale:} \quad e^{\hat{\beta}_j \pm z^* \cdot SE(\hat{\beta}_j)}$$

<br>

. . . 

&rarr; We are $(1-\alpha)\times100\%$ confident that a 1-unit increase in $X_j$ is associated with a multiplicative change in the expected count by a factor between $e^L$ and $e^U$, holding all other variables constant.

## Inference for $\lambda$

::: medium

Reminder: $\log \lambda = \eta$

1. Compute the predicted value and standard error on the **link (log) scale**:
   $$\widehat{\eta} = \widehat{\beta}_0 + \widehat{\beta}_1 X_1 + \dots + \widehat{\beta}_p X_p$$
2. Build the interval on the log scale: $[\widehat{\eta} - z^*SE, \,\, \widehat{\eta} + z^*SE]$
3. Exponentiate the endpoints to map back to the **original count scale**:
   $$\left[ e^{\widehat{\eta} - z^*SE}, \,\, e^{\widehat{\eta} + z^*SE} \right]$$

In R: 

```r
predict(model, newdata = data.frame(x = ...), type = "link", se.fit = TRUE)
predict(model, newdata = data.frame(x = ...), type = "response", se.fit = TRUE)
```

:::

## Example: `link` scale

::: medium
Minnesota historically votes `dem` and has 73.3% of the population in urban areas.


```{r}
#| echo: true

predict(full_glm, newdata = data.frame(historical = "dem", 
                                       percent_urban = 73.3), 
        type = "link", 
        se.fit = TRUE)
```

:::

$3.06 \pm 2\cdot .07 = [2.92, 3.20] \to [e^{2.92}, e^{3.20}] = [18.54, 24.53]$

## Example: `response` scale

::: medium

```{r}
#| echo: true


predict(full_glm, newdata = data.frame(historical = "dem", 
                                       percent_urban = 73.3), 
        type = "response", 
        se.fit = TRUE)
```

:::

$21.275 \pm 2\cdot 1.47 = [18.33, 24.22]$

. . . 

&rarr; We are 95% confident that the average number of anti-discrimination laws in states like Minnesota is between 18.33 and 24.22

## Likelihood Ratio Test

**Hypotheses:**

$H_0: \beta_i = \beta_j = ... = 0$

$H_A$: at least one $\ne$ 0

<br>

**Test Statistic**: 

$$\Delta  = D_{\text{Reduced}} - D_{\text{Full}}$$

**Reference distribution**:

- Depends on standard poisson vs quasipoisson

## LRT reference distribution

**Standard Poisson GLM:** We compare $\Delta$ to a $\chi^2$ distribution with $df = \Delta p$.

```{r}
#| eval: false
#| echo: true

anova(reduced_glm, full_glm, test = "Chisq")
```

<br>

**Quasipoisson GLM**: Variance has been inflated by $\widehat{\psi}$. Instead, we scale the drop in deviance and perform an F-test: 

R automatically handles the scale adjustment when you ask for test = "F":

```{r}
#| eval: false
#| echo: true
anova(reduced_quasi, full_quasi, test = "F")
```

