---
title: Intro to Multiple Linear Regression
subtitle: Day 09
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

# wildfires stolen from Adam
# wildfires <- read.csv("https://aloy.github.io/stat230-materials/data/wildfires.csv")
# wildfires <- write.csv(wildfires, here("data", "wildfires.csv"))

wildfires <- read.csv(here("data", "wildfires.csv"))
wildfires <- filter(wildfires, Year >= 1960)
```


## `bikes` again

```{r}
bikes_slr <- lm(rides ~ temp_actual, data = bikes)

equatiomatic::extract_eq(bikes_slr, use_coefs = TRUE)
```

::::: columns
::: {.column width="50%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003300", size = 1) |>
  gf_lm(rides ~ temp_actual, size = 1.5, color = "#f15a31") 
```

:::


::: {.column width="50%"}

<br> $R^2 =$ `r round(summary(bikes_slr)$r.squared, 3)`

:::
:::::

## Pairs plot

```{r}
bikes |> 
  select(temp_actual, windspeed, rides) |>
  GGally::ggpairs()
```

## Correlation matrix

```{r}
#| echo: true


bikes |> 
  select(temp_actual, windspeed, rides) |>
  cor()
```


## Multiple Linear Regression (MLR)



$$Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots + \beta_p x_{ip} + \varepsilon_i, \quad \varepsilon_i \overset{iid}{\sim} N(0, \sigma^2)$$

<br>

. . .

This model requires that...

1. $\mu\{Y|x_1, x_2, \ldots, x_p\}$ is a linear function

2. The error terms are independent 

3. The error terms follow a normal distribution with mean 0 and variance $\sigma^2$


## Fitting the model

**Procedure:**

Least squares estimation: choose the coefficients to minimize 
$$\text{SSE} = \sum_{i=1}^n \left(Y_i - \widehat{Y}_i \right)^2$$
Then use $\widehat{\sigma} = \sqrt{\frac{\text{SSE}}{n-(p+1)}}$

## Fitting the model 

**R:** 

```{r}
#| echo: true
bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)
coef(bikes_mlr)
```

## Interpreting $\widehat{\beta}_0$ {.incremental}


```{r echo=FALSE}
# knitr::kable(tidy(supervisor_lm), format = "html", digits = 4)
coef(bikes_mlr) |> round(3)
```

- The expected value of the response variable when all explanatory variables are 0

- The expected overall number of rides is about -1525.538 when temperature = 0 and windspeed = 0 🫠

## Interpreting $\widehat{\beta}_j$ {.incremental}


```{r echo=FALSE}
# knitr::kable(tidy(supervisor_lm), format = "html", digits = 4)
coef(bikes_mlr) |> round(3)
```



- The expected change in the response for a 1-unit increase in $x_j$, when all other variables are held constant.

- We expect 87.02 *additional* rides for every 1 degree increase in actual temperature, assuming constant windspeed

- We expect 38.63 *fewer* rides for every 1 MPH increase in windspeed, assuming constant temperature

## Geometric interpretation: 

MLR "line" is actually a **plane**

```{r}
library(plotly)
p <- plot_ly(bikes, x = ~temp_actual, y = ~windspeed, z = ~rides, 
             type = "scatter3d", mode = "markers", color = ~rides)
p
```

## Geometric interpretation: 

MLR "line" is actually a **plane**

```{r}
# code from gemini

# 2. Create sequences of values across the range of your raw data
x_seq <- seq(min(bikes$temp_actual), max(bikes$temp_actual), length.out = 50)
y_seq <- seq(min(bikes$windspeed), max(bikes$windspeed), length.out = 50)

# 3. Create a 2D prediction grid of all possible combinations
pred_grid <- expand.grid(temp_actual = x_seq, windspeed = y_seq)

# 4. Generate predictions from your model
z_pred <- predict(bikes_mlr, newdata = pred_grid)

# 5. Reshape the predictions into a matrix (required by add_surface)
z_matrix <- t(matrix(z_pred, nrow = length(x_seq), ncol = length(y_seq)))

# 6. Add the surface to your existing plot using the raw sequences!
p |> 
  add_surface(
    x = ~x_seq, 
    y = ~y_seq, 
    z = ~z_matrix, 
    opacity = 0.5,          # Make the plane slightly transparent
    showscale = FALSE       # Hide the color legend for the plane
  )
```


## Centering predictor variables

*Idea:* make intercept more interpretable

```{r}
p <- plot_ly(bikes, x = ~(temp_actual - mean(temp_actual)), y = ~(windspeed - mean(windspeed)), z = ~rides, 
             type = "scatter3d", mode = "markers", color = ~rides)
p
```

## Making centered variables

```{r}
#| echo: true

bikes <- bikes |>
  mutate(
    temp_centered = temp_actual - mean(temp_actual, na.rm = TRUE), 
    wind_centered = windspeed - mean(windspeed, na.rm = TRUE)
  )
```

Mean shifts, standard deviation stays the same:

```{r}
bikes |>
  summarize(
    mean_orig = mean(temp_actual),
    mean_centered = mean(temp_centered),
    sd_orig = sd(temp_actual),
    sd_centered = sd(temp_centered)
  ) |>
  round(2)
```

## Interpreting $\widehat{\beta}_0$ for centered predictors {.incremental}


```{r echo=FALSE}
bikes_mlr_c <- lm(rides ~ temp_centered + wind_centered, data = bikes)
coef(bikes_mlr_c) |> round(3)
```

- The expected value of the response variable when all explanatory variables are 0

- The expected overall number of rides is about 3,482 on an average temperature and windspeed day 😌

## Interpreting $\widehat{\beta}_j$ (doesn't change) {.incremental} 


```{r echo=FALSE}
# knitr::kable(tidy(supervisor_lm), format = "html", digits = 4)
coef(bikes_mlr_c) |> round(3)
```


- The expected change in the response for a 1-unit increase in $x_j$, when all other variables are held constant.

- We expect 87.02 *additional* rides for every 1 degree increase in actual temperature, assuming constant windspeed

- We expect 38.63 *fewer* rides for every 1 MPH increase in windspeed, assuming constant temperature

## Standardizing predictor variables

*Idea:* 1 MPH in windspeed is different than 1 degree temp change

```{r}
p1 <- gf_histogram(~temp_centered, data = bikes, col = "white", fill = "#003300")
p2 <- gf_histogram(~wind_centered, data = bikes, col = "white", fill = "#003300")
p1 + p2
```


## Making standardized variables

```{r}
#| echo: true

bikes <- bikes |>
  mutate(
    temp_std = (temp_actual - mean(temp_actual, na.rm = TRUE))/sd(temp_actual, na.rm = TRUE), 
    wind_std = (windspeed - mean(windspeed, na.rm = TRUE))/sd(windspeed, na.rm = TRUE)
  )
```

Now, mean = 0 and sd = 1

```{r}
bikes |>
  summarize(
    mean_orig = mean(temp_actual),
    mean_std = mean(temp_std),
    sd_orig = sd(temp_actual),
    sd_std = sd(temp_std)
  ) |>
  round(2)
```

## Interpreting $\widehat{\beta}_0$ for standardized predictors {.incremental}


```{r echo=FALSE}
bikes_mlr_std <- lm(rides ~ temp_std + wind_std, data = bikes)
coef(bikes_mlr_std) |> round(3)
```

- The expected value of the response variable when all explanatory variables are 0

- The expected overall number of rides is about 3,482 on an average temperature and windspeed day

## Interpreting $\widehat{\beta}_j$ for standardized predictors {.incremental} 

```{r echo=FALSE}
# knitr::kable(tidy(supervisor_lm), format = "html", digits = 4)
coef(bikes_mlr_std) |> round(3)
```


- The expected change in the response for a 1 *standard deviation* increase in $x_j$, when all other variables are held constant.

- We expect 885 *additional* rides for every 1 SD increase in actual temperature (10.2 degrees), assuming constant windspeed

- We expect 213 *fewer* rides for every 1 SD increase in windspeed (5.5 MPH), assuming constant temperature

