---
title: Inference for SLR
subtitle: Day 03
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)
```

## Data for today 

*Capital Bikeshare* is a bike sharing service in the Washington, D.C. area. We're interested in analyzing the relationship between *number of rides taken* and *temperature*

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003300", size = 1) |>
  gf_lm(rides ~ temp_actual, linewidth = 1.5, color = "#f15a31") 
```

## Last Time

1. $Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

2. Scatterplots + correlation

3. Fit the regression line in R and find a `summary()`

```{r}
#| echo: true
bikes_lm <- lm(rides ~ temp_actual, data = bikes)
```

4. Making confidence intervals using `confint`

. . . 

### Today: moving from *descriptive* to *inferential*

## `lm()` Summary {.smaller}

```{r}
summary(bikes_lm)
```

## Let's take a step back: Sampling Variability

```{r}
#| echo: false

set.seed(1234)

boots <- bootstraps(bikes, times = 20, apparent = TRUE)

fit_lm_on_bootstrap <- function(split) {
    lm(rides ~ temp_actual, data = analysis(split))
}

boot_models <-
  boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))

boot_coefs <- 
  boot_models %>% 
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

boot_aug3 <- 
  boot_models %>% 
  slice(1:3) %>% 
  mutate(augmented = map(model, augment)) %>% 
  unnest(augmented)

labels <- c("Bootstrap01" = "Sample 1", "Bootstrap02" = "Sample 2", "Bootstrap03" = "Sample 3")
```

```{r}
#| echo: false
#| fig.height: 3
#| fig.width: 5
hop <- 
  ggplot(data = boot_coefs) +
  geom_abline(aes(slope = temp_actual, intercept = `(Intercept)`, group = id), color = "#00aeef", size = 0.5) +
  geom_point(data = bikes, aes(temp_actual, rides), color = "grey60", inherit.aes = FALSE) +
  labs(x = "Temperature", y = "No. rides") +
  theme(
    strip.text = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    # plot.margin = margin(7, 1.5, 3.5, 1.5)
  ) +
  # panel_border() +
  transition_states(id, 0, 1) +
  shadow_mark(future = TRUE, color = "gray50", alpha = 1/20)

animate(hop)
```

## Sampling distribution 

```{r}
#| echo: false
#| fig.height: 3
#| fig.width: 6

boots1k <- bootstraps(bikes, times = 1000, apparent = TRUE)


boot_models_1k <-
  boots1k |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))

boot_coefs <- 
  boot_models_1k %>% 
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  rename(
    intercept = `(Intercept)`,
    slope = temp_actual
  )

slope_dist <- ggplot(boot_coefs) +
   ggdist::stat_dotsinterval(aes(x = slope), quantiles = 500, scale = 0.6, position = position_nudge(y = 0.175)) +
  labs(y = "Density")

intercept_dist <- ggplot(boot_coefs) +
  ggdist::stat_dotsinterval(aes(x = intercept), quantiles = 500, scale = 0.6, position = position_nudge(y = 0.175))  +
  labs(y = "Density")

slope_dist + intercept_dist 
```


## Sampling distribution

We use the t distribution with df =  n - 2 to model the sampling distribution of a regression coefficient

```{r}
#| echo: false
#| fig.height: 3
#| fig.width: 6

mod <- bikes_lm

sdsn_df <- tibble(
  term = c("slope", "intercept"),
  mean = c(mean(boot_coefs$slope), mean(boot_coefs$intercept)),
  df = c(62, 62),
  sd = diag(vcov(mod)) |> sqrt() |> unname() |> rev()
)



slope_slab <- 
  sdsn_df |> 
  filter(term == "slope") |>
  ggplot() +
  ggdist::stat_dotsinterval(data = boot_coefs, aes(x = slope), quantiles = 500, scale = 0.6, position = position_nudge(y = 0), inherit.aes = FALSE) +
  ggdist::stat_slab(aes(xdist = distributional::dist_student_t(df = df, mu = mean, sigma = sd)), 
                    slab_size = 0.5, 
                    fill = "#0072B2",
                    alpha = 0.4) +
  xlab("Slope")

intercept_slab <- 
  sdsn_df |> 
  filter(term == "intercept") |>
  ggplot() +
  ggdist::stat_dotsinterval(data = boot_coefs, aes(x = intercept), quantiles = 500, scale = 0.6, position = position_nudge(y = 0), inherit.aes = FALSE) +
  ggdist::stat_slab(aes(xdist = distributional::dist_student_t(df = df, mu = mean, sigma = sd)), 
                    slab_size = 0.5, 
                    fill = "#0072B2",
                    alpha = 0.4) +
  xlab("Intercept")

slope_slab + intercept_slab
```

## Sampling distribution

- t distribution with $df =  n - 2$

. . .

- Mean = $\beta_i$

. . .

- Standard deviation = $SE(\widehat{\beta}_1) = \widehat{\sigma} / \sqrt{\sum (x_i - \overline{x} )^2}$


## Confidence interval

::::{.columns}
::: {.column width="15%"}
statistic
:::
::: {.column width="7%"}
±
:::
::: {.column width="70%"}
critical value × standard error
:::
::::

<br>

::::{.columns}
::: {.column width="15%"}
$\widehat{\beta}_i$
:::
::: {.column width="7%"}
±
:::
::: {.column width="33%"}
$t^* \cdot SE(\widehat{\beta}_i)$
:::
::::


## Critical Value

::::{.columns}
::: {.column width="33%"}

```{r}
#| echo: true

qt(.025, df = 30)
```

```{r}
my_p = .025
my_q = qt(my_p, df = 30)

gf_dist("t", df = 30, color = "black", linewidth = 1) |>
  gf_dist("t", df = 30, geom = "area", fill = "#0072B2", alpha = 0.5, 
          xlim = c(-4, my_q)) |>
  gf_vline(xintercept = ~my_q, color = "#f15a31", linetype = "dashed", linewidth = 1) |>
  gf_point(0 ~ my_q, color = "#f15a31", size = 3)
```

```{r}
#| echo: true

qt(.025, df = 30, 
   lower.tail = FALSE)
```

```{r}
my_p = .025
my_q = qt(my_p, df = 30,
   lower.tail = FALSE)

gf_dist("t", df = 30, color = "black", linewidth = 1) |>
  gf_dist("t", df = 30, geom = "area", fill = "#0072B2", alpha = 0.5, 
          xlim = c(-4, my_q)) |>
  gf_vline(xintercept = ~my_q, color = "#f15a31", linetype = "dashed", linewidth = 1) |>
  gf_point(0 ~ my_q, color = "#f15a31", size = 3)
```

:::
::: {.column width="33%"}


```{r}
#| echo: true

qt(.1, df = 30)
```

```{r}
my_p = .1
my_q = qt(my_p, df = 30)

gf_dist("t", df = 30, color = "black", linewidth = 1) |>
  gf_dist("t", df = 30, geom = "area", fill = "#0072B2", alpha = 0.5, 
          xlim = c(-4, my_q)) |>
  gf_vline(xintercept = ~my_q, color = "#f15a31", linetype = "dashed", linewidth = 1) |>
  gf_point(0 ~ my_q, color = "#f15a31", size = 3)
```

```{r}
#| echo: true

qt(.1, df = 30,
   lower.tail = FALSE)
```

```{r}
my_p = .1
my_q = qt(my_p, df = 30,
   lower.tail = FALSE)

gf_dist("t", df = 30, color = "black", linewidth = 1) |>
  gf_dist("t", df = 30, geom = "area", fill = "#0072B2", alpha = 0.5, 
          xlim = c(-4, my_q)) |>
  gf_vline(xintercept = ~my_q, color = "#f15a31", linetype = "dashed", linewidth = 1) |>
  gf_point(0 ~ my_q, color = "#f15a31", size = 3)
```

:::
::: {.column width="33%"}

```{r}
#| echo: true

qt(.25, df = 30)
```

```{r}
my_p = .25
my_q = qt(my_p, df = 30)

gf_dist("t", df = 30, color = "black", linewidth = 1) |>
  gf_dist("t", df = 30, geom = "area", fill = "#0072B2", alpha = 0.5, 
          xlim = c(-4, my_q)) |>
  gf_vline(xintercept = ~my_q, color = "#f15a31", linetype = "dashed", linewidth = 1) |>
  gf_point(0 ~ my_q, color = "#f15a31", size = 3)
```

```{r}
#| echo: true

qt(my_p, df = 30, 
   lower.tail = FALSE)
```

```{r}
my_p = .25
my_q = qt(my_p, df = 30, lower.tail = FALSE)

gf_dist("t", df = 30, color = "black", linewidth = 1) |>
  gf_dist("t", df = 30, geom = "area", fill = "#0072B2", alpha = 0.5, 
          xlim = c(-4, my_q)) |>
  gf_vline(xintercept = ~my_q, color = "#f15a31", linetype = "dashed", linewidth = 1) |>
  gf_point(0 ~ my_q, color = "#f15a31", size = 3)
```

:::
::::

## Hypothesis test {.scrollable}



$H_0: \beta_i = \text{null value}$ vs. $H_a:$ choose one of $\begin{array}{c} \beta_i \ne  \text{null value}\\ \beta_i < \text{null value}  \\ \beta_i > \text{null value} \end{array}$ 

<br>

. . .

::::{.columns}
::: {.column width="30%"}
test statistic
:::
::: {.column width="70%"}
$= \dfrac{\text{estimate} - \text{null value}}{\text{SE}}$
:::
::::

::::{.columns}
::: {.column width="30%"}

:::
::: {.column width="70%"}
$=\dfrac{\widehat{\beta}_i - \text{null value}}{SE(\widehat{\beta}_i)}$
:::
::::

. . .

Find appropriate tail areas under the  t-distribution with $df = n - 2$


## Coefficient table

Standard output for a regression model includes a coefficient table

```{r}
#| echo: true
library(broom)
tidy(mod)
```


## Example: 90% confidence interval for $\beta_1$

::::{.columns}
::: {.column width="15%"}
statistic
:::
::: {.column width="7%"}
±
:::
::: {.column width="70%"}
critical value × standard error
:::
::::

<br>

::::{.columns}
::: {.column width="15%"}
$\widehat{\beta}_i$
:::
::: {.column width="7%"}
±
:::
::: {.column width="33%"}
$t^* \cdot SE(\widehat{\beta}_i)$
:::
::::

. . . 

```{r}
#| echo: true

qt(.05, df = 30)
```

<br>

<br>

::: callout-tip
## CI Interpretation

We are _______% confident that for every 1 _________ increase in __________________________, _____________________________ also increases by between ____________ and _____________

:::

# Your turn: on paper

# Intro to maize/RStudio


