---
title: Inference for Linear Combinations
subtitle: Day 12
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

bikes <- bikes |> select(date, season, year, month, weekend, holiday, temp_actual, temp_feel, humidity, windspeed, rides)

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt", sep = "\t", header = TRUE) |>
  mutate(
    Smoke = as.logical(Smoke)
  )

bikes_slr <- lm(rides ~ temp_actual, data = bikes)

fev_lm <- lm(FEV ~ Age * Smoke, data = fev)

```

## Inference for MLR

<br> 

1. Inference for individual $\beta$ coefficients
2. Inference for comparing two models
3. Inference for predictions


## FEV example

```{r echo=FALSE , fig.width = 4, fig.height = 3, fig.align='center', out.width = "60%"}
#| warning: false

age <- fev$Age
nonsmoke.mean <- function(age) 0.7574802 + 0.2028886 * age
smoke.mean <- function(age) (0.7574802 + 1.3071504) + (0.2028886 -0.1128448) * age
ggplot(data = fev, mapping = aes(x = Age, y = FEV, color = Smoke)) +
  geom_point( alpha = 0.5) +
  geom_lm(aes(group = Smoke, color = Smoke)) + 
  scale_color_manual(values = c("#f15a31", "#000036")) +
  labs(y = "FEV (in L/s)") +
  theme(legend.position = "top")


```

## Coefficient table

```{r}
fev_coefs <- fev_lm |> broom::tidy()
fev_coefs |> 
  mutate(p.value = style_pvalue(p.value)) |>
  knitr::kable()
```

## FEV example

$${\rm E}(Y|X) = \beta_0 + \beta_1 {\tt age} + \beta_2 {\tt smoke} + \beta_3 {\tt age \times smoke}$$
<br>

. . .

Mean functions for each group

. . .

:::columns
:::{.column width="30%"}
Smokers:
:::
:::{.column width="70%"}
${\rm E}(Y|X) = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) {\tt age}$
:::
:::


. . .

:::columns
:::{.column width="30%"}
Non-smokers:
:::
:::{.column width="70%"}
${\rm E}(Y|X) = \beta_0 + \beta_1  {\tt age}$
:::
:::

. . . 

::: task
**Warm up:** How would I build a confidence interval for the slope of the smokers line? 
:::

## SEs for linear combinations



Estimate: 

$$\widehat{\gamma} = c_1 \widehat{\beta}_i + c_2 \widehat{\beta}_j$$


<br> 

. . .

Standard error:

:::{style="font-size: 0.9em"}

$$
SE(\widehat{\gamma}) = \sqrt{c_1^2 \lbrace SE(\widehat{\beta}_i)\rbrace^2 + c_2^2 \lbrace SE(\widehat{\beta}_j) \rbrace^2 + 2c_1c_2 Cov(\widehat{\beta}_i, \widehat{\beta}_j)} 
$$
:::



## FEV example

:::{.callout-warning}
Our standard table of coefficients doesn't contain the SEs we need to construct CIs or perform tests for these linear combinations.
:::

```{r}
#| echo: false
fev_coefs |> 
  mutate(p.value = style_pvalue(p.value)) |>
  knitr::kable()
```

## Covariance matrix

::: medium

- The diagonal (top left to lower right) displays variances, $\lbrace SE(\widehat{\beta}_j) \rbrace^2$

- The off-diagonals display covariances, $Cov(\widehat{\beta}_i, \widehat{\beta}_j)$

- Covariance is like correlation, but not scaled to be between $-1$ and $1$

    - $r_{xy} = \frac{Cov(x,y)}{s_x s_y} \to Cov(x,y) = r s_x s_y$
:::

```{r}
#| echo: false
vcov(fev_lm)
```

## SE for a linear combination

:::{style="font-size: 0.9em;"}
Smokers: ${\rm E}(Y|X) = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) {\tt age}$

Target: $\widehat{\gamma} = \widehat{\beta}_1 + \widehat{\beta}_3$
:::
```{r}
#| echo: false
vcov(fev_lm)
```

. . .

:::{style="font-size: 0.8em;"}
$$
SE(\widehat{\gamma}) = \sqrt{c_1^2 \lbrace SE(\widehat{\beta}_1)\rbrace^2 + c_3^2 \lbrace SE(\widehat{\beta}_3) \rbrace^2 + 2c_1c_3 Cov(\widehat{\beta}_1, \widehat{\beta}_3)} 
$$
:::
. . .

:::{style="font-size: 0.8em;"}
$c_1 = 1$, $c_3 = 1$
:::

. . .

:::{style="font-size: 0.8em;"}
:::columns
:::{.column width="50%"}
$SE(\widehat{\beta_1})^2 = 0.0004423$
:::
:::{.column width="50%"}
$SE(\widehat{\beta_3})^2 = 0.0017798$
:::
:::
:::

. . .

:::{style="font-size: 0.8em;"}
$Cov(\widehat{\beta_1}, \widehat{\beta_3}) = -0.0004423$
:::

## CI for a linear combination

Once we have $SE(\widehat{\gamma})$, we can construct a CI as usual:

$$
\widehat{\gamma} \pm t^*_{n - (p+1)} \cdot SE(\widehat{\gamma})
$$

. . .

$\widehat{\gamma} = \widehat \beta_1 + \widehat \beta_3$

. . . 

```{r}
#| echo: false
V <- vcov(fev_lm)
SE_gamma <- sqrt(V[2,2] + V[4,4] + 2*V[2,4])
```


$SE(\widehat{\gamma}) =`r SE_gamma`$ 

. . . 

$df = n - (p+1) = 345 - 4 = 341$

. . .

$t^*_{341} \approx 1.649$ for a 90% CI


## Test for a linear combination

We can also test hypotheses about linear combinations of coefficients using a t-test:

$H_0: \gamma = \gamma_0$ vs. $H_a: \gamma \underset{>}{\overset{<}{\ne}} \gamma_0$


$t = \dfrac{\widehat{\gamma} - \gamma_0}{SE(\widehat{\gamma})}$

Reference distribution: $t_{n-(p+1)}$

## Your turn: smokers vs. nonsmokers at age 15

::: task
At age 15, how different is the mean FEV of smokers and nonsmokers?

1. Write the mean function for each group at $\tt age = 15$. Subtract (smoker $-$ nonsmoker) to get $\gamma$ in terms of the $\beta$'s.
2. What are $c_2$ and $c_3$? Write out $SE(\widehat{\gamma})$.
3. Construct a 95% CI for $\gamma$ and test $H_0: \gamma = 0$.
4. Interpret your results in context.
:::

```{r}
#| echo: false
vcov(fev_lm)
```

::: footer
Hint: $c_3 \ne 1$ this time. $df = 341$, $t^*_{341} \approx 1.967$ for a 95% CI.
:::

## Your turn: solution

```{r}
#| echo: false
cc <- c(0, 0, 1, 15)
g_hat <- sum(cc * coef(fev_lm))
se_g <- sqrt(drop(t(cc) %*% V %*% cc))
tstar95 <- qt(0.975, df = fev_lm$df.residual)
```

::: medium
$\gamma = (\beta_0 + \beta_2) + (\beta_1 + \beta_3)(15) - (\beta_0 + \beta_1 (15)) = \beta_2 + 15\beta_3$

. . .

$\widehat{\gamma} = `r round(coef(fev_lm)[3], 3)` + 15(`r round(coef(fev_lm)[4], 3)`) = `r round(g_hat, 3)`$

. . .

$SE(\widehat{\gamma}) = \sqrt{\{SE(\widehat{\beta}_2)\}^2 + 15^2 \{SE(\widehat{\beta}_3)\}^2 + 2(1)(15) Cov(\widehat{\beta}_2, \widehat{\beta}_3)} = `r round(se_g, 3)`$

. . .

95% CI: $`r round(g_hat, 3)` \pm 1.967(`r round(se_g, 3)`) = (`r round(g_hat - tstar95 * se_g, 3)`, `r round(g_hat + tstar95 * se_g, 3)`)$

. . .

$t = `r round(g_hat / se_g, 2)`$, p-value $= `r format(2 * pt(-abs(g_hat / se_g), fev_lm$df.residual), digits = 2)`$

. . .

At age 15, smokers' mean FEV is estimated to be about `r abs(round(g_hat, 2))` L/s lower than nonsmokers' (95% CI: `r abs(round(g_hat + tstar95 * se_g, 2))` to `r abs(round(g_hat - tstar95 * se_g, 2))` L/s lower). The difference is statistically significant.
:::


