---
title: Binomial Logistic Regression
subtitle: Day 22
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```

## `moths` from last homework

```{r}
moths <- read.csv("https://aluby.github.io/stat230/data/moths.csv") |>
  mutate(notremoved = Placed - Removed, 
               log_odds_removed = log(Removed / notremoved),
               prop_removed = Removed / Placed, 
               dark = ifelse(Morph=="dark",1,0), 
               distance = Distance, 
               morph = Morph) 

moths |>
  select(Morph, Distance, Placed, Removed)
```
 
## [Moth Coloration and Natural Selection]{style="color:#FFFFFF"} {background-image="http://catherinephamevolution.weebly.com/uploads/4/9/7/3/49739619/189059_orig.jpg" background-size="cover" background-position="bottom" style="color: "}

## What is the Y variable?

```{r}
#| echo: true
model1 <- lm(log_odds_removed ~ distance + morph + distance * morph, 
             data = moths)
```

```{r}
tidy(model1)
```

## Why doesn't `glm` work?

```{r}
#| echo: true
#| error: true
model2 <- glm(Removed ~ distance + morph + distance * morph, 
              data = moths, 
              family = "binomial")
```


## Binomial response

- $Y_i$: the number of moths removed (i.e. "success") on each tree for each morph

- We will **assume** each trial (success/failure) is independent

::: callout-note
## Binomial distribution

The sum of independent and identically distributed success/failure trials follows a Binomial(n, p) distribution.

:::

## Binomial logistic regression

$$\pi_i = P(success | x_1, ..., x_3)$$

Log-odds scale: 

$$\text{logit}(\pi_i) = \log (\frac{\pi_i}{1-\pi_i}) = \beta_0 + \beta_1 x_1 + ... + \beta_p x_p = \eta_i$$

Probability scale: 

$$\pi_i = \dfrac{e^{\eta_i}}{1 + e^{\eta_i}}= \dfrac{e^{\beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \cdots + \beta_p x_{pi}}}{1 + e^{\beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \cdots + \beta_p x_{pi}}}$$

## Fitted model

::: small

```{r}
equatiomatic::extract_eq(model1, use_coefs = TRUE, coef_digits = 3)
```
:::


```{r echo=FALSE, fig.height = 3, fig.width=5.5, out.width = 700, fig.align='center'}
gf_point(log_odds_removed ~ Distance, color =~morph, data = moths) |>
  gf_lm() |>
  gf_refine(scale_color_colorblind())
```

## 

This approach required calculating the empirical log-odds "by hand" and adding to the dataset.  Alternatively, we can fit with `glm()` using the `weights` argument

```{r}
#| echo: true
#| error: true
model3 <- glm(Removed/Placed ~ distance + morph + distance * morph, 
              weights = Placed,
              data = moths, 
              family = "binomial")
```

```{r}
tidy(model3)
```

::: footer
The estimates are slightly different due to different assumptions (eg $\varepsilon$)
:::

## Binomial model assumptions

- **Binomial counts**

- **Independence**: observations are independent

- **Linearity**: the log odds is a linear function of the predictors

- **Variance structure**: the variance of a binomial random variable is $n \pi (1 - \pi)$

    $\Longrightarrow$ so the variance of $Y_i$ is $n_i \pi_i (1 - \pi_i)$ (not constant!)
    
## Pearson residuals

$${\rm Pres}_i = \dfrac{Y_i - n_i \widehat{\pi}_i}{\sqrt{n_i \widehat{\pi}_i (1 - \widehat{\pi}_i)}}$$

For large enough $n_i$, Pearson residuals tend to behave like they come from $N(0,1)$

How to use?

- Plot against fitted values, predictors

- Check for curvature, outliers

## Pearson residuals

```{r}
p1 <- ggResidpanel::resid_panel(model3, type = "pearson", plots = c("resid", "qq"), smoother = TRUE)
p2 <- ggResidpanel::resid_xpanel(model3, type = "pearson", smoother = TRUE, nrow = 1)

p1 / p2
```

## Deviance residuals

::: medium

${\rm Dres}_i = {\rm sign}(Y_i - n_i \widehat{\pi}_i) \sqrt{2 \left[ Y_i \log \left( \dfrac{Y_i}{n_i \widehat{\pi}_i} \right) + (n_i - Y_i) \log \left( \frac{n_i - Y_i}{n_i - n_i \widehat{\pi}_i} \right) \right]}$

:::

For large enough $n_i$, deviance residuals tend to behave like they come from $N(0,1)$

How to use?

- Plot against fitted values, predictors

- Check for curvature, outliers

## Deviance residuals

```{r}
p1 <- ggResidpanel::resid_panel(model3, type = "deviance", plots = c("resid", "qq"), smoother = TRUE)
p2 <- ggResidpanel::resid_xpanel(model3, type = "deviance", smoother = TRUE, nrow = 1)

p1/p2
```
