---
title: Transformations and polynomial regression
subtitle: Day 08
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

# wildfires stolen from Adam
# wildfires <- read.csv("https://aloy.github.io/stat230-materials/data/wildfires.csv")
# wildfires <- write.csv(wildfires, here("data", "wildfires.csv"))

wildfires <- read.csv(here("data", "wildfires.csv"))
wildfires <- filter(wildfires, Year >= 1960)
```

## 

<br> 

### Exam01: SLR

  - Inference 
  - Conditions
  - (Simple) Transformations
  
<br> 

### Where we're going

- More complicated transformations
- Additional predictors
- Different types of response variables


## Warm up: transformations

::: {.task}
Consider the following two models: 

- $\log_{2} (y) = 1 + 2 \log_{2}(x) + \varepsilon, \varepsilon \sim N(0,1)$
- $y = 2 x^2+ \varepsilon, \varepsilon \sim N(0,1)$

Show that the two models have the same $\widehat y$'s

:::

::: callout-tip
## Hint: log rules might be useful

:::: {.columns}

::: {.column width="45%"}
- $\log(a + b) = \log(a) \log(b)$
- $\log(a^b) = b \log (a)$
:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}
- $\log(a - b) = \log(a) /\log(b)$
- $\log_{b} b = 1$ or $\log e = 1$
:::

::::
:::

## 

## 

```{r}
#| fig-width: 8
#| fig-height: 3
set.seed(041526)
tibble(x = seq(.1, 2, len = 100),
       y1 = 2^(1 + 2 * log10(x) + rnorm(100)), 
       y2 = 2*x^2 + rnorm(100)) |>
  pivot_longer(-x) |>
  ggplot(aes(x = x, y = value), col = "#003300") + 
  geom_point() + 
  facet_grid(cols = vars(name)) + 
  theme(legend.position = "none")
```

Even though they give the same **predictions**, Model (2) has constant variance but Model (1) does not. If we use Model (2) to make prediction intervals for data that look like Model (1), our interval will be too wide for small x and too narrow for large x


## Wildfires

- The National Interagency Coordination Center at the National Interagency Coordination Center compiles annual wildland fire statistics for federal and state agencies. 

- This information is provided through Situation Reports, which have been in use for several decades.

- Our goal is to model the number of acres burned over the years


::: footer
Data source: https://www.nifc.gov/fireInfo/fireInfo_stats_totalFires.html
:::

##

```{r fig.height=4, fig.width=5, message=FALSE, out.width=800}
#| echo: true


gf_point(Acres ~ Year, data = wildfires) %>%
  gf_labs(x = "Year", y = "Acres burned") %>%
  gf_smooth(size = .5)
```


## Option 1: SLR model


$\mu \lbrace y | x \rbrace = \beta_0 + \beta_1 x$

Is the fit reasonable?


```{r fig.height=2.75, fig.width=9, message=FALSE, echo=FALSE}
linear_mod <- lm(Acres ~ Year, data = wildfires)
regression_panel(Acres ~ Year, data = wildfires, xlab = "Year", ylab = "Acres burned")
```


## Option 2: Transform X


$\mu \lbrace y | x \rbrace = \beta_0 + \beta_1 x^2$

Is the fit reasonable?


```{r fig.height=2.75, fig.width=9, message=FALSE, echo=FALSE}
tform_mod <- lm(Acres ~ I(Year^2), data = wildfires)
regression_panel(Acres ~ I(Year^2), scat.formula = Acres ~ Year, data = wildfires, xlab = "Year", ylab = "Acres burned")
```

## Option 3: Polynomial model


$\mu \lbrace y | x \rbrace = \beta_0 + \beta_1 x + \beta_2x^2$

Is the fit reasonable?


```{r fig.height=2.75, fig.width=9, message=FALSE, echo=FALSE}
quad_mod <- lm(Acres ~ Year + I(Year^2), data = wildfires)
regression_panel(Acres ~ Year + I(Year^2), scat.formula = Acres ~ Year, data = wildfires, xlab = "Year", ylab = "Acres burned")
```


## The polynomial regression model


$$Y_i = \beta_0 + \beta_1 x_{i} + \beta_2 x_{i}^2 + \cdots + \beta_k x_{i}^k + \varepsilon_i, \quad \varepsilon_i \overset{iid}{\sim} N(0, \sigma)$$

Assumptions — same as in SLR

1. $\mu\{Y|x_i\}$, is a linear function

1. For each $x_i$, the sub-population of responses is normally distributed 

1. The standard deviation for each sub-population is $\sigma$

1.  Independent observations


## Interpreting the model

::: columns
:::{.column width="50%"}

```{r fig.height=3.25, fig.width=4, message=FALSE, echo=FALSE}
gf_point(Acres ~ Year, data = wildfires, xlab = "Year", ylab = "Acres burned", color = "#003069") %>%
  gf_lm(formula = y ~ x + I(x^2), color = "#f15a31")
```
:::
:::{.column width="50%"}
Focus on the expected change in $y$ for a specific increase in $x$

- e.g. change in acres burned from 1975 to 1976
    
- e.g. change in acres burned from 2010 to 2011
:::
:::

::: {style="font-size: 0.8em;"}
\begin{aligned}
\mu\{ y | x + 1 \} - \mu\{ y | x \} 
  &=  \left[ \beta_0 + \beta_1 (x+1) + \beta_2 (x+1)^2 \right] - \left[ \beta_0 + \beta_1 x + \beta_2 x^2 \right]\\
  &= \beta_1 + \beta_2 \left( 2x + 1 \right)
\end{aligned}

:::


## Inferences about coefficients

Inference uses the same t-based tools as SLR, but with 

$$\widehat{\sigma}  = \sqrt{\dfrac{{\sum_{i=1}^n e_i^2}}{n - (k+1)}}$$

i.e. the degrees of freedom of the t-distribution change



## Testing a single coefficient

::: {style="font-size: 0.9em;"}
**Hypotheses:** $H_0: \ \beta_j = \#$ vs. $H_a: \ \beta_j \underset{>}{\overset{<}{\ne}} \#$
:::

. . .

::: {style="font-size: 0.9em;"}
**Test statistic:** $t = \dfrac{\hat{\beta}_j - \#}{SE(\beta_j)}$
:::
. . .

::: {style="font-size: 0.9em;"}
**Reference distribution:** $t$ distribution with d.f. = $n-(k+1)$


**p-value:** Area in the tail(s) specified by $H_a$
:::


```{r echo=FALSE, fig.height = 2, fig.width = 7.5, out.width=850}
p1 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] < 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p2 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] != 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p3 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] > 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

(p1 | p2 | p3)
```

## CIs for a single coefficient


$\widehat{\beta}_j \pm t^*_{n-(k+1)} \cdot SE(\widehat{\beta}_j )$


. . .

<br>

#### Wildfires example

::: {style="font-size: 0.8em;"}

```{r echo=FALSE, results='asis'}
coef_table <- tidy(quad_mod, conf.int = TRUE) |>
  select(-statistic, -p.value)


knitr::kable(coef_table, format = "pipe", digits = 0, booktabs = TRUE, col.names = c("Term", "Estimate", "SE", "Lower", "Upper"))
```

:::


## Your turn

::: task 

Would a higher-order polynomial (e.g. cubic, quartic, quintic) provide a better fit to the wildfire data?

<br>

Work through the R handout with your groups
:::

## A warning about this analysis


> Prior to 1983, sources of these figures are not known, or cannot be confirmed, and were not derived from the current situation reporting process. As a result the figures prior to 1983 should not be compared to later data.
>
>— NIFC

::: footer
https://www.nifc.gov/fireInfo/fireInfo_stats_totalFires.html]
:::


## A warning about polynomials


High-order polynomial regression models will over-fit your data (i.e. pick up on peculiarities specific to your one sample from the population)


```{r echo=FALSE, fig.height = 2.5, fig.width = 7.5, out.width=850}
poly2 <- makeFun(quad_mod)
p1 <- gf_point(Acres ~ Year, data = wildfires, color = "#003069") %>%
  gf_labs(x = "Year", y = "Acres burned") %>%
  gf_fun(poly2(x) ~ x, color = "#f15a31") %>%
  gf_labs(title = "2nd-order polynomial") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

poly5_mod <- lm(Acres ~ poly(Year, 5), data = wildfires)
poly5 <- makeFun(poly5_mod)
p5 <- gf_point(Acres ~ Year, data = wildfires, color = "#003069") %>%
  gf_labs(x = "Year", y = "Acres burned") %>%
  gf_fun(poly5(x) ~ x, color = "#f15a31")  %>%
  gf_labs(title = "5th-order polynomial") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

poly25_mod <- lm(Acres ~ poly(Year, 25), data = wildfires)
poly25 <- makeFun(poly25_mod)
p25 <- gf_point(Acres ~ Year, data = wildfires, color = "#003069") %>%
  gf_labs(x = "Year", y = "Acres burned") %>%
  gf_fun(poly25(x) ~ x, color = "#f15a31") %>%
  gf_labs(title = "25th-order polynomial") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

(p1 | p5 | p25)
```


# Explained variance (aka $R^2$ again)


## The "null" model

- Use the mean of Y as the prediction for all observations

- $\mu \lbrace Y | X \rbrace = \beta_0$

- Leaves a lot of variability unexplained

::: columns
:::{.column width="50%"}
```{r echo=FALSE, fig.height = 3, fig.width = 3.5, out.width="90%", fig.align='center'}
gf_point(Acres ~ Year, data = wildfires, ylab = "Acres burned", color = "#003069") %>%
  gf_hline(yintercept = mean(wildfires$Acres), color = "#f15a31")
```
:::
:::{.column width="50%"}
$SD(Y) = `r round(sd(wildfires[["Acres"]]), 2)`$ cm
:::
:::

## Polynomial model

- $\mu \lbrace Y | X \rbrace = \beta_0 + \beta_1x + \beta_2 x^2$

- Using a predictor explains more variability

<br>


::: columns
:::{.column width="50%"}
```{r echo=FALSE, fig.height = 3, fig.width = 3.5, out.width="90%", fig.align='center'}
gf_point(Acres ~ Year, data = wildfires, ylab = "Acres burned", color = "#003069") %>%
  gf_lm(formula = y ~ poly(x, 2), color = "#f15a31")
```
:::
:::{.column width="50%"}
$SD(\varepsilon) = `r round(sigma(quad_mod), 2)`$ cm
:::
:::



## SSTotal

::: columns
:::{.column width="50%"}
${\rm SSTotal} = \sum(Y_i - \bar{Y})^2$

Measures the overall variability in $Y$
:::

::: {.column width="50%"}
```{r echo=FALSE, fig.height = 3.25, fig.width = 3.75, fig.align='center'}
quad_aug <- augment(quad_mod)
quad_aug <- mutate(quad_aug, mean.y = mean(Acres))

gf_point(Acres ~ Year, data = quad_aug, ylab = "Acres burned") %>%
  gf_hline(yintercept = mean(quad_aug$Acres), color = "#f15a31") %>%
  gf_segment(mean.y + Acres ~ Year + Year, color = "gray70") 
```
:::
:::


## SSResidual (aka SSError)

::: columns
::: {.column width="50%"}
${\rm SSResidual} = \sum(Y_i - \widehat{Y}_i)^2$

<br>

Measures the variability unexplained by the model

<br>

 $\widehat \sigma = \sqrt{\text{SSResidual}/\text{df}}$

:::

::: {.column width="50%"}
```{r echo=FALSE, fig.height = 3.25, fig.width = 3.75, fig.align='center'}

gf_point(Acres ~ Year, data = quad_aug, ylab = "Acres burned") %>%
  gf_lm(formula = y ~ poly(x, 2), color = "#f15a31") %>%
  gf_segment(.fitted + Acres ~ Year + Year, color = "gray70") 
```
:::
:::


## SSRegression

Measures the variability explained by the model

<br>

::: columns
:::{.column width="30%"}
```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
quad_anova <- anova(quad_mod)

ssdf <- tribble(
  ~Source, ~SS,
  "Error",      sigma(quad_mod)^2 * quad_anova$Df[3],
  "Total",       sum(quad_anova$`Sum Sq`),
  "Regression", sum(quad_anova$`Sum Sq`[1:2])
)

sst_plot <- ggplot(data = ssdf |> filter(Source == "Total"), aes(x = "", y = SS)) +
  geom_bar(stat="identity", width=1) +
  theme_void() +
  labs(title = "SSTotal")


decomp_plot <- ggplot(data = ssdf |> filter(Source != "Total"), aes(x = "", y = SS, fill = Source)) +
  geom_bar(stat="identity", width=1) +
  geom_text(aes(label = paste0(Source)), position = position_stack(vjust = 0.5), color = "white") +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#f15a31", "#003069"))

sst_plot + decomp_plot
```
:::
:::{.column width="70%"}
<br>

<br>

<br>

${\rm SSRegression} = SSTotal - SSError$
:::
:::

## Coefficient of Determination: R<sup>2</sup>

Proportion of the total variation in $y$ explained by the linear regression model

::: columns
:::{.column width="30%"}
```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
sst_plot + decomp_plot
```
:::
:::{.column width="70%"}
<br>

<br>



$\begin{split}
R^2 &= `r round(glance(quad_mod)[["r.squared"]], 3)`\\
&= \dfrac{SSRegr}{SST} \\
&= 1 - \dfrac{SSE}{SST} 
\end{split}$
:::
:::


## ⚠️ Caution {.center}

$R^2$ only addresses how close the fitted values are to the data, on average. It says nothing about the validity of the model.


