---
title: Model Selection
subtitle: Day 15
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

```

## MLR so far

- Interpreting coefficients
- Writing lines for different groups (categorical predictors) 
- Inference for a single coefficient
- Inference for linear combinations of coefficients
- F tests
- Influential points and diagnostics (leverage, standardized residuals, cook's distance)
- Multicollinearity 

. . . 

Today &rarr; putting it all together

## Idea: how do we decide what variables to include? {.smaller}

```{r}
bikes |> head(10) |> knitr::kable()
```

## Review: F test and sum of squares


 
## Sequential ANOVA (the `anova()` function)

::: medium

When you use `anova(my_lm)` on a single model, R provides a table of **extra Sum of Squares (SS)** for terms added sequentially


```{r}
#| echo: true
model_1 <- lm(rides ~ temp_actual + windspeed + weekend + temp_actual:weekend + windspeed:weekend, data = bikes)

anova(model_1)
```

:::

## Breaking down the `anova()` table

::: {.medium .nonincremental}

The First Row: `temp_actual`  

  - Df: Including this variable adds 1 term to the model.  
  - Sum Sq: $SS_{reg}(\text{temp_actual})$.  
  
The Second Row: `windspeed` 

  - Sum Sq: Represents the extra SS for adding `windspeed` to a model already containing `temp_actual`.  
  - $SS_{reg}(\text{temp_actual, windspeed}) - SS_{reg}(\text{temp\actual})$  


The Third Row: `weekend`   

  - Sum Sq: Represents the extra SS for adding `weekend` to a model already containing `temp_actual` and `windspeed`.  
  - $SS_{reg}(\text{temp_actual, windspeed, weekend}) - SS_{reg}(\text{temp_actual, windspeed})$  
  
:::

## Confirm results with a nested F test

"intermediate" p values in the table may not be valid; confirm proposed and full model with a nested F test:

```{r}
#| echo: true


model_2 <- lm(rides ~ temp_actual + windspeed + weekend, data = bikes)

anova(model_1, model_2)
```


## Variable selection: prediction vs understanding 

:::: {.columns}

::: {.column width="45%"}
**Prediction:**

- More complex models might be okay if you have a lot of data.  

- The model doesn't need to "make sense" as much if it predicts well.  

- Watch out for overfitting!

- Stat270
:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}
**Understanding:**

- Complexity can hurt our ability to interpret results meaningfully.

- Keep effects you are interested in, even if they are insignificant, to "control" for them.
:::

::::

## Comparing non-nested models

What if I want to compare:

```{r}
#| echo: true


model_2 <- lm(rides ~ temp_actual + windspeed + weekend, 
              data = bikes)
```

```{r}
#| echo: true


model_3 <- lm(rides ~ temp_actual + windspeed*weekend + 
                holiday + temp_feel , 
              data = bikes)
```

## Comparing non-nested models

Adjusted $R^2$

  - can be used for nested and non-nested models
  - no formal test/rules for "better enough" 
    
```{r}
#| echo: true


summary(model_2)$adj.r.squared
summary(model_3)$adj.r.squared
```

## Information criteria

The following "information criteria" can be used to compare nested or non-nested models (smaller values are better)

<br>

$$AIC =  n \log(SSE/n) + 2p$$

<br>

$$BIC =  n \log(SSE/n) + \log(n) \cdot p$$

## AIC in R

```{r}
#| echo: TRUE

AIC(model_2)
AIC(model_3)
```

- suggests more complicated model is worth the added complexity

## BIC in R

```{r}
#| echo: TRUE

BIC(model_2)
BIC(model_3)
```

- BIC imposes a larger complexity penalty, so prefers the simpler model 

## AIC vs BIC

<br>

**AIC:** for large $n$, gives an estimate of out-of-sample predictive performance (i.e. better model for **prediction**)

<br>

**BIC:** for large $n$, will pick the "true" model if it is one of the candidates (i.e. better model for **understanding**)


## Prediction: Overfitting 

::: medium

The goal when building a model is to capture the trends in the data while still making something that can applied to new data. When a model is **overfit**, a model will perform very well on the data used to fit it but perform very poorly on any new data.


Model building is like making a suit that can generally fit everyone around a given size, not making a custom-tailored suit that can only fit one person perfectly.

:::

## Prediction: Training & Test Datasets   

If the goal is prediction, we test the model on data it hasn't seen:

  - Training dataset: Used to estimate ("train") your model.  
  - Test dataset: The data held out to test your predictions.  
  - RMSE: Root Mean Squared Error measures the average size of error (squared residuals).
  
$$RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}$$

## Code: make training and testing data

```{r}
#| echo: true

set.seed(050726)

bikes_split <- initial_split(bikes, prop = .8)
bikes_split

bikes_train <- training(bikes_split)
bikes_test <- testing(bikes_split)
```

## Code: fit both models on training data

```{r}
#| echo: true


model_1_train <- lm(rides ~ temp_actual + windspeed + weekend + 
                      temp_actual:weekend +  windspeed:weekend,
                     data = bikes_train)
model_3_train <- lm(rides ~ temp_actual + windspeed * weekend + 
                      holiday + temp_feel,
                      data = bikes_train)
```


## Code: find $\widehat{y}$ for both models, then RMSE

```{r}
#| echo: true


model_1_test <- predict(model_1_train, newdata = bikes_test)
model_3_test <- predict(model_3_train, newdata = bikes_test)

sqrt(mean((bikes_test$rides - model_1_test)^2))
sqrt(mean((bikes_test$rides - model_3_test)^2))
```

## Code: find RMSE for **training data**

```{r}
#| echo: true


sqrt(mean(model_1_train$residuals^2))
sqrt(mean(model_3_train$residuals^2))
```

::: medium

- RMSE is smaller on the model 3 for training data (predicting better)
- RMSE is **bigger** using the model 3 for testing data (predicting worse)
    - when the model is used on data it hasn't seen before, it's doing worse
    - this is what it means to overfit
:::


## Tools for Model Comparison 

::: medium

1. ANOVA F-tests: Used for nested models to test if coefficients equal zero.  
2. Adjusted $R^2$: Can be used for non-nested models; penalizes adding unnecessary predictors.
3. AIC or BIC
    - Smaller values are better.  
    - Penalizes high numbers of coefficients to prevent over-complexity
    
Use on the full data if you are building an explanatory model, or on the training data *only* if you are building a predictive model 

:::

## Inference after selection

- Model selection results in keeping terms with smaller p-values
- Actual Type I error may be much bigger than than "expected" $\alpha = .05$
- This is an active area of research! 



## A Note on Automatic Selection

I do **not** recommend using automated "Stepwise" (Forward/Backward) selection.  

- Why? They are easily abused and ignore domain knowledge, outliers, and collinearity.  
- Better Approach:

   > "The single most important tool in selecting a subset of variables is the analyst’s knowledge of the area under study." — Sandy Weisberg. 
  
## Summary of Variable Selection

- Identify research goals and use domain knowledge.  
- Perform EDA and consider transformations.  
- Compare models using ANOVA/Adj $R^2$/BIC (for nested) or AIC/RMSE (for prediction).  
- **Justify your model choice!**
