---
title: Simple Linear Regression (SLR)
subtitle: Day 02
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)
```


# Regression in Intro

## 

::::: columns
::: {.column width="60%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003300", size = 1) |>
  gf_lm(rides ~ temp_actual, size = 1.5, color = "#f15a31") 
```
:::

::: {.column width="40%"}

<br>

-   Describe the scatterplot
-   Write the equation of the regression line
-   Interpret the slope and intercept
-   Make a prediction
:::
:::::

## Q: Describe the scatterplot

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 70%
#| echo: false
 gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003300", size = 1) # |>
#   gf_lm(rides ~ temp_actual, size = 1.5, color = "#f15a31") 
```

::::: columns
::: {.column width="50%"}

:::{.small}
::: {.conceptest}
1. There is a strong, positive, linear association between temperature and number of rides.

2. There is a moderate, positive, linear association between temperature and number of rides.
:::
:::

:::

::: {.column width="50%"}


:::{.small}
::: {.conceptest}
3. There is a moderate, positive, non-linear association between temperature and number of rides.

4. There is a moderate, positive, linear association between temperature and number of rides, but there are a few extreme outliers.
:::
:::

:::
:::::


## Q: Write the equation of the regression line

```{r}
#| echo: true


lm(rides ~ temp_actual, data = bikes) 
```

What is the equation of the regression line?

<br>


## Q: Interpret the slope in context

::: {.conceptest}
1. For every one degree increase in temperature, the model predicts an average decrease of 89.23 rides

1. For every one-degree decrease in temperature, the model predicts an average increase of 89.23 rides

1. If the daily temperature is 0, we'd expect 89.23 rides

1. If the temperature increases by one degree, it will cause an increase of 89.23 rides every day 
:::


## Q: Make a prediction

The high temperature forecast tomorrow is 72 degrees F. What does the model predict for ridership?


# Regression in Stat230

## Simple Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| echo: false


library(ggdist)

bike_lm <- lm(rides ~ temp_actual, data = bikes)
bike_dsn <- data.frame(
  temp_actual = c(50, 60, 70, 80),
  yhat = predict(bike_lm, data.frame(temp_actual = c(50, 60, 70, 80))),
  resid_sd = rep(sigma(bike_lm), 4)
)


ggplot(data = bikes, aes(x = temp_actual, y = rides)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#003300") +
  geom_lm(color = "#f15a31") +
  stat_slab(aes(x = temp_actual, ydist = "norm", arg1 = yhat, arg2 = resid_sd), data = bike_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  xlab("Temperature") +
  ylab("No. rides")
```

## A deterministic model

- $Y_i = \beta_0 + \beta_1 x_i$

- No uncertainty!

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

bikes |>
  ggplot(aes(x = temp_actual, y = rides)) +
  # geom_point(alpha = 0.6, color = "#003069") +
  geom_lm(color = "#f15a31") +
  labs(x = "Temperature", y = "No. rides") 
```

## Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

 bike_lm <- lm(rides ~ temp_actual, data = bikes)
 bike_dsn <- data.frame(
   temp_actual = c(50, 60, 70, 80),
   yhat = predict(bike_lm, data.frame(temp_actual = c(50, 60, 70, 80))),
   resid_sd = rep(sigma(bike_lm), 4)
 )

# mod <- lm(Head ~ Mass, data = blue_jays_male)
# 
# blue_jay_dsn <- 
#   tibble(
#     Mass = c(60, 65, 70, 75, 80),
#     yhat = predict(mod, newdata = data.frame(Mass = c(60, 65, 70, 75, 80))),
#     resid_sd = sigma(mod)
#   )

bikes |>
  ggplot(aes(x = temp_actual, y = rides)) +
  stat_slab(aes(x = temp_actual, ydist = "norm", arg1 = yhat, arg2 = resid_sd), data = bike_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  geom_lm(color = "#f15a31") +
  labs(x = "Temperature", y = "No. rides") 
```

## Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

predict_head <- function(n, x) {
  rnorm(n, mean = predict( bike_lm , data.frame(temp_actual = x)), sd = sigma( bike_lm ))
}

set.seed(1234)

mod_sims <- tibble(
  temp_actual = rep( c(50, 60, 70, 80), each = 20),
  rides = c(predict_head(20, 50), predict_head(20, 60), predict_head(20, 70), predict_head(20, 80))
)

p <- bikes |>
  ggplot(aes(x = temp_actual, y = rides)) +
  stat_slab(aes(x = temp_actual, ydist = "norm", arg1 = yhat, arg2 = resid_sd), 
            data = bike_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  geom_point(data = mod_sims, aes(temp_actual, rides), inherit.aes = FALSE, alpha = 0.4, color = "#003300") +
  geom_lm(color = "#f15a31") +
  labs(x = "Temperature", y = "No. rides") 
p
```

## Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

bikes |>
  ggplot(aes(x = temp_actual, y = rides)) +
  geom_point(alpha = 0.6, color = "#003300") +
  geom_lm(color = "#f15a31") +
  labs(x = "Temperature", y = "No. rides") 
```

## Simple Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br>

-   Linear relationship between $x$ and $y$
-   Errors are independent and identically distributed (iid)
-   Errors are normally distributed
-   Errors have mean 0
-   Variance of the errors doesn't depend on $x$

## Notation

Mean function:

$\quad E(Y|X)=\mu\{Y|X\} = \beta_0 + \beta_1 x_i$

<br>

Fitted model equation:

$\quad \widehat{Y}_i = \widehat{\mu}\{Y|X\} = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$

## Uncertainty in the parameters

```{r}
#| echo: false
#| fig.height: 2.5
#| fig.width: 6

set.seed(1234)

boots <- bootstraps(bikes, times = 20, apparent = TRUE)

fit_lm_on_bootstrap <- function(split) {
    lm(rides ~ temp_actual, data = analysis(split))
}

boot_models <-
  boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))

boot_coefs <- 
  boot_models %>% 
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

boot_aug3 <- 
  boot_models %>% 
  slice(1:3) %>% 
  mutate(augmented = map(model, augment)) %>% 
  unnest(augmented)

labels <- c("Bootstrap01" = "Sample 1", "Bootstrap02" = "Sample 2", "Bootstrap03" = "Sample 3")

ggplot(boot_aug3, aes(temp_actual, rides)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = .fitted, group = id), col = "#00aeef") +
  facet_wrap(~id, labeller = as_labeller(labels)) +
  labs(x = "Temperature", y = "No. rides") 
```

## Uncertainty in the parameters

```{r}
#| echo: false
#| fig.height: 3
#| fig.width: 5
hop <- 
  ggplot(data = boot_coefs) +
  geom_abline(aes(slope = temp_actual, intercept = `(Intercept)`, group = id), color = "#00aeef", size = 0.5) +
  geom_point(data = bikes, aes(temp_actual, rides), color = "grey60", inherit.aes = FALSE) +
  labs(x = "Temperature", y = "No. rides") +
  theme(
    strip.text = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    # plot.margin = margin(7, 1.5, 3.5, 1.5)
  ) +
  # panel_border() +
  transition_states(id, 0, 1) +
  shadow_mark(future = TRUE, color = "gray50", alpha = 1/20)

animate(hop)
```

## Uncertainty in the parameters

```{r}
#| echo: false
#| fig.width: 5
#| fig.height: 3
ggplot(data = boot_coefs) +
  geom_abline(aes(slope = temp_actual, intercept = `(Intercept)`, group = id), color = "#0072B2", size = 0.5, alpha = 0.2) +
  # theme_dviz_grid() +
  geom_point(data = bikes, aes(temp_actual, rides), color = "grey60", inherit.aes = FALSE) +
  theme(
    # strip.text = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.margin = margin(7, 1.5, 3.5, 1.5)
  ) +
  labs(x = "Temperature", y = "No. rides") 
```

## Rest of today

::: {.task .nonincremental}
- Open the SLR in R activity on Moodle and complete in your groups
- Make note of any questions
- Choose someone responsible for reporting out your answers
:::
