---
title: Model Comparison for Logistic Regression
subtitle: Day 21
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```

## Model from last time 

::: small
```{r}
equatiomatic::extract_eq(news_mod, intercept = "beta", terms_per_line = 2,
                         swap_var_names = c("type_num" = "type", "title_words" = "words",
                                            "title_has_excl" = "excl", "title_caps_percent" = "pct_caps"))
```
:::

```{r}
tidy(news_mod)
```

## Do we need an interaction term between `title_caps_percent` and `has_excl`? 

 - EDA 
 
 - Formal Test
 
## EDA (original data)

```{r}
gf_jitter(type_num ~ title_caps_percent, color = ~has_excl, data = news, height = .25, width = 1, alpha = .5)
```

## EDA (aggregated data)

First, make an "aggregated dataset" 

```{r}
#| output-location: slide
#| echo: true
news_agg <- news |>
  mutate(tcp_bin = ntile(title_caps_percent, 13)) |>
  group_by(tcp_bin, has_excl) |>
  summarize(
    prop_fake = mean(type_num),
    log_odds = log(prop_fake/(1-prop_fake))
  )

news_agg
```

## Plot empirical log odds

```{r}
#| echo: true
#| fig-width: 8
#| fig-height: 4
gf_point(log_odds ~ tcp_bin, color = ~has_excl, data = news_agg) |>
  gf_lm()
```

- Is there a nonlinear relationship? 
- Should we try a different lines model?

## Model with interaction

```{r}
#| echo: true
news_mod2 <- glm(type_num ~ title_words + 
                   has_excl*title_caps_percent, 
                data = news, family = "binomial")
```

::: xsmall

```{r}
equatiomatic::extract_eq(news_mod2, intercept = "beta", terms_per_line = 2,
                         swap_var_names = c("type_num" = "type", "title_words" = "words",
                                            "title_has_excl" = "excl", "title_caps_percent" = "pct_caps"))
```

:::

:::medium
```{r}
tidy(news_mod2)
```
:::

## 

::: task
Write out regression model for each group separately
:::


## 

::: task
Do we need the interaction term? Write out null and alternative hypothesis. 
:::

## Solution: 

## 
::: task
Do we need `has_excl` at all? Write out the null and alternative hypothesis for this question.
:::

## Likelihood Ratio Test

In OLS, we used: 

For logistic regression, we use a **likelihood ratio test**

::: medium
```{r}
#| echo: true
#| eval: false

anova(reduced_model, full_model, test = "LRT")
```


```{r}
#| echo: true

reduced_model <- glm(type_num ~ title_words + title_caps_percent, 
                data = news, family = "binomial")

anova(reduced_model, news_mod2, test = "LRT")
```

:::

## Likelihood Ratio Test

:::: {.columns}
::: {.column width="25%"}
**Hypotheses:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=1}
$H_0: \beta_i = \beta_j = ... = 0 \quad$ vs. $\quad H_a: \text{at least 1 } \beta \ne 0$
:::
:::

::::

:::: {.columns}
::: {.column width="25%"}
**Test statistic:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=2}

$$\Lambda = 2 \log \frac{\text{max likelihood in full model}}{\text{max likelihood in reduced model}}$$
:::
:::

::::

<br>


:::: {.columns}
::: {.column width="25%"}
**Reference distribution:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=3}
$\chi^2_{df}$ where $df=$ number of extra terms in the full model
:::
:::

::::


**Idea:** how much did the full model increase the likelihood over the small model? (High likelihood = good, low likelihood = less good)

## `anova` table

Deviance = $-2 \log L$

::: small

| Resid DF | Resid Dev | Df | Deviance | Pr(>Chi) |
|:---:|:---:|:---:|---|---|
| n - coefs(reduced)| -2 log L(reduced) |  |  |  |
| n - coefs(full) | -2 log L(full) | # extra coefs | $\Lambda$ | p-value |

:::

## 

::: xsmall

```{r}
#| echo: true


summary(news_mod)
```

:::

## 

::: task
Fill in the ANOVA table below for the test $H_0: \beta_1 = \beta_2 = \beta_3 = 0$ against $H_A$ at least 1 $\ne 0$
:::

|  | Resid DF | Resid Dev | Df | Deviance | Pr(>Chi) |
|---|:---:|:---:|:---:|---|---|
| 1 |  |  |  |  |  |
| 2 |  |  |  |  |  |

## Z-test vs LRT 

Test for one coefficient $H_0: \beta_i = 0$ against $H_A: \beta_i \ne 0$: 

- In OLS, the t-test and F-test give *exactly the same p-values* if you are comparing 1 coefficient
- In logistic regression, the z-test and LRT are **not** equivalent
    - Hopefully they are close
    - If not, the LRT is usually more accurate

## AIC and BIC for logistic regression

The following "information criteria" can be used to compare nested or non-nested models (smaller values are better)

$$IC = \text{Measure of Fit} + \text{Penalty}$$

<br>

$$AIC = n \log(L) + 2p$$

<br>

$$BIC =  n \log(L) + \log(n) \cdot p$$

where L is the likelihood of the model 

## In R

```{r}
#| echo: true


AIC(news_mod, news_mod2, reduced_model)

BIC(news_mod, news_mod2, reduced_model)
```

