---
title: Logistic Regression
subtitle: Day 19
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

```

## Running Example

This dataset has information on 150 news articles, some that are real and some that are fake. Our goal is to build a statistical model to explain `type` based on features of the article

::: small

```{r}
news <- bayesrules::fake_news |>
  select(title, url, type, title_words, title_has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

write_csv(news, here("data/fake_news_bayesrules.csv"))

knitr::kable(head(news))
```

:::

## Exploratory Data Analysis

```{r}
ggplot(news, aes(y = type, fill = title_has_excl)) + 
  geom_bar() + 
  theme(legend.position = "bottom") + 
  scale_fill_viridis_d(option = "plasma", end = .75)
```

## Empirical Probability and Odds

```{r}
table(news$title_has_excl, news$type) |>
  addmargins()
```

- P(fake | has exclamation) = 
- Odds(fake | has exclamation) = 

- P(fake | no exclamation) = 
- Odds(fake | no exclamation) = 

- Odds Ratio: 

## EDA II 

```{r}
#| fig-width: 12
#| fig-height: 4


p1 <- gf_density_ridges(type ~ title_caps_percent, data = news, scale = 2, fill = ~type, show.legend = FALSE) + 
  scale_fill_viridis_d(option = "plasma", end = .75)

p2 <- gf_jitter(type_num ~ title_caps_percent, width = 0.2, height = .05, data = news) + labs(y = "Fake = 1")

p1 + p2
```

## SLR?

```{r}
#| fig-width: 5
#| fig-height: 4


p2 |>
  gf_lm(color = "#f15a31")
```

```{r}
lm(type_num ~ title_caps_percent, data = news) |> coef()
```

## Logistic Regression

```{r}
#| fig-width: 5
#| fig-height: 4


p2 |>
  gf_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, color = "#f15a31")
```

## Logistic Regression: Theoretical Model


$$\log(\frac{\pi(x)}{1-\pi(x)}) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_p x_p$$

- $\pi(x) = P(Y = 1 | x)$
- $\log(\frac{\pi}{1-\pi})$ is called the **logit** function or the **log-odds**
- Not constrained to be between 0 and 1 (like probabilities)
- Not constrained to be between 0 and $\infty$ (like odds) 

## Logistic Regression: R 

```{r}
#| echo: true 


news_glm <- glm(type_num ~ title_caps_percent, 
                data = news, 
                family = "binomial")
```

```{r}
news_glm
```

## Logistic Regression: Fitted model and interpreting $\beta_1$

$$\hat\eta = \log(\frac{\hat\pi(x)}{1-\hat\pi(x)}) = -0.93 + 0.1 x$$

. . . 

$\widehat \beta_1 = 0.1$ is the expected change in the log odds associated with a 1-unit increase title cap percent...

. . . 

<br>

but we don't "think" in terms of log odds 


## Log-odds &rarr; odds

$$\widehat\eta_i = \log(\frac{\widehat\pi(x)}{1-\widehat\pi(x)}) = -0.93 + 0.1 x$$

. . . 


$$\widehat{\text{odds}} = \frac{\widehat\pi(x)}{1-\widehat\pi(x)} = e^{-0.93 + 0.1 x} = e^{-.93}e^{.1x}$$

. . . 

if `title_caps_percent` increases by 1 unit: 

$$\frac{\frac{\widehat\pi(x+1)}{1-\widehat\pi(x+1)}}{\frac{\widehat\pi(x)}{1-\widehat\pi(x)}} = \frac{e^{-.93}e^{.1(x+1)}}{e^{-.93}e^{.1x}} = e^{.1} \approx 1.105$$
&rarr; the odds of fake news increases by a factor of 1.105

## Estimating probabilities

What's the probability a news article with 20% capitalization in the title is fake news? 

$$\widehat \eta(20) = \log \frac{\widehat\pi(20)}{1-\widehat\pi(20)} = -0.93 + 0.1(20) = 1.07$$

::: task
How do we get from the estimated log odds to a probability?
:::

## Log odds &rarr; probability

**Log odds:** 

$\widehat{\eta} = \log \left( \dfrac{\widehat{\pi}(\mathtt{X})}{1-\widehat{\pi}(\mathtt{X})} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 X_1 + \cdots + \widehat{\beta}_p X_p$

<br>


**Odds of success:** 

$\widehat{\omega} = \dfrac{\widehat{\pi}(\mathtt{X})}{1-\widehat{\pi}(\mathtt{X})} = e^{\widehat{\beta}_0 + \widehat{\beta}_1 X_1 + \cdots + \widehat{\beta}_p X_p}$

<br>

. . .

**Probability of success:** 


$\widehat{\pi}(X) = \dfrac{e^{\widehat{\beta}_0 + \widehat{\beta}_1 X_1 + \cdots + \widehat{\beta}_p X_p}}{1+e^{\widehat{\beta}_0 + \widehat{\beta}_1 X_1 + \cdots + \widehat{\beta}_p X_p}}$

## Estimating probabilities

What's the probability a news article with 20% capitalization in the title is fake news? 

$$\widehat \eta(20) = \log \frac{\widehat\pi(20)}{1-\widehat\pi(20)} = -0.93 + 0.1(20) = 1.07$$

::: task
How do we get from the estimated log odds to a probability?
:::

$$\widehat\pi(20) = \frac{e^{1.07}}{1+e^{1.07}} \approx 0.745$$

## Fitted regression model 

${\rm logistic}(\widehat{\eta}) = \dfrac{ \exp(\widehat{\eta}) }{ 1+  \exp(\widehat{\eta})}$ is the fitted curve on the probability scale

## Fitted regression model: 3 scales


:::: {.columns}
::: {.column width="33%" .small}
$\log(\frac{\widehat p}{1-\widehat p}) = \widehat\beta_0 + \widehat\beta_1 x$
:::

::: {.column width="33%" .small}
$\text{odds}(x) = e^{\widehat\beta_0 + \widehat\beta_1 x}$
:::

::: {.column width="33%" .small}
$P(Y = 1 | x) = \frac{e^{\widehat\beta_0 + \widehat\beta_1 x}}{1 + e^{\widehat\beta_0 + \widehat\beta_1 x}}$
:::
::::

```{r}
#| fig-width: 12
#| fig-height: 4
tibble(
  x = seq(0, 60, by = .5),
  log_odds = -0.9318 + 0.09772 * x,
  odds = exp(log_odds),
  prob = exp(log_odds)/(1 + exp(log_odds))
) |>
  pivot_longer(-x) |>
  ggplot(aes(x = x, y = value)) + 
  geom_line(color = "#003300", linewidth = 1.5) + 
  facet_wrap(vars(name), scales = "free_y") + 
  labs(y = "")
```
