---
title: Case Study Tips
subtitle: Day 24
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```


## Common issues on case studies

- Correct things in an incorrect order
- What goes in an EDA? 
- What types of plots go in the main text? 
- Report appendix vs quarto appendix

## General order

- Intro: What is the research question? What background information should the reader have to understand your study? What is the response and (general) predictor variables?

- EDA/Data: What dataset are you using? How many observations/variables? What is the distribution of the response variable? How are the predictors related to the response variable? Is there collinearity or skewness that should be addressed before fitting any models? 

- Results: 
    (a) Model building: what were your candidate models? what were your criteria for choosing a model? (You can report *theoretical* model here)
    (b) Conditions and diagnostics: does everything look OK? Did you need to do any further investigation? 
    (c) Now that you have verified the validity of your model, you can report the *fitted* model equation or the coefficient table
    (d) Interpretations of coefficients (with uncertainty!) or results from tests. Answer your research question here
    
- Discussion: 
  - Reiterate your main result without any technical results
  - Are there any limitations to your analysis? 
  - If you had another week, what could you investigate? If you had another dataset, what could you investigate? 
  
- Report appendix

- Quarto appendix

## Presenting equation

```{r}
moths <- read.csv("https://aluby.github.io/stat230/data/moths.csv") |>
  mutate(notremoved = Placed - Removed, 
               log_odds_removed = log(Removed / notremoved),
               prop_removed = Removed / Placed, 
               dark = ifelse(Morph=="dark",1,0), 
               distance = Distance, 
               morph = Morph) 


moth_glm <- glm(Removed/Placed ~ morph*distance, weights = Placed,
                data = moths, 
                family = "binomial")
```

::: small

My recommendation is to report the *theoretical* equation first, then include a coefficient table. 

`r equatiomatic::extract_eq(moth_glm, intercept = "beta")`

```{r}
broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  select(-c("statistic", "p.value")) |>
  knitr::kable(digits = 2)
```

:::

## Alternatively, put coefficient table in the report appendix and include a coefficent graph

```{r}
broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  gf_pointrange(estimate + conf.low + conf.high ~ term) |>
  gf_hline(yintercept = 0, linetype = "dashed", col = "gray") +
  coord_flip()
```

## Removing intercept

```{r}
broom::tidy(moth_glm, conf.int = TRUE, conf.level = .9) |>
  filter(term != "(Intercept)") |>
  gf_pointrange(estimate + conf.low + conf.high ~ term) |>
  gf_hline(yintercept = 0, linetype = "dashed", col = "gray") +
  coord_flip()
```

## Examples: 

- [Multiple linear regression example](https://www.causeweb.org/usproc/sites/default/files/usclap/2024-2/usclap%203418%20-%20community%20factors%20on%20violent%20crime%20rates.pdf))

- [Logistic regression example](https://causeweb.org/usproc/sites/default/files/usclap/2025-2/usclap%203657%20-%20predicting%20health%20insurance%20coverage.pdf)

- [Advanced method example](https://causeweb.org/usproc/sites/default/files/usclap/2025-2/usclap%203675%20-%20modeling%20sockeye%20salmon%20populations.pdf)

- [Experiment example](https://causeweb.org/usproc/sites/default/files/usclap/2025-1/usclap%203471.pdf)

- [A logistic regression example](https://causeweb.org/usproc/sites/default/files/usclap/2025-1/usclap%203502.pdf)
