---
title: Adding Categorical Predictors
subtitle: Day 10
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

bikes <- bikes |> select(date, season, year, month, weekend, holiday, temp_actual, temp_feel, humidity, windspeed, rides)
```


## Recap: `bikes` MLR

```{r}
bikes_slr <- lm(rides ~ temp_actual, data = bikes)

bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)

equatiomatic::extract_eq(bikes_mlr, use_coefs = TRUE)
```

::::: columns
::: {.column width="50%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003300", size = 1) |>
  gf_lm(rides ~ temp_actual, size = 1.5, color = "#f15a31") 
```

:::


::: {.column width="50%"}

<br> $R^2 =$ `r round(summary(bikes_mlr)$r.squared, 3)`

:::
:::::

## $R^2$ for MLR

The $R^2$ for the original SLR model using only `temp_actual` was `r round(summary(bikes_slr)$r.squared, 3)`

```{r}
summary(bikes_mlr)
```

## $R^2$ Recap

## Adjusted $R^2$

<br> Adjusts for the **number of predictors** in a model: 

$$R^2_{\text{adj}} =$$
<br> 

- $R^2$ (original) would **never decrease** if we add predictors
- $R^2_{adj}$ could **decrease** if we add a "bad" predictor

. . . 

<br> &rarr; better to use **adjusted** $R^2$ for MLR

## `bikes` data

```{r}
bikes |> 
  knitr::kable()
```

## Plan for today: remove `windspeed` but add `weekend`

```{r}
p1 <- gf_point(rides ~ temp_actual, color = ~weekend, data = bikes) 

p2 <- gf_boxplot(rides ~ weekend, fill = ~weekend, data = bikes)

p1 + p2 + plot_layout(widths = c(3, 1)) & theme(legend.position = "none")
```

## Indicator variables

Regression requires a numeric representation of all variables, so we treat `TRUE = 1` and `FALSE = 0`

::: task
For each regression model on the handout, sketch the fitted model on the whiteboard

- Each fitted model will have two lines: one for weekends and one for weekdays
- Work with your groups (last day together!) 
:::


## Model 1 

## Model 2

## Model 3

## Parallel lines model

The number of rides increases at the same rate, but always lower on weekends

```{r}
p1 |>
  gf_lm()
```

## Different slopes model

The number of rides starts at the same baseline, but the relationship between temperature and rides is different on weekends vs weekdays

```{r}
mod_same_int <- lm(rides ~ temp_actual:weekend, data = bikes)
bikes_aug <- augment(mod_same_int, data = bikes)
p1 |>
 gf_line(.fitted ~ temp_actual, color = ~weekend, data = bikes_aug, linewidth = 1)
```

## Separate lines model

Weekends/weekdays have different baselines and different relationships with the number of rides

```{r}
mod_sep_lines <- lm(rides ~ temp_actual + weekend + temp_actual*weekend, data = bikes)
bikes_aug <- augment(mod_sep_lines, data = bikes)
p1 |>
 gf_line(.fitted ~ temp_actual, color = ~weekend, data = bikes_aug, linewidth = 1)
```


## Parallel lines model

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{temp} + \beta_2 \texttt{weekend}$$

<br> 

- $\beta_0$: intercept for weekdays
- $\beta_0 + \beta_2$: intercept for weekends
- $\beta_1$: slope for both groups

## Different slopes model

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{temp} + \beta_2 \texttt{temp} \times \texttt{weekend}$$

<br> 

- $\beta_0$: intercept for both groups
- $\beta_1$: slope for weekdays
- $\beta_1 + \beta_2$: slope for weekends

## Separate lines model

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{temp} + \beta_2 \texttt{weekend} + \beta_3 \texttt{temp} \times \texttt{weekend}$$

<br> 

- $\beta_0$: intercept for weekdays
- $\beta_0 + \beta_2$: intercept for weekends
- $\beta_1$: slope for weekdays
- $\beta_1 + \beta_3$: slope for weekends

## More than 2 categories

What if we want to incorporate `season`? (fall, spring, summer, winter)

## A possibly bad idea

Convert the column to numeric: 

- Fall $\to$ 1
- Winter $\to$ 2
- Spring $\to$ 3
- Summer $\to$ 4

$$\mu(\text{rides} | \text{season}) = \beta_0 + \beta_1 \text{season}$$

## A good idea

Create a set of indicator (dummy) variables to represent the four categories

:::: {.columns}

::: {.column width="25%"}
|Original|
|--------|
| winter |
| fall |
| spring  |
| spring |
| fall |
| summer |
:::

::: {.column width="20%" .fragment}
| winter |
|--------|
| 1|
| 0|
|0 |
| 0|
| 0|
| 0|
:::

::: {.column width="20%" .fragment}
| spring |
|--------|
| 0|
| 0|
| 1|
| 1|
| 0|
| 0|
:::

::: {.column width="20%" .fragment}
|summer|
|--------|
| 0|
| 0|
| 0|
| 0|
| 0|
| 1|
:::

::::

## Indicator variables

For a categorical variable with $k$ levels, create $k-1$ indicator variables to fully represent the original variable

## Example 2

::: task
For each sketched regression model, write the mean function for a regression model that matches on the whiteboard

- Note that line type = season
- Work with your groups

:::
