---
title: Intro to Poisson Regression
subtitle: Day 26
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```

## Motivating data: number of anti-discrimination laws

The `equality_index` data includes the observed number of anti-discrimination laws in each state,  $Y$ . We are interested in modeling  $Y$ by a state’s `percent_urban` and its `historical` presidential voting patterns. The `historical` variable has 3 levels:

- `dem` = the state typically votes for the Democratic candidate
- `gop` = the state typically votes for the Republican (GOP) candidate
- `swing` = the state often swings back and forth between Democratic and Republican candidates

::: footer
Source: the {bayesrules} R package
:::

## EDA


```{r}
#| echo: false 
#| fig-width: 8
#| fig-height: 3


data("equality_index")

p1 <- ggplot(equality_index, aes(x = laws)) + 
  geom_histogram(col = "white")

p2 <- ggplot(equality_index, aes(y = laws, x = percent_urban)) + 
  geom_point(size = .5)

p3 <- ggplot(equality_index, aes(y = laws, x = historical)) + 
  geom_boxplot(aes(fill = historical), alpha = .6) +
  geom_point(size = .5) +
  theme(legend.position = "none")

p2 + p1 + p3
```
- `laws` is skewed
- as percent urban increases, laws also increases
- `dem` and `swing` generally have more laws than `gop`

## EDA (log scale)


```{r}
#| echo: false 
#| fig-width: 8
#| fig-height: 3

data("equality_index")

p1 <- ggplot(equality_index, aes(x = log(laws))) + 
  geom_histogram(col = "white")

p2 <- ggplot(equality_index, aes(y = log(laws), x = percent_urban)) + 
  geom_point(size = .5)

p3 <- ggplot(equality_index, aes(y = log(laws), x = historical)) + 
  geom_boxplot(aes(fill = historical), alpha = .6) +
  geom_point(size = .5) +
  theme(legend.position = "none")

p2 + p1 + p3
```

- `log(laws)` is still a little skewed
- `log(laws)` and `percent_urban` is more linear
- if we had states with 0 laws, might have to do `log(laws + 1)`

## SLR Model (ignoring `historical` for now)

```{r}
ggplot(equality_index, aes(y = log(laws), x = percent_urban)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)

equality_slr <- lm(log(laws) ~ percent_urban, data = equality_index)
tidy(equality_slr)
```

## 

```{r}
ggResidpanel::resid_panel(equality_slr)
```

. . . 

&rarr; some remaining nonlinearity, but is it worth going the transformation route? Why might we prefer a different model entirely?

## Assumed SLR model (simulated data)


```{r fig.height = 3.5, fig.width = 6, fig.align='center'}
# Now make Poisson regression picture
set.seed(0)
old_dat <- data.frame(x=(x=runif(50, 35, 95)),
                  y=rnorm(50, exp(-1.21 + .04179*x), 
                          sd = 4*sigma(equality_slr)))


old_slab_dat <- data.frame(
  x = c(45, 65,85),
  yhat = exp(predict(equality_slr, newdata = data.frame(percent_urban = c(45, 65, 85))))
)

library(ggdist)
ggplot(old_dat, aes(x, y)) +
  geom_point(size = 0.1, alpha = 0.4) +
  geom_function(fun = \(x) exp(-1.21 + .04179*x)) +
    stat_slab(aes(x = x, ydist = "norm", arg1 = yhat, arg2 = 4*sigma(equality_slr)), data = old_slab_dat, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "darkgreen", side="left") +
  geom_vline(xintercept=c(45, 65, 85), lty=2) +
  labs(x = "percent_urban", y = "laws")
```

## Model for count data

```{r fig.height = 3.5, fig.width = 6, fig.align='center'}
# Now make Poisson regression picture

equality_glm <- glm(laws ~ percent_urban, data = equality_index, family = "poisson")

library(tidyverse)

# 1. Setup your model and generate data
set.seed(0)
dat <- data.frame(
  x = runif(100, 35, 95), # Increased sample size slightly for better visualization
  y = rpois(100, exp(-2.18 + 0.061 * x))
)

# 2. Define the exact vertical slices where you want distributions shown
# (We pick clean locations within your data range)
anchor_points <- c(45, 65, 85)
scale_factor <- 45  # Controls how far left the distributions bulge

# 3. Generate the theoretical Poisson staircase steps at each anchor point
dens_steps <- lapply(anchor_points, function(x_anchor) {
  # Calculate the true theoretical lambda at this specific X position
  lambda <- exp(-2.18 + 0.061 * x_anchor)
  
  # Define the range of counts (Y) to calculate probabilities for
  y_vals <- 0:max(dat$y + 5)
  p_vals <- dpois(y_vals, lambda)
  
  # Create clean staircase steps manually by duplicating coordinates:
  # This makes a constant horizontal probability profile for each integer count
  data.frame(
    x = rep(x_anchor - scale_factor * p_vals, each = 2),
    y = rep(y_vals, each = 2) + c(-0.5, 0.5),
    group = as.character(x_anchor)
  )
}) |> do.call(rbind, args = _)

# 4. Generate the final plot
ggplot(dat, aes(x, y)) +
  geom_point(size = 1.5, alpha = 0.5, color = "black") +
  # Plot the continuous Poisson mean function line
  geom_function(fun = \(x) exp(-2.18 + 0.061 * x), linewidth = 1, color = "black") +
  # Add the vertical dashed anchor lines
  geom_vline(xintercept = anchor_points, linetype = "dashed", alpha = 0.5) +
  # Overlay the beautiful staircase PMF paths
  geom_path(data = dens_steps, aes(x, y, group = group), color = "orange", linewidth = 1.1) +
  labs(
    x = "percent_urban", 
    y = "Laws",
  )
```

## Generalized linear model

::: {.fragment fragment-index=1}
**Random component: **
:::

::: {.fragment fragment-index=2}
What distribution does the response variable follow?
:::

<br>


::: {.fragment fragment-index=1}
**Linear predictor (systematic component): **
:::

::: {.fragment fragment-index=3}
What function of the predictors should we use?
:::

<br>


::: {.fragment fragment-index=1}
**Link function: **
:::

::: {.fragment fragment-index=4}
How can we relate the expectation of $Y$ and our linear predictor?
:::


## Random component

What distribution does the response variable follow?

<br>

**Possible values:** We are counting the number of events that occur over a fixed interval, so $Y =0, 1, 2, \ldots$

<br>

**Poisson distribution**

$P(Y = y) = \frac{e^{-\lambda} \lambda^y}{y!}, \quad \text{for} \quad y = 0, 2, 1, \ldots$



## Linear predictor

What function of the predictors should we use?

<br>

$\eta = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$

:::{.aside}
$\eta =$ "eta"
:::

##

{{< video https://www.youtube.com/embed/6pDH66X3ClA?si=IsolyWEfJizfItxS width="100%" height="100%" >}}


## Link function

How can we relate the expectation of $Y$ and our linear predictor?

. . .

<br>

$E(Y) = \lambda$ between 0 and $\infty$

<br>

$\eta = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p \longrightarrow$ between $-\infty$ and $\infty$

<br>

What function takes numbers between between 0 and $\infty$ as input and maps them to the real line?


## Link function

How can we relate the expectation of $Y$ and our linear predictor?


$\log(\lambda) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p = \eta$


## Poisson regression model for counts

**Random component:** $Y_i = \#$ anti-discrimination laws in a state; Poisson distribution is reasonable

. . .

<br>

**Link function:** $E(Y) = \lambda$ and $\lambda >0$

$\Longrightarrow \log(\lambda) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p = \eta$

. . .

<br>

**Inverse-link function:** $E(Y) = e^{\beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p}$

## EDA

:::: {.columns}

::: {.column width="60%"}
```{r fig.height = 4, fig.width=4, out.width="100%"}
gf_point(log(laws) ~ percent_urban, data = equality_index, height = 0.25, width = 0.25, alpha = 0.5) 
```

:::

::: {.column width="40%"}

<!-- - Why did we use <br> `log(nests + 0.5)`? -->

What structure are we looking for in this plot?

<br>

<br>


There are no 0 counts in this data set, how could we create this plot if we had observed 0s?

:::

::::

## Model fitting


```{r eval=FALSE, echo=TRUE}
glm(laws ~ percent_urban, data = equality_index, 
    family = poisson)
```

```{r}
pois1 <- glm(laws ~ percent_urban, data = equality_index, family = poisson)
summary(pois1)$coef
```

<br>

**Fitted model:** $\log(\widehat{\lambda}) = -2.18 + 0.06 \cdot \tt{percent urban}$

. . .

**How can we interpret the coefficients?**


## Interpreting the coefficients

$\log(\widehat{\lambda}) = \widehat{\beta}_0 + \widehat{\beta}_1 x$


How does the expected (average) number of events, $\widehat{\lambda}$, change for a 1-unit increase in $x$?

. . . 

Let's consider $x=0$ and $x=1$:

* $\widehat{\lambda} = e^{\widehat{\beta}_0 + \widehat{\beta}_1(0)}=e^{\widehat{\beta}_0}$

* $\widehat{\lambda} = e^{\widehat{\beta}_0 + \widehat{\beta}_1(1)}=e^{\widehat{\beta}_0} e^{\widehat{\beta}_1}$

. . .

The expectation at $x+1$ is $e^{\widehat{\beta}_1}$ times the expectation at $x$

## Interpreting the coefficients

In our example:  $\log(\widehat{\lambda}) = -2.18 + 0.06 \cdot \tt{percent urban}$

. . .

$e^{0.06} \approx 1.06$

. . . 

<br>


&rarr; A 1 percent increase in the percent of urban residents is associated with an increase in the average number of antidiscrimination laws by a factor of 1.06
(i.e., a 6% increase).


## Multiple predictors {.smaller}


```{r echo=TRUE}
full <- glm(laws ~ percent_urban + historical,
            data = equality_index,
            family = poisson)
summary(full)
```


## 

<br>

&rarr; A 1 percent increase in the percent of urban residents is associated with an increase in the average number of antidiscrimination laws by a factor of $e^{.03} = 1.03$, holding `historical` constant
(i.e., a 3% increase).

. . . 

<br>

&rarr; Compared to `dem` states, `gop` states have fewer antidiscrimination laws by a factor of $e^{-1.6} = .202$, assuming `percent_urban` is held constant
(i.e., a 80% decrease).

. . . 

<br>

&rarr; Compared to `dem` states, `swing` states have fewer antidiscrimination laws by a factor of $e^{-.83} = .44$, assuming `percent_urban` is held constant
(i.e., a 56% decrease).


## Today: 

- EDA
- Interpretations
- Wald-based tests and CI's

### Next time: 

- Conditions for inference
- LRT 
- quasipoisson



