---
title: Inference for MLR
subtitle: Day 11
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

bikes <- bikes |> select(date, season, year, month, weekend, holiday, temp_actual, temp_feel, humidity, windspeed, rides)

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt", sep = "\t", header = TRUE)

bikes_slr <- lm(rides ~ temp_actual, data = bikes)
bikes_mlr <- lm(rides ~ temp_actual * weekend, data = bikes)
bikes_par <- lm(rides ~ temp_actual + weekend, data = bikes)

set.seed(230)
bikes_noise <- lm(rides ~ temp_actual + n1 + n2 + n3 + n4 + n5,
                  data = bikes |> mutate(n1 = rnorm(n()), n2 = rnorm(n()), n3 = rnorm(n()),
                                         n4 = rnorm(n()), n5 = rnorm(n())))

bikes_season <- lm(rides ~ temp_actual + season, data = bikes)
bikes_season_w <- lm(rides ~ temp_actual + season_w,
                     data = bikes |> mutate(season_w = relevel(season, ref = "winter")))

a_full <- anova(bikes_slr, bikes_mlr)
a_slopes <- anova(bikes_par, bikes_mlr)
a_shift <- anova(bikes_slr, bikes_par)
d_full <- a_full$Df[2]
f_full <- a_full$F[2]
f_full_p <- a_full$`Pr(>F)`[2]

fmt_p <- function(p) ifelse(p < 0.001, "<0.001", format(round(p, 3), nsmall = 3))
sci <- function(p) format(p, digits = 2, scientific = TRUE)

coef_tbl <- function(m) {
  tidy(m) |>
    mutate(across(c(estimate, std.error, statistic), \(x) round(x, 2)),
           p.value = fmt_p(p.value)) |>
    knitr::kable()
}


```

## MLR model for `bikes` data

```{r}
bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)
broom::tidy(bikes_mlr) |> select(term, estimate)
```

::: {.task .medium}
**Warm up:** Write out the theoretical and fitted regression model
:::

::: notes
Timing: ~7 min for this block (warm-up, different slopes?, t-test/CI review). Don't re-derive the t-test; it is SLR review.
:::

## Do we _need_ a different slopes model? {.smaller}

```{r}
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes) |>
  gf_lm()

coef_tbl(bikes_mlr) 
```

## Hypothesis test for a single coefficient

Let $p$ = the number of predictors in the model (an interaction counts as one)

<br>

**Hypotheses:** $H_0: \ \beta_j = \#$ vs. $H_a: \ \beta_j \ne, <, \text{or} > \#$

**Test statistic:** $t = \dfrac{\widehat{\beta}_j - \#}{SE(\widehat{\beta}_j)}$

**Reference distribution:** $t$ distribution with d.f. = $n-(p+1)$

**p-value:** Area in the tail(s) specified by $H_a$

## CIs for a single coefficient

$\widehat{\beta}_j \pm t^*_{n-(p+1)} \cdot SE(\widehat{\beta}_j )$

<br>

```{r}
#| echo: true
tidy(bikes_mlr, conf.int = TRUE) |> select(term, estimate, std.error, conf.low, conf.high)
qt(0.975, df = 496)
```

By hand, for `temp_actual`: $`r round(coef(bikes_mlr)[2], 2)` \pm `r round(qt(0.975, 496), 3)` \times `r round(summary(bikes_mlr)$coef[2, 2], 3)` = (`r round(confint(bikes_mlr)[2, 1], 1)`, `r round(confint(bikes_mlr)[2, 2], 1)`)$

## Answer: Do we need a different slopes model?

- **Graph:** the weekday and weekend lines are nearly parallel
- **Table:** the interaction coefficient is only `r round(coef(bikes_mlr)[4], 2)` rides per degree, with $p =$ `r round(summary(bikes_mlr)$coef[4, 4], 3)`

So there is no evidence that the slope for temperature differs on weekends. 

. . .

**Next question:** do we need `weekend` *at all*?

## Do we need `weekend` at all?

Full model with the interaction:

```{r}
coef_tbl(bikes_mlr)
```


## ...but wait

Drop only the interaction (parallel lines model):

```{r}
coef_tbl(bikes_par)
```

Now `weekendTRUE` **is** significant!

. . .

::: callout-caution
## Caution!

$\widehat \beta$'s and p-values **are highly dependent**. A t-test tells us: "is this one term needed, *given everything else is already in the model*?"
:::

. . . 

We need a test that asks about several coefficients at once

::: notes
Timing: puzzle block ~7 min (minutes 7-14). Let students argue which table to believe before revealing the resolution.
:::

## Explained Variability

```{r echo=FALSE, fig.height = 2.5, fig.width = 7, fig.align='center'}
slr_aug <- augment(bikes_slr)
slr_aug <- mutate(slr_aug, mean.y = mean(rides))

p1 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Total", size = .2) %>%
  gf_segment(mean.y + rides ~ temp_actual + temp_actual, color = "gray70", alpha = .4) %>%
  gf_hline(yintercept = mean(slr_aug$rides), color = "#f15a31") 

p2 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Error", size = .2) %>%
  gf_segment(.fitted + rides ~ temp_actual + temp_actual, color = "#003069", alpha = .4) %>%
  gf_lm(color = "#f15a31") 

p1 + p2
```

. . . 

**Idea**: Compare the *explained variability* of the two models

## Explained Variability


```{r}
ssdf <- tribble(
  ~Source, ~SS,
  "Error",      6,
  "Total",       10,
  "Regression", 4
)

sst_plot <- ggplot(data = ssdf |> filter(Source == "Total"), aes(x = "", y = SS)) +
  geom_bar(stat="identity", width=1) +
  theme_void() +
  labs(title = "SSTotal")


decomp_plot <- ggplot(data = ssdf |> filter(Source != "Total"), aes(x = "", y = SS, fill = Source)) +
  geom_bar(stat="identity", width=1) +
  geom_text(aes(label = paste0(Source)), position = position_stack(vjust = 0.5), color = "white") +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#f15a31", "#003069"))

ssdf2 <- tribble(
  ~Source, ~SS,
  "Error",      5,
  "Total",       10,
  "Regression", 5
)

sst_plot2 <- ggplot(data = ssdf2 |> filter(Source == "Total"), aes(x = "", y = SS)) +
  geom_bar(stat="identity", width=1) +
  theme_void() +
  labs(title = "SSTotal")


decomp_plot2 <- ggplot(data = ssdf2 |> filter(Source != "Total"), aes(x = "", y = SS, fill = Source)) +
  geom_bar(stat="identity", width=1) +
  geom_text(aes(label = paste0(Source)), position = position_stack(vjust = 0.5), color = "white") +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#f15a31", "#003069"))
```

::: columns
:::{.column width="30%"}
"Reduced" Model


```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
sst_plot + decomp_plot
```
:::
:::{.column width="30%"}
"Full" Model


```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
sst_plot2 + decomp_plot2
```
:::

:::{.column width="30%" .fragment .center}
<br>

**Idea:** Does the full model reduce the error (SSE) by *more than chance alone*?

:::
:::


## Comparing nested models

:::: {.columns align=center}
::: {.column width="20%"}
Full model
:::

::: {.column width="75%"}
$\mu \lbrace Y | X \rbrace = \beta_0 + \beta_1 \text{temp} + \beta_2 \text{weekend} + \beta_3 \text{temp} \times \text{weekend}$
:::

::: {.column width="20%"}
Reduced model
:::

::: {.column width="75%"}
$\mu \lbrace Y | X \rbrace = \beta_0 + \beta_1 \text{temp}$
:::

::::

<br>

The reduced model is the full model with some $\beta$'s set to 0, so we say the models are **nested**

. . .

<br>

:::: {.columns align=center}
::: {.column width="20%"}
Hypotheses
:::

::: {.column width="75%"}
$H_0:\ \beta_{2} = \beta_3 = 0$ (`weekend` doesn't matter: it neither shifts the line nor changes its slope)

$H_a:$ at least one of $\beta_2, \beta_3 \ne 0$
:::

::::

## The extra sum of squares F test

$$F = \frac{(\text{SSE}_{\text{reduced}} - \text{SSE}_{\text{full}}) / d}{\text{MSE}_{\text{full}}} = \frac{\text{how much error drops per } \beta}{\text{estimate of } \sigma^2}$$

<br>


::: fragment

- If $H_0$ is true, both parts estimate $\sigma^2$, so $F \approx 1$
- If the extra terms really help, $F$ is much bigger than 1

:::

::: small
where 

- $d = \text{df}_{reduced} - \text{df}_{full}$ = number of $\beta$'s being tested, 
- $\text{df}_{i} = n - (p_i + 1)$ is the error d.f. for model $i$, 
- $\text{MSE}_{\text{full}} = \dfrac{\text{SSE}_{\text{full}}}{\text{df}_{\text{full}}}$
:::

## Upper-tail p-values

The F-statistic follows an $F$ distribution with $d$ and $\text{df}_{full}$ d.f. It is right-skewed, and only large values of $F$ are evidence against $H_0$, so the p-value is the **upper tail**

```{r fig.height=3, fig.width=5, echo=FALSE, message=FALSE, fig.align='center'}
 ggplot(data = data.frame(x = c(.00001, 5.5)), aes(x)) +
  stat_function(fun = df, args = list(df1=4, df2=23), color = "navy", linewidth = 1) +
  stat_function(fun = df, args = list(df1=4, df2=23), xlim = c(3, 5.5),
                  geom = "area", fill = "navy", alpha = .4) +
  labs(x = "F", y = "Density") +
  scale_y_continuous(limits = c(0, 0.8), expand = c(0,0)) +
   geom_curve(aes(x = 5, y = .22, xend = 4.25, yend = .03), arrow = arrow(length = unit(0.03, "npc")), curvature = .2, color = "navy") +
  annotate("text", x = 5.26, y = .25, label = "p-value", color = "navy", size = 6) 
```

To obtain upper-tail areas: `pf(stat, df1, df2, lower.tail = FALSE)`

## Putting it all together: by hand

```{r}
tibble(
  Model = c("Reduced: `temp_actual`", "Full: `temp_actual * weekend`"),
  SSE = c(deviance(bikes_slr), deviance(bikes_mlr)),
  df = c(df.residual(bikes_slr), df.residual(bikes_mlr))
) |>
  knitr::kable(format.args = list(big.mark = ",", scientific = FALSE), digits = 0)
```

. . .

$$
F = \frac{(`r format(round(deviance(bikes_slr)), big.mark = ",")` - `r format(round(deviance(bikes_mlr)), big.mark = ",")`) / `r d_full`}{`r format(round(deviance(bikes_mlr)), big.mark = ",")` / `r df.residual(bikes_mlr)`} \approx `r round(f_full, 2)`
$$

. . .

```{r}
#| echo: true
pf(15.26, df1 = 2, df2 = 496, lower.tail = FALSE)
```

## Extra sums of squares F-test in R

```{r}
#| echo: true
full <- lm(formula = rides ~ temp_actual * weekend, data = bikes)    #<1>
reduced <- lm(formula = rides ~ temp_actual, data = bikes) #<2>
anova(reduced, full) #<3>
```

1. Fit the full model
2. Fit the reduced model
3. Use `anova()` to compare the two models. The first argument should be the reduced model.

## Reading the `anova()` output

```
Analysis of Variance Table

Model 1: rides ~ temp_actual
Model 2: rides ~ temp_actual * weekend
  Res.Df       RSS Df    Sum of Sq      F               Pr(>F)    
1    df1      SSE1                                  
2    df2      SSE2  d    SSE1-SSE2  (SumSq/d)/(SSE2/df2) p-val
```

. . .

**Conclusion:** $p =$ `r sci(f_full_p)`. Strong evidence that `weekend` matters (as a shift in the line, a change in slope, or both)

## Resolving the puzzle {.smaller}

```{r}
tibble(
  Question = c("Different slopes needed?", "Weekend shift needed (no interaction)?", "Weekend needed at all?"),
  Reduced = c("`temp + weekend`", "`temp`", "`temp`"),
  Full = c("`temp * weekend`", "`temp + weekend`", "`temp * weekend`"),
  d = c(a_slopes$Df[2], a_shift$Df[2], a_full$Df[2]),
  F = as.character(signif(c(a_slopes$F[2], a_shift$F[2], a_full$F[2]), 3)),
  `p-value` = fmt_p(c(a_slopes$`Pr(>F)`[2], a_shift$`Pr(>F)`[2], a_full$`Pr(>F)`[2]))
) |>
  knitr::kable()
```

<br>

- One tool answers all three questions
- When $d = 1$, the F-test and the t-test are the same test ($F = t^2$)
- The t-tests couldn't answer the *joint* question (row 3)

::: notes
Timing: F-test block ~16 min (minutes 22-38). Spend the time on the "Putting it all together: by hand" and "Reading the anova() output" slides; if short on time, skip "Resolving the puzzle" and revisit it at the start of Day 12.
:::

## `anova()` on a single model

```{r}
#| echo: true
anova(bikes_mlr)
```

- Each row adds one term to the model, in order, and tests whether it improves the fit given the terms above it
- The **last row** is the extra sums of squares F-test for the last term and matches `anova(bikes_par, bikes_mlr)`

## Reading R output

```
lm(formula = rides ~ temp_actual * weekend, data = bikes)

Coefficients:
                         Estimate Std. Error t value Pr(>|t|)    
(Intercept)             -1843.867    415.700  -4.436 1.13e-05 ***
temp_actual                87.173      6.439  13.538  < 2e-16 ***
weekendTRUE              -714.938    795.698  -0.899    0.369    
temp_actual:weekendTRUE     0.411     12.539   0.033    0.974    

Residual standard error: 1253 on 496 degrees of freedom
Multiple R-squared:  0.3709,	Adjusted R-squared:  0.3671 
F-statistic: 97.47 on 3 and 496 DF,  p-value: < 2.2e-16
```

::: footer
Output was edited to fit on one slide
:::

. . .

The **F-statistic** at the bottom is the same kind of test: the full model vs. an *intercept-only* model. $H_0: \beta_1 = \beta_2 = \beta_3 = 0$ (none of the predictors help)

## Last piece of the `summary()` table: adjusted $R^2$

::: columns
:::{.column width="50%"}

$$R^2 = 1 - \frac{SSE}{SST}$$

:::
:::{.column width="50%" .fragment}
$$R^2_{adj} = 1 - \frac{SSE/(n-p-1)}{SST/(n-1)}$$
$$R^2_{adj} = 1 - \frac{MSE}{MST}$$
:::

:::

Adjusted $R^2$ adds a penalty for having a more complex model (model with more $\beta$s)

## $R^2$ vs Adjusted $R^2$ {.incremental}

```{r}
tibble(
  Model = c("`temp_actual`", "`temp_actual + weekend`", "`temp_actual * weekend`", "`temp_actual` + 5 random noise variables"),
  `$R^2$` = c(summary(bikes_slr)$r.squared, summary(bikes_par)$r.squared,
              summary(bikes_mlr)$r.squared, summary(bikes_noise)$r.squared),
  `Adjusted $R^2$` = c(summary(bikes_slr)$adj.r.squared, summary(bikes_par)$adj.r.squared,
                       summary(bikes_mlr)$adj.r.squared, summary(bikes_noise)$adj.r.squared)
) |>
  knitr::kable(digits = 4)
```

. . .

- $R^2$ never goes down when predictors are added, even for pure noise! 
- Bigger $R^2$ alone can't tell us whether the improvement is "real"
- $R^2_{adj}$ could decrease if unnecessary predictors are added 
- Better to use $R^2_{adj}$ than $R^2$ to compare models
- Neither is a formal test



## Takeaways

- A t-test asks about **one** coefficient, given everything else in the model
- To test several coefficients at once, compare nested models with an **extra sums of squares F-test**
- $R^2$ always goes up; adjusted $R^2$ accounts for model complexity; the F-test tells us whether the improvement is "significant"

## Activity: Inference for MLR

Open the Day 11 activity. Questions 10 and 11 are the extra sums of squares F-tests.

```{r}
countdown::countdown(minutes = 20)
```

## Appendix: t distribution tail areas {visibility="hidden"}

```{r echo=FALSE, fig.height = 2, fig.width = 7.5, out.width=850}
p1 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] < 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p2 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] != 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p3 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] > 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

(p1 | p2 | p3)
```

## Appendix: F distributions {visibility="hidden"}

```{r fig.height=3, fig.width=5, echo=FALSE, message=FALSE, fig.align='center'}
par(mar = c(2, .1, .1, .1))
plot(c(0, 5), c(0, df(.01, 1, 1)), type = "n", axes = FALSE)
at <- seq(0, 5, 1)
axis(1, at)
X <- seq(0, 5, 0.01)
DF1 <- c(1, 2, 5, 10, 100)
DF2 <- c(1, 1, 2, 1, 100)
COL. <- RColorBrewer::brewer.pal(n = 7, name = "Blues")[-1]
for (i in 1:5) {
  Y <- df(X, as.numeric(DF1[i]), as.numeric(DF2[i]))
  lines(X, Y, col = COL.[i])
}
legend(2.5, 3.25,
       legend = c(paste('df1 = ', DF1[1:5], ', df2 = ', DF2[1:5], sep = '')),
       col = COL.,
       text.col = COL.,
       lty = rep(1, 5))
```
