---
title: Overdispersion and Inference for Poisson Regression
subtitle: Day 27
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```

## Motivating data: number of anti-discrimination laws

The `equality_index` data includes the observed number of anti-discrimination laws in each state,  $Y$ . We are interested in modeling  $Y$ by a state’s `percent_urban` and its `historical` presidential voting patterns. The `historical` variable has 3 levels:

- `dem` = the state typically votes for the Democratic candidate
- `gop` = the state typically votes for the Republican (GOP) candidate
- `swing` = the state often swings back and forth between Democratic and Republican candidates

::: footer
Source: the {bayesrules} R package
:::

## EDA


```{r}
#| echo: false 
#| fig-width: 8
#| fig-height: 3


data("equality_index")

p1 <- ggplot(equality_index, aes(x = laws)) + 
  geom_histogram(col = "white")

p2 <- ggplot(equality_index, aes(y = laws, x = percent_urban)) + 
  geom_point(size = .5)

p3 <- ggplot(equality_index, aes(y = laws, x = historical)) + 
  geom_boxplot(aes(fill = historical), alpha = .6) +
  geom_point(size = .5) +
  theme(legend.position = "none")

p2 + p1 + p3
```
- `laws` is skewed
- as percent urban increases, laws also increases
- `dem` and `swing` generally have more laws than `gop`

## Review: Poisson regression model

- If $Y_i =$ # of events in a fixed interval of time/space, then $Y_i \sim \text{Poisson}(\lambda)$ 


- This leads to the Poisson regression model for counts:

    $$\log(\lambda) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$
    or equivalently
    

## Assumptions

1. **Linearity**: the log of the mean of Y, $\log(\lambda)$, must be a linear function of the predictors

2. **Mean = Variance**: the mean of a Poisson random variable must be equal to its variance

3. **Independence**: the observations must be independent

## Model fitting {.smaller}

```{r echo=TRUE}
full_glm <- glm(laws ~ percent_urban + historical,
            data = equality_index,
            family = poisson)
summary(full_glm)
```

## Pearson residuals

$$Pres_i = \frac{Y_i - \widehat{\text{mean}(Y_i)}}{\widehat{SD(Y_i)}}$$


```{r}
ggResidpanel::resid_xpanel(full_glm, type = "pearson")
```


## For Poisson model

$$Pres_i = \frac{Y_i - \hat \mu_i}{\widehat{SD(Y_i)}}$$

- Mean = $\hat \lambda$, SD = $\sqrt{\hat \lambda}$
- Pearson residual: $\frac{Y_i - \hat \lambda_i}{\sqrt{\hat \lambda_i}}$
- If $\hat\lambda$'s are "big" ($\ge 5$), Pearson residuals are $\approx N(0,1)$ if Poisson model is correct. Can use to diagnose problems with linearity
- If overdispersed, then SD of Pearson residuals is $>1$

## Goodness of Fit test

We want a formal way to test whether our Poisson model is a good fit for the data, or if it leaves a significant amount of variation unexplained.

* **$H_0$:** The Poisson model fits the data adequately.
* **$H_A$:** The Poisson model is deficient (a more complex, saturated model is required).

The **Residual Deviance** ($G^2$) serves as our test statistic:
$$G^2 = 2 \sum_{i=1}^n \left[ y_i \log\left(\frac{y_i}{\hat{\lambda}_i}\right) - (y_i - \hat{\lambda}_i) \right]$$

* Under $H_0$, $G^2$ follows an approximate $\chi^2$ distribution with $df = n - p - 1$.
* A very small $p$-value means we reject $H_0$, indicating poor fit (often due to overdispersion).

## Are means big enough for GoF test? 

```{r}
summary(predict(full_glm, type = "response"))
```


## Quasipoisson model 

```{r}
full_quasi <- glm(laws ~ percent_urban + historical,
            data = equality_index,
            family = quasipoisson)
summary(full_quasi)
```

## Pearson residuals

```{r}
ggResidpanel::resid_xpanel(full_quasi, type = "pearson")
```

## Studentized residuals

Pearson residuals tell us how many standard deviations an observation falls from its predicted mean, but they don't account for **leverage** ($h_i$). 

$$r_i = \frac{Pres_i}{\sqrt{1 - h_i}} = \frac{Y_i - \hat{\lambda}_i}{\sqrt{\hat{\lambda}_i(1 - h_i)}}$$

* **Why adjust for leverage?** Points with extreme predictor values (high leverage) naturally force the model line close to them, artificially shrinking their raw and Pearson residuals.
* **Interpretation:** If the model assumptions hold, studentized residuals should be roughly normally distributed with a standard deviation of 1. 
* Look out for any extreme outliers where $|r_i| > 3$.

## Inference for coefficients

Under the Poisson model, we can perform inference on our slopes ($\beta_j$) using **Wald tests** and standard normal critical values ($z^*$).

### Testing $H_0: \beta_j = 0$
The test statistic is a $z$-score rather than a $t$-statistic:
$$z = \frac{\hat{\beta}_j}{SE(\hat{\beta}_j)} \sim N(0,1)$$

### Confidence Intervals
We construct intervals on the **log (link) scale** first, then exponentiate:
$$\text{Scale of Log-Means:} \quad \hat{\beta}_j \pm z^* \cdot SE(\hat{\beta}_j)$$
$$\text{Scale of Rate Ratios:} \quad e^{\hat{\beta}_j \pm z^* \cdot SE(\hat{\beta}_j)}$$

* **Interpretation:** We are $(1-\alpha)\times100\%$ confident that a 1-unit increase in $X_j$ is associated with a multiplicative change in the expected count by a factor between $e^L$ and $e^U$, holding all other variables constant.

## Inference for $\lambda$


1. Compute the predicted value and standard error on the **link (log) scale**:
   $$\hat{\eta} = \hat{\beta}_0 + \hat{\beta}_1 X_1 + \dots + \hat{\beta}_p X_p$$
2. Build the interval on the log scale: $[\hat{\eta} - z^*SE, \,\, \hat{\eta} + z^*SE]$
3. Exponentiate the endpoints to map back to the **original count scale**:
   $$\left[ e^{\hat{\eta} - z^*SE}, \,\, e^{\hat{\eta} + z^*SE} \right]$$

In R:

```r
predict(model, newdata = data.frame(x = ...), type = "link", se.fit = TRUE)
```


## Likelihood Ratio Test

When we want to test a collection of slopes simultaneously (e.g., evaluating whether the multi-level categorical variable `historical` can be dropped entirely), we use a **Likelihood Ratio Test (LRT)**.

$$\Delta G^2 = G^2_{\text{Reduced}} - G^2_{\text{Full}}$$

* **Standard Poisson GLM:** We compare $\Delta G^2$ to a $\chi^2$ distribution with $df = \Delta p$.

```{r}
#| eval: false

anova(reduced_glm, full_glm, test = "Chisq")
```

* **Quasipoisson GLM**: 🛑 We cannot use the standard $\chi^2$ distribution because the variance has been inflated by $\hat{\phi}$. Instead, we scale the drop in deviance and perform an F-test: $$F = \frac{\Delta G^2 / \Delta p}{\hat{\phi}}$$

R automatically handles the scale adjustment when you ask for test = "F":

```{r}
#| eval: false
anova(reduced_quasi, full_quasi, test = "F")
```

