---
title: Intro to Multiple Linear Regression
subtitle: Day 09
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: true
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

# wildfires stolen from Adam
# wildfires <- read.csv("https://aloy.github.io/stat230-materials/data/wildfires.csv")
# wildfires <- write.csv(wildfires, here("data", "wildfires.csv"))

wildfires <- read.csv(here("data", "wildfires.csv"))
wildfires <- filter(wildfires, Year >= 1960)
```


## `bikes` data {.smaller}

*Capital Bikeshare* is a bike sharing service in the Washington, D.C. area. To best serve its registered members, the company must understand the demand for its service. To help them out, we can analyze the number of rides taken on a random sample of *n* days. 

::::: columns
::: {.column width="60%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_histogram(~rides, data = bikes, xlab = "No. Rides", fill = "#003300", col = "white", bins = 25) 
```
:::

::: {.column width="40%"}

<br>
The data also has information about: 

- Weekend/holiday/etc
- Day of the measurement
- Weather information

:::
:::::

We're interested in predicting daily rides based on `temp_actual` and `windspeed`

## SLR model: `rides ~ temp_actual`

```{r}
bikes_slr <- lm(rides ~ temp_actual, data = bikes)

equatiomatic::extract_eq(bikes_slr, use_coefs = TRUE)
```

::::: columns
::: {.column width="50%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003300", size = 1) |>
  gf_lm(rides ~ temp_actual, size = 1.5, color = "#f15a31") 
```

:::


::: {.column width="50%"}

<br> $R^2 =$ `r round(summary(bikes_slr)$r.squared, 3)`

:::
:::::

## SLR model: `rides ~ windspeed`

```{r}
bikes_slr_2 <- lm(rides ~ windspeed, data = bikes)

equatiomatic::extract_eq(bikes_slr_2, use_coefs = TRUE)
```

::::: columns
::: {.column width="50%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_point(rides ~ windspeed, data = bikes, xlab = "Windspeed", ylab = "No. rides", color = "#003300", size = 1) |>
  gf_lm(rides ~ windspeed, size = 1.5, color = "#f15a31") 
```

:::


::: {.column width="50%"}

<br> $R^2 =$ `r round(summary(bikes_slr_2)$r.squared, 3)`

:::
:::::

## Geometric interpretation: 

This relationship is actually 3 dimensional:

```{r}
library(plotly)
p <- plot_ly(bikes, x = ~temp_actual, y = ~windspeed, z = ~rides, 
             type = "scatter3d", mode = "markers", color = ~rides)
p
```


## Multiple Linear Regression (MLR)



$$Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots + \beta_p x_{ip} + \varepsilon_i, \quad \varepsilon_i \overset{iid}{\sim} N(0, \sigma^2)$$

<br>

. . .

This model requires that...

1. $\mu\{Y|x_1, x_2, \ldots, x_p\}$ is a linear function

2. The error terms are independent 

3. The error terms follow a normal distribution with mean 0 and variance $\sigma^2$

## EDA: Pairs plot

```{r}
bikes |> 
  select(temp_actual, windspeed, rides) |>
  GGally::ggpairs()
```

## EDA: Correlation matrix

```{r}
#| echo: true


bikes |> 
  select(temp_actual, windspeed, rides) |>
  cor()
```


## Fitting the model

**Procedure:**

Least squares estimation: choose the coefficients to minimize 
$$\text{SSE} = \sum_{i=1}^n \left(Y_i - \widehat{Y}_i \right)^2$$
Then use $\widehat{\sigma} = \sqrt{\frac{\text{SSE}}{n-(k+1)}}$ where $k$ is the number of $x$ variables

## Fitting the model 

**R:** 

```{r}
#| echo: true
bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)
coef(bikes_mlr)
```

. . . 

`r equatiomatic::extract_eq(bikes_mlr, use_coefs = TRUE)`

## Interpreting $\widehat{\beta}_0$


```{r echo=FALSE}
# knitr::kable(tidy(supervisor_lm), format = "html", digits = 4)
coef(bikes_mlr) |> round(3)
```

- The expected value of the response variable when all explanatory variables are 0

- The expected overall number of rides is about -1525.538 when temperature = 0 and windspeed = 0 🫠

## Interpreting $\widehat{\beta}_j$ 


```{r echo=FALSE}
# knitr::kable(tidy(supervisor_lm), format = "html", digits = 4)
coef(bikes_mlr) |> round(3)
```



- The expected change in the response for a 1-unit increase in $x_j$, when all other variables are held constant.

- We expect 87.02 *additional* rides for every 1 degree increase in actual temperature, assuming constant windspeed

- We expect 38.63 *fewer* rides for every 1 MPH increase in windspeed, assuming constant temperature

- Compare these to the SLR estimates: $\hat \beta_{\text{temp}} = 89.23$ and $\hat \beta_{\text{wind}} = -55.5$

## Geometric interpretation: 

MLR "line" is actually a **plane**

```{r}
# code from gemini

# 2. Create sequences of values across the range of your raw data
x_seq <- seq(min(bikes$temp_actual), max(bikes$temp_actual), length.out = 50)
y_seq <- seq(min(bikes$windspeed), max(bikes$windspeed), length.out = 50)

# 3. Create a 2D prediction grid of all possible combinations
pred_grid <- expand.grid(temp_actual = x_seq, windspeed = y_seq)

# 4. Generate predictions from your model
z_pred <- predict(bikes_mlr, newdata = pred_grid)

# 5. Reshape the predictions into a matrix (required by add_surface)
z_matrix <- t(matrix(z_pred, nrow = length(x_seq), ncol = length(y_seq)))

# 6. Add the surface to your existing plot using the raw sequences!
p |> 
  add_surface(
    x = ~x_seq, 
    y = ~y_seq, 
    z = ~z_matrix, 
    opacity = 0.5,          # Make the plane slightly transparent
    showscale = FALSE       # Hide the color legend for the plane
  )
```

## Takeaways: 

- EDA for more than 1 predictor variable
- Write the theoretical MLR model
- Write the fitted MLR model
- Interpret the MLR coefficients
