---
title: Overdispersion and the Quasibinomial model
subtitle: Day 23
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)
library(ggResidpanel)
library(ggthemes)

notes_theme = theme_minimal(base_size = 14, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

news <- bayesrules::fake_news |>
  mutate(has_excl = ifelse(title_has_excl | text_excl > 0, TRUE, FALSE)) |>
  select(title, url, type, title_words, has_excl, title_caps_percent, negative, positive) |>
  mutate(type_num = as.numeric(type == "fake"))

news_mod <- glm(type_num ~ title_words + has_excl + title_caps_percent, 
                data = news, family = "binomial")

```

## `moths` from last homework

```{r}
moths <- read.csv("https://aluby.github.io/stat230/data/moths.csv") |>
  mutate(notremoved = Placed - Removed, 
               log_odds_removed = log(Removed / notremoved),
               prop_removed = Removed / Placed, 
               dark = ifelse(Morph=="dark",1,0), 
               distance = Distance, 
               morph = Morph) 

moths |>
  select(Morph, Distance, Placed, Removed)
```
 

## binomial logistic regression model

```{r}
moth_glm <- glm(Removed/Placed ~ morph*distance, weights = Placed,
                data = moths, 
                family = "binomial")
```

```{r}
tidy(moth_glm)
```

## Model assumptions

- **Binomial counts**

- **Independence**: observations are independent

- **Linearity**: the log odds is a linear function of the predictors

- **Variance structure**: the variance of a binomial random variable is $n \pi (1 - \pi)$

    $\Longrightarrow$ so the variance of $Y_i$ is $m_i \pi_i (1 - \pi_i)$ (not constant!)


## Goodness-of-fit test

Deviance can be used to assess the **overall fit** of binomial logistic regression

Observed | Predicted | True
:---------:|:---------:|:---------:
$Y_i$ | $\widehat{\pi}_i$ | $\pi_i$


<br>



**Deviance:** $D = 2 \sum \left[ Y_i \log \left( \dfrac{Y_i}{m_i\widehat{\pi}_i} \right)  + (m_i - Y_i)  \log \left( \dfrac{m_i - Y_i}{m_i - m_i \widehat{\pi}_i} \right) \right]$

<br>


:::{style="font-size:0.9em"}

$D \overset{\cdot}{\sim} \chi^2$ with $\text{d.f.}= n-(p+1)$, **if** the model is correct and all $m_i$ are large enough (>5)


$n =$ \# of binomial observations (\# of proportions, not # of moths)
:::


## Goodness-of-fit test

<div style="float: right; clear: right; margin: 10px;">(the model is adequate)</div>

:::{style="font-size:0.9em"}
H<sub>0</sub>: ${\rm logit}(\pi_i) = \beta_0 + \beta_1 x_{1i} + \cdots + \beta_p x_{pi}$
:::

<div style="float: right; clear: right; margin: 10px;">(the model is inadequate)</div>

:::{style="font-size:0.9em"}
H<sub>a</sub>: ${\rm logit}(\pi_i) = \alpha_i$
:::


<br>

```{r eval=FALSE}
summary(moth_glm)
```
```
Call:
glm(formula = Removed/Placed ~ Distance * Morph,
    family = binomial, data = case2102, weights = Placed)

    Null deviance: 35.385  on 13  degrees of freedom
*Residual deviance: 13.230  on 10  degrees of freedom
```

<br>

```{r}
#| echo: true
1- pchisq(13.230, 10)
```


<br>

:::{style="font-size:0.5em"}
Note: This test is *only valid for the binomial logistic regression*, it is not valid for binary logistic regression.
:::

## Communication Development Study


