---
title: Model Selection
subtitle: Day 1
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)
library(gt)
library(gtsummary)
library(carData)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}


bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)

```

## MLR so far

- Interpreting coefficients
- Writing lines for different groups (categorical predictors) 
- Inference for a single coefficient
- Inference for linear combinations of coefficients
- F tests
- Influential points and diagnostics (leverage, standardized residuals, cook's distance)
- Multicollinearity 

. . . 

Today &rarr; putting it all together to compare models

## Idea: how do we decide what variables to include? {.smaller}

```{r}
bikes |> head(10) |> knitr::kable()
```

## Review: SST/SSReg/SSE

```{r}
library(ggplot2)
library(patchwork)

# 1. Setup Data and Model
set.seed(42)
n <- 15
x <- seq(1, 10, length.out = n)
y <- 2 + 0.8 * x + rnorm(n, sd = 1.5)
df <- data.frame(x = x, y = y)

# Fit model and get predictions/mean
fit <- lm(y ~ x, data = df)
df$y_hat <- predict(fit)
df$y_mean <- mean(df$y)

# 2. Plot SST: Variation from the Mean (Total)
# Formula: sum((y - y_mean)^2)
p_sst <- ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hline(aes(yintercept = y_mean), linetype = "dashed", color = "blue") +
  geom_segment(aes(xend = x, yend = y_mean), color = "blue", alpha = 0.5) +
  geom_line(aes(y = y_hat), color = "darkgreen", linewidth = .5, alpha = .5) +
  labs(title = "SST", subtitle = "Distance: Obs to Mean") +
  theme_minimal()

# 3. Plot SSR: Variation Explained by Regression
# Formula: sum((y_hat - y_mean)^2)
p_ssr <- ggplot(df, aes(x, y)) +
  geom_point(alpha = 0.2) +
  geom_hline(aes(yintercept = y_mean), linetype = "dashed", color = "blue") +
  geom_line(aes(y = y_hat), color = "darkgreen", linewidth = 1) +
  geom_segment(aes(y = y_hat, xend = x, yend = y_mean), color = "darkgreen", alpha = 0.5) +
  labs(title = "SSReg", subtitle = "Distance: Fit to Mean") +
  theme_minimal()

# 4. Plot SSE: Unexplained Variation (Residuals)
# Formula: sum((y - y_hat)^2)
p_sse <- ggplot(df, aes(x, y)) +
  geom_point() +
  geom_line(aes(y = y_hat), color = "darkgreen", linewidth = 1) +
  geom_segment(aes(xend = x, yend = y_hat), color = "red", alpha = 0.5) +
  labs(title = "SSE", subtitle = "Distance: Obs to Fit") +
  theme_minimal()

# 5. Combine Plots
(p_sst | p_ssr | p_sse) 
```
 

## But, we work with sum of **squares**

```{r}
library(ggplot2)
library(patchwork)

# 1. Setup minimal data
set.seed(1)
n <- 6
x <- c(1, 2.5, 4, 6, 8, 9.5)
y <- c(1.5, 4, 3.5, 7, 7.5, 10)
df <- data.frame(x = x, y = y)

# Fit model
fit <- lm(y ~ x, data = df)
df$y_hat <- predict(fit)
df$y_mean <- mean(df$y)

# Helper function to create squares
# The side of the square is the absolute difference between y1 and y2
make_squares <- function(data, y1_col, y2_col, color) {
  data$diff <- data[[y1_col]] - data[[y2_col]]
  geom_rect(data = data, aes(
    xmin = x, 
    xmax = x + abs(diff), 
    ymin = pmin(data[[y1_col]], data[[y2_col]]), 
    ymax = pmax(data[[y1_col]], data[[y2_col]])
  ), fill = color, alpha = 0.3, color = color)
}

# 2. Plot SST: Total variation from the Mean
p1 <- ggplot(df, aes(x, y)) +
  geom_hline(aes(yintercept = y_mean), linetype = "dashed", color = "blue") +
  make_squares(df, "y", "y_mean", "blue") +
  geom_point() +
  labs(title = "SST", subtitle = "Squares of (Obs - Mean)") +
  theme_minimal() + coord_fixed() # coord_fixed ensures squares look like squares

# 3. Plot SSReg: Variation explained by the line
p2 <- ggplot(df, aes(x, y)) +
  geom_hline(aes(yintercept = y_mean), linetype = "dashed", color = "blue") +
  geom_line(aes(y = y_hat), color = "darkgreen") +
  make_squares(df, "y_hat", "y_mean", "darkgreen") +
  geom_point(alpha = 0.2) +
  labs(title = "SSReg", subtitle = "Squares of (Fit to Mean)") +
  theme_minimal() + coord_fixed()

# 4. Plot SSE: Residual variation (the 'error')
p3 <- ggplot(df, aes(x, y)) +
  geom_line(aes(y = y_hat), color = "darkgreen") +
  make_squares(df, "y", "y_hat", "red") +
  geom_point() +
  labs(title = "SSE", subtitle = "Squares of (Obs to Fit)") +
  theme_minimal() + coord_fixed()

# Combine
(p1 | p2 | p3) 
```

## 

::: medium

- Linear Regression tilt and shifts the line until the total area of the **red squares** are as small as possible ("least squares")
- $R^2$: proportion of the blue area (SSTotal) that is successfully covered by the green squares (SSRegr) 
- $h_i$: a point becomes more influential as it moves further away from $\bar{x}$ because the corresponding red square grows quadratically
- Overall F test: $F = \frac{MSReg}{MSE}$
    - is the "average" green square significantly larger than the "average" red square?
- Sequential sum-of-squares F test: $F=\frac{(SSE(reduced)-SSE(full))/(df_{reduced}-df_{full})}{SSE(full)/df_{full}}$
    - are the red squares for a "larger" model significantly smaller than the red squares for a "smaller" model?

:::

## Sequential ANOVA (the `anova()` function)

::: medium

When you use `anova()` on a single model, R provides a table of **extra Sum of Squares (SS)** for terms added sequentially


```{r}
#| echo: true
model_1 <- lm(rides ~ temp_actual + windspeed + weekend + temp_actual:weekend + 
                windspeed:weekend, data = bikes)
anova(model_1)
```

:::

## Breaking down the `anova()` table

::: {.medium .nonincremental}

The First Row: `temp_actual`  

  - Df: Including this variable adds 1 term to the model.  
  - Sum Sq: $SSE(\text{null}) - SSE(\text{temp_actual})$.  
  
The Second Row: `windspeed` 

  - Sum Sq: Represents the extra SS for adding `windspeed` to a model already containing `temp_actual`.  
  - $SSE(\text{temp_actual}) - SSE(\text{temp_actual, windspeed})$  


The Third Row: `weekend`   

  - Sum Sq: Represents the extra SS for adding `weekend` to a model already containing `temp_actual` and `windspeed`.  
  - $SSE(\text{temp_actual, windspeed}) - SSE(\text{temp_actual, windspeed, weekend})$  
  
:::

## Confirm results with a nested F test

"intermediate" p values in the table may not be valid; confirm proposed and full model with a nested F test:

```{r}
#| echo: true


model_2 <- lm(rides ~ temp_actual + windspeed + weekend, data = bikes)

anova(model_1, model_2)
```


## Comparing non-nested models

What if I want to compare:

```{r}
#| echo: true


model_2 <- lm(rides ~ temp_actual + windspeed + weekend, 
              data = bikes)
```

```{r}
#| echo: true


model_3 <- lm(rides ~ temp_actual + windspeed*weekend + 
                holiday + temp_feel , 
              data = bikes)
```

## Comparing non-nested models

Adjusted $R^2$

  - can be used for nested and non-nested models
  - no formal test/rules for "better enough" 
    
```{r}
#| echo: true


summary(model_2)$adj.r.squared
summary(model_3)$adj.r.squared
```

## Information criteria

The following "information criteria" can be used to compare nested or non-nested models (smaller values are better)

$$IC = \text{Measure of Fit} + \text{Penalty}$$

<br>

$$AIC = n \log(SSE/n) + 2p$$

<br>

$$BIC =  n \log(SSE/n) + \log(n) \cdot p$$

## AIC in R

```{r}
#| echo: TRUE

AIC(model_2)
AIC(model_3)
```

- suggests more complicated model is worth the added complexity

## BIC in R

```{r}
#| echo: TRUE

BIC(model_2)
BIC(model_3)
```

- BIC imposes a larger complexity penalty, so prefers the simpler model 

## Variable selection: prediction vs understanding 

:::: {.columns}

::: {.column width="45%"}
**Prediction:**

- More complex models might be okay if you have a lot of data.  

- The model doesn't need to "make sense" as much if it predicts well.  

- Watch out for overfitting!

- Stat270
:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}
**Understanding:**

- Complexity can hurt our ability to interpret results meaningfully.

- Keep effects you are interested in, even if they are insignificant, to "control" for them.
:::

::::

## AIC vs BIC

<br>

**AIC:** for large $n$, gives an estimate of out-of-sample predictive performance (i.e. better model for **prediction**)

<br>

**BIC:** for large $n$, will pick the "true" model if it is one of the candidates (i.e. better model for **understanding**)


## Overfitting 

::: medium

The goal when building a model is to capture the trends in the data while still making something that can applied to new data. When a model is **overfit**, a model will perform very well on the data used to fit it but perform very poorly on any new data.
:::

```{r}
#| fig-width: 8
#| fig-height: 2


library(ggplot2)
library(patchwork)

# 1. Generate Synthetic Data (A simple curve with noise)
set.seed(123)
n <- 30
x <- seq(0, 1, length.out = n)
y <- sin(2 * pi * x) + rnorm(n, sd = 0.3)
df <- data.frame(x = x, y = y)

# Helper function to generate plots
plot_fit <- function(degree, title) {
  ggplot(df, aes(x, y)) +
    geom_point(alpha = 0.6, color = "black") +
    stat_smooth(method = "lm", formula = y ~ poly(x, degree), 
                se = FALSE, color = "#f15a31", linewidth = 1) +
    labs(title = title) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))
}

# 2. Underfit: Linear model (too simple to capture the curve)
p1 <- plot_fit(1, "Underfit")

# 3. Just Right: Degree 3 (captures the signal but ignores noise)
p2 <- plot_fit(3, "Balanced")

# 4. Overfit: Degree 15 (chases every random noise point)
p3 <- plot_fit(15, "Overfitt")

# 5. Combine using Patchwork
(p1 | p2 | p3)
```

::: medium

Model building is like making a suit that can generally fit everyone around a given size, not making a custom-tailored suit that can only fit one person perfectly. 

:::

## Prediction: Training & Test Datasets   

If the goal is prediction, we test the model on data it hasn't seen:

  - Training dataset: Used to estimate ("train") your model.  
  - Test dataset: The data held out to test your predictions.  
  - RMSE: Root Mean Squared Error measures the average size of error (squared residuals).
  
$$RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}$$

## Code: make training and testing data

```{r}
#| echo: true

set.seed(050726)

bikes_split <- initial_split(bikes, prop = .8)
bikes_split

bikes_train <- training(bikes_split)
bikes_test <- testing(bikes_split)
```

## Code: fit both models on training data

```{r}
#| echo: true


model_1_train <- lm(rides ~ temp_actual + windspeed + weekend + 
                      temp_actual:weekend +  windspeed:weekend,
                     data = bikes_train)
model_3_train <- lm(rides ~ temp_actual + windspeed * weekend + 
                      holiday + temp_feel,
                      data = bikes_train)
```


## Code: find $\widehat{y}$ for both models, then RMSE

```{r}
#| echo: true


model_1_test <- predict(model_1_train, newdata = bikes_test)
model_3_test <- predict(model_3_train, newdata = bikes_test)

sqrt(mean((bikes_test$rides - model_1_test)^2))
sqrt(mean((bikes_test$rides - model_3_test)^2))
```

## Code: find RMSE for **training data**

```{r}
#| echo: true


sqrt(mean(model_1_train$residuals^2))
sqrt(mean(model_3_train$residuals^2))
```

::: medium

- RMSE is smaller on the model 3 for training data (predicting better)
- RMSE is **bigger** using the model 3 for testing data (predicting worse)
    - when the model is used on data it hasn't seen before, it's doing worse
    - this is what it means to overfit
:::


## Tools for Model Comparison 

::: medium

1. ANOVA F-tests: Used for nested models to test if coefficients equal zero.  
2. Adjusted $R^2$: Can be used for non-nested models; penalizes adding unnecessary predictors.
3. AIC or BIC
    - Smaller values are better.  
    - Penalizes high numbers of coefficients to prevent over-complexity
    
Use on the full data if you are building an explanatory model, or on the training data *only* if you are building a predictive model 

:::

## Inference after selection

- Model selection results in keeping terms with smaller p-values
- Actual Type I error may be much bigger than than "expected" $\alpha = .05$
- This is an active area of research! 

&rarr; if you rely on AIC/BIC/RMSE and compare *lots* of models, your p-values and confidence intervals might not be valid


## A Note on Automatic Selection

I do **not** recommend using automated "Stepwise" (Forward/Backward) selection.  

- Why? They are easily abused and ignore domain knowledge, outliers, and collinearity.  
- Better Approach:

   > "The single most important tool in selecting a subset of variables is the analyst’s knowledge of the area under study." — Sandy Weisberg. 
  
## Summary of Variable Selection

- Identify research goals and use domain knowledge.  
- Perform EDA and consider transformations.  
- Compare models using ANOVA/Adj $R^2$/BIC (for nested) or AIC/RMSE (for prediction).  
- **Justify your model choice!**
