---
title: Inference for MLR
subtitle: Day 11
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
library(here)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

regression_panel <- function(formula, scat.formula = NULL, data, xlab, ylab) {
  mod <- lm(formula, data = data)
  
  aug <- augment(mod)
  
  plot_fit <- makeFun(mod)
  
  if(is.null(scat.formula)) {
    orig_data <- gf_point(formula, data = data, xlab = xlab, ylab = ylab, color = "#003300")
  } else {
    orig_data <- gf_point(scat.formula, data = data, xlab = xlab, ylab = ylab, color = "#003300") 
  }
  
  orig_data <- orig_data %>% gf_fun(plot_fit(x) ~ x, color = "#f15a31")
  
  resid_plot <- gf_point(.std.resid ~ .fitted, data = aug, xlab = "Fitted values", ylab = "Standardized residuals") %>%
    gf_hline(yintercept = 0, linetype = 2, color = "#f15a31")
  
  qq_plot <- gf_qq(~.std.resid, data = aug, xlab = "N(0,1) quantiles", ylab = "Standardized residuals", color = "#003300") %>%
    gf_qqline(color = "#f15a31")
  
  (orig_data | resid_plot | qq_plot)
}

bikes <- bikes |> select(date, season, year, month, weekend, holiday, temp_actual, temp_feel, humidity, windspeed, rides)

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt", sep = "\t", header = TRUE)

```

## Warm up: new groups!

::: task

- Introduce yourselves
- Group expectations worksheet

:::


## `bikes`

```{r}
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes) 
```

## MLR model

```{r}
bikes_mlr <- lm(rides ~ temp_actual*weekend, data = bikes)
bikes_mlr
```

::: task
Write out the theoretical and fitted regression model
:::

## 


## Do we _need_ a different slopes model? 

```{r}
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes) |>
  gf_lm()

tidy(bikes_mlr) 
```



## Testing a single coefficient

::: {style="font-size: 0.9em;"}
**Hypotheses:** $H_0: \ \beta_j = \#$ vs. $H_a: \ \beta_j \underset{>}{\overset{<}{\ne}} \#$
:::

. . .

::: {style="font-size: 0.9em;"}
**Test statistic:** $t = \dfrac{\hat{\beta}_j - \#}{SE(\beta_j)}$
:::
. . .

::: {style="font-size: 0.9em;"}
**Reference distribution:** $t$ distribution with d.f. = $n-(k+1)$


**p-value:** Area in the tail(s) specified by $H_a$
:::


```{r echo=FALSE, fig.height = 2, fig.width = 7.5, out.width=850}
p1 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] < 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p2 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = .4) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] != 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

p3 <- ggplot(data = data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = .4) +
  ggtitle(expression(H[a]: beta[j] > 0)) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

(p1 | p2 | p3)
```

## CIs for a single coefficient


$\widehat{\beta}_j \pm t^*_{n-(k+1)} \cdot SE(\widehat{\beta}_j )$


## Answer: Do we need a different slopes model? 


## Do we need `weekend` at all?


```{r}
tidy(bikes_mlr) 
```

::: callout-caution
## Caution!

<br>

<br>

<br>

<br>

:::

## Do we need `weekend` at all?

```{r}
bikes_mlr2 <- lm(rides ~ temp_actual + weekend, data = bikes)
tidy(bikes_mlr2)
```


## Explained Variability


::: columns
:::{.column width="30%"}
```{r}
#| fig-width: 1.75
#| fig-height: 3
#| echo: FALSE
#| out-width: 100%
sst_plot + decomp_plot
```
:::
:::{.column width="60%"}
<br>

<br>



$\begin{split}
R^2 &= `r round(glance(quad_mod)[["r.squared"]], 3)`\\
&= \dfrac{SSRegr}{SST} \\
&= 1 - \dfrac{SSE}{SST} 
\end{split}$
:::
:::

# Idea: is the proportion of variability explained by Model 1 significantly different than Model 2?

## F Test





