---
title: Evaluating SLR models
subtitle: Day 05
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format: 
  revealjs:
    incremental: false
    scrollable: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(patchwork)
library(fontawesome)
notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

city <- readr::read_csv("https://raw.githubusercontent.com/zief0002/modeling/main/data/riverview.csv")
```

## Topics for Exam I


## Last Time

1. Inference for predictions

    - $\widehat{\mu}(Y|X)$
    - $\text{Pred}(Y|X)$
    
### Today: how "good" is our model?

## Two definitions for "good"

<br> 

1. The model is appropriate and our inferences are valid

<br> 

2. The X variable does a good job at explaining the Y variable

## Notation

<br> 

Theoretical model: 

$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br> 

Fitted model: 

$\widehat{y}_i = \widehat\beta_0 + \widehat\beta_1 x_i$

$y_i = \widehat{y}_i + \widehat{\varepsilon}_i$
 

## Riverview Data

```{r}
#| fig-width: 4
#| fig-height: 2.5
#| fig-align: 'left'

gf_point(income ~ education, data = city, xlab = "Education (years)", ylab = "Income ($1000s)") |>
  gf_lm()

city_lm = lm(income ~ education, data = city)
```


```{r}
lm(income ~ education, data = city)
```

## Conditions for inference 

Our model must be valid in order for our p-values or confidence intervals to be valid: 

$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br>

Conditions to check:

- **L**inear relationship is appropriate
- **I**ndependent and identically distributed errors (iid)
- **N**ormally distributed errors
- Errors have **E**qual variance


## Residuals


**Definition**: $e_i =\widehat{\varepsilon}_i = y_i - \widehat{y}_i$

<br>


**Properties:**

-  sum to zero $\Longrightarrow$ mean is 0

-  **uncorrelated** with $x$ and $\widehat{y}$

- normally distributed 
 
- $SD(e_i) = \widehat{\sigma} \sqrt{1 - \dfrac{1}{n} - \dfrac{(x_i - \overline{x})^2}{\sum (x_i - \overline{x})^2}}$


## What is a residual plot?


```{r}
#| fig-width: 8
#| fig-height: 3


p1 <- ggplot(broom::augment(city_lm), aes(x = education, y = income)) +
  geom_segment(aes(xend = education, yend = .fitted), col = "darkorange") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5)

p2 <- ggplot(broom::augment(city_lm), aes(x = .fitted, y = .resid)) +
  geom_segment(aes(xend = .fitted, yend = 0), col = "darkorange") +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, linewidth = 1.5, col = "blue")

p1 + p2
```




## A "good" residual plot

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| layout-ncol: 2
set.seed(72384)
x <- rnorm(100, mean = 10, sd = 3.5)
y <- 1 + 1*x + rnorm(100, mean = 0, sd = 10)
good_mod <- lm(y ~ x)

gf_point(residuals(good_mod) ~ fitted(good_mod), 
         ylab = "Residual", xlab = "Fitted values") |>
  gf_hline(yintercept = 0, color = "blue", linetype = 2) 

gf_point(residuals(good_mod) ~ x, 
         ylab = "Residual", xlab = "x") |>
  gf_hline(yintercept = 0, color = "blue", linetype = 2) 
```

## Your turn

::: {.task}
- Work in groups

- On the whiteboards, sketch a plot of $y$ vs. $x$ and a corresponding residual plot that would indicate a violation of the

  1. linearity condition

  2. equal variance condition
    
:::

## Assessing Linearity

```{r}
#| fig-width: 8
#| fig-height: 3

p1 <- gf_blank(residuals(good_mod) ~ fitted(good_mod),
         ylab = "Y variable", xlab = "X variable") +
          theme(axis.text = element_blank()) +
          labs(title = "Violates linearity")

p2 <- gf_blank(residuals(good_mod) ~ fitted(good_mod),
         ylab = "Residual", xlab = "Fitted values") +
          theme(axis.text = element_blank()) +
          labs(title = "Violates linearity")

p1 + p2
```

## Assessing Equal Variance

```{r}
#| fig-width: 8
#| fig-height: 3

p1 <- gf_blank(residuals(good_mod) ~ fitted(good_mod),
         ylab = "Y variable", xlab = "X variable") +
          theme(axis.text = element_blank()) +
          labs(title = "Violates equal variance")

p2 <- gf_blank(residuals(good_mod) ~ fitted(good_mod),
         ylab = "Residual", xlab = "Fitted values") +
          theme(axis.text = element_blank()) +
          labs(title = "Violates equal variance")

p1 + p2
```

## Assessing Normality

- histogram of residuals
- normal Q-Q plot of residuals

**Examples of "good" plots:**

```{r}
#| echo: false
#| layout-ncol: 4
#| fig.height: 3.5
#| fig.width: 5
x2 <- rnorm(30, mean = 10, sd = 3.5)
y2a <- 1 + 1*x2 + rnorm(30, mean = 0, sd = 10)
y2b <- 1 + 1*x2 + rnorm(30, mean = 0, sd = 10)
good_mod2a <- lm(y2a ~ x2)
good_mod2b <- lm(y2b ~ x2)

gf_histogram(~residuals(good_mod2a), color = "white", bins = 12, xlab = "Residual", ylab = "Count") 

gf_histogram(~residuals(good_mod2b), color = "white", bins = 12, xlab = "Residual", ylab = "Count")

gf_qq(~residuals( good_mod2a), ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline() 

gf_qq(~residuals( good_mod2b), ylab = "Residual", xlab = "N(0, 1) quantiles") |>
  gf_qqline()
```

::: {.callout-note}
Normality is **harder to judge at small sample sizes** -- even true random draws from a normal distribution can look lumpy or have a wandering Q-Q plot when $n$ is small. Don't over-interpret minor wiggles with only 30-50 points.
:::

## Assessing independence

- plot residuals vs. variable inducing dependence (e.g. time, location, subject ID)

**Examples of "good" plots:**

```{r}
#| echo: false
#| layout-ncol: 2
#| fig.height: 3.5
#| fig.width: 5


gf_point(rstandard(good_mod2a) ~ 1:30, ylab = "Standardized Residual", xlab = "Order") |>
  gf_line() |>
  gf_hline(yintercept = 0)

gf_point(rstandard(good_mod2b) ~ 1:30, ylab = "Standardized Residual", xlab = "Order") |>
  gf_line() |>
  gf_hline(yintercept = 0)
```

# Making residual plots in R

## `augment()`

The `augment()` function (from {broom}) adds `.fitted` and `.resid` columns to your data, so you don't have to compute them by hand.

```{r}
#| echo: true
city_aug <- augment(city_lm)
city_aug
```

## Residual plots in R

```{r}
#| echo: true
#| layout-ncol: 4
#| fig-width: 8
#| fig-height: 5
gf_point(.resid ~ .fitted, data = city_aug)
gf_histogram(~.resid, data = city_aug, col = "white")
gf_qq(~ .resid, data = city_aug) |> gf_qqline()
gf_point(.resid ~ 1:nrow(city_aug), data = city_aug)
```

# Your turn

::: {.task}

Work through Example 1 on the activity

:::

## What happens the conditions aren't valid? 


-   **Linearity:** if nonlinear, everything breaks!

-   **Independence:** estimates are still unbiased (i.e. we fit the right line) but measures of the accuracy of those estimates (the SEs) are typically too small

-   **Normality:** estimates are still unbiased (i.e. we fit the right line), SEs are correct BUT prediction intervals (which rely on the data points being normally distribued around the mean) are wrong 

-   **Equal variance:** estimates are still unbiased but standard errors are wrong (and we don't know how wrong)


## What do we do if our assumptions are violated?

1. Change our assumptions (hard, need more stats)

2. Transform $y$, $x$, or both

3. Change the type of inference (remember the bootstrap?)

# Is the X variable a "good" predictor of the Y variable? 

## 

```{r}
#| fig-width: 8
#| fig-height: 4
#|
p1 <- gf_point(income ~ education, data = city) |>
  gf_lm() +
  xlim(c(5, 29)) +
  ylim(c(0, 140))

p2 <- tibble(
  x = rnorm(50, sd = 4.5) + 16,
  y = 11.3 + 2.65*x + rnorm(50, sd = 32)
) |>
  gf_point(y ~ x) |>
  gf_abline(slope = 2.65, intercept = 11.3, col = "#3366FF", linewidth = 1) +
  xlim(c(5, 29)) +
  ylim(c(0, 140))

p3 <- tibble(
  x = rnorm(50, sd = 4.5) + 16,
  y = 11.3 + 2.65*x + rnorm(50, sd = 2)
) |>
  gf_point(y ~ x) |>
  gf_lm() +
  xlim(c(5, 29)) +
  ylim(c(0, 140))

p1 + p2 + p3
```

## Partitioning variability

```{r}
#| fig-width: 9
#| fig-height: 3.2

slr_aug <- broom::augment(city_lm) |> mutate(mean.y = mean(income))

p1 <- ggplot(slr_aug, aes(x = education, y = income)) +
  geom_segment(aes(xend = education, yend = mean.y), col = "darkorange") +
  geom_point() +
  geom_hline(yintercept = mean(slr_aug$income), col = "blue", linewidth = 1) +
  labs(title = "SS Total") + 
  ylim(20, 85)

p2 <- ggplot(slr_aug, aes(x = education, y = .fitted)) +
  geom_segment(aes(xend = education, yend = mean.y), col = "darkorange") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, col = "blue", linewidth = 1) +
  geom_hline(yintercept = mean(slr_aug$income), col = "blue", linetype = "dashed") +
  labs(title = "SS Regression", y = "income") + 
  ylim(20, 85)

p3 <- ggplot(slr_aug, aes(x = education, y = income)) +
  geom_segment(aes(xend = education, yend = .fitted), col = "darkorange") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, col = "blue", linewidth = 1) +
  labs(title = "SS Error") + 
  ylim(20, 85)

p1 + p2 + p3
```

## $R^2$

\begin{align*}
\sum(Y - \bar{Y})^2 &= \sum (\hat{Y} -\bar{Y})^2  \hspace{.9cm} + \sum e_i^2 \\
\text{total variability} &= \text{variability of } \hat{Y}\text{'s} + \text{variability of residuals}\\
\end{align*}

Then:

$$R^2 = \frac{\sum (\hat{Y} - \bar{Y})^2}{\sum (Y - \bar{Y})} = \frac{\text{variability of } \widehat{Y}}{\text{variability of }Y} $$


::: callout-note
## $R^2$

The $R^2$ value gives the proportion of total variability in Y explained by the SLR model
:::


## Notation reference

| Symbol | Meaning |
|:---|:---|
| $r$ | sample correlation coefficient |
| $\rho$ | population correlation coefficient |
| $R^2$ | coefficient of determination |

## Finding $R^2$ in R {.smaller}

```{r}
#| echo: true
summary(city_lm)
```

and you can access it directly with `$`:

```{r}
#| echo: true
summary(city_lm)$r.squared
```

## Properties of $R^2$

- Between 0 and 1
- Big when $\sigma$ is small relative to the overall variability
    - big proportion of variability explained by the model
- For SLR, $R^2 = r^2$ 
    - ($r$ = correlation coefficient between X and Y)

## $R^2$ dangers


```{r}
#| echo: false

set.seed(040726)

fake_data <- tibble(
  x = c(runif(50, 0, 200), 400),
  y = c(.5 * x[1:50]^2 + rnorm(50, 0, 50), 30000)
)

fake_lm <- lm(y ~ x, data = fake_data)

p1 <- ggplot(fake_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = paste0("R² = ", round(summary(fake_lm)$r.squared, 3))
  )

p2 <- ggplot(broom::augment(fake_lm), aes(x = .fitted, y = .resid)) +
  geom_point()

p1 + p2

# cor(fake_data$x, fake_data$y)
```

# Your turn

::: {.task}

Work through Example 2 on the activity

:::
