---
title: Simple Linear Regression (SLR)
subtitle: Day 02
title-slide-attributes:
  data-background-color: "#003300"
  data-slide-number: none
format:
  revealjs:
    incremental: false
    scrollable: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(countdown)
library(tidyverse)
library(ggformula)
library(rsample)
library(broom)
library(gganimate)
library(ggdist)
library(patchwork)

notes_theme = theme_minimal(base_size = 12, base_family = "Source Sans 3")
theme_set(notes_theme)

nhanes <- read_csv(here::here("data/nhanes-sitting-depression.csv")) 

median_hrs <- median(nhanes$sedentary_hours)

nhanes <- nhanes |>
  mutate(dep_score = log(phq9 + 1),
          sitting_group = if_else(sedentary_hours > median_hrs, "High", "Low"))
```

# Meet your groups! 

Name, pronouns, year, last stats class
```{r}
countdown::countdown(5)
```

## Group work: highs and lows

- Think about some people you have liked working with in the past. What have they done well? 
- Think about some people you have *not* liked working with in the past. What made them hard to work with? 

```{r}
countdown::countdown(1.5)
```

## In your groups 

Fill out the proposed engagement rubric on your sheets. I'll compile these after class and we'll discuss them next time. 

```{r}
countdown::countdown(5)
```

## Recap of last time {.smaller}

The *National Health and Nutrition Examination Survey* (NHANES) is an ongoing CDC survey that collects health and behavioral data from a nationally representative sample of the U.S. population. participants aged 18-25 were classified into two groups: those who reported **more** sitting time than average ("High"), and those who reported **less** ("Low"). Their depression symptoms were measured with the PHQ-9, a standard 9-item screening scale (0-27, higher = more symptoms). This was translated into a **depression risk score**, with a larger number indicating a higher risk. 

Last time, we used the sample statistics and our trusty intro stat skills to show that there was **no significant difference** between the two groups.

## Let's take a closer look at the data:

::::: columns
::: {.column width="60%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false


gf_boxplot(dep_score ~ sitting_group, data = nhanes)
```

:::

::: {.column width="40%"}
<br>

- How do you think "High" and "Low" sitting group were defined?
:::
:::::

## Data for today: the original variable {.smaller}

Survey participants were asked to report how many hours they spend sitting per day, on average.

::::: columns
::: {.column width="60%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_histogram(~sedentary_hours, data = nhanes, xlab = "Hours spent sedentary/sitting", fill = "#003300", col = "white", bins = 15) |>
  gf_vline(xintercept = median_hrs, col = "darkred", linetype = "dashed")
```
:::

::: {.column width="40%"}

<br>
The data also has information about:

- Age
- Sex

For our first analysis, we'll investigate the relationship between **sitting time** and **depression symptoms**
:::
:::::

## What do you notice that you couldn't see before?

```{r}
#| fig.align: center
#| out.width: 100%
#| echo: false

p1 <- gf_histogram(~sedentary_hours, data = nhanes, xlab = "Hours spent sedentary/sitting", fill = "#003300", col = "white", bins = 15) 

p2 <- gf_histogram(~dep_score, data = nhanes, xlab = "Depression Risk Score", fill = "#003300", col = "white", bins = 15)

p1 + p2
```

# Regression in Intro

##

::::: columns
::: {.column width="60%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
#| echo: false
gf_jitter(dep_score ~ sedentary_hours, data = nhanes, xlab = "Sitting time (hrs/day)", ylab = "Depression risk score", color = "#003300", size = 1, width = 0.15, height = 0.3) |>
  gf_lm(dep_score ~ sedentary_hours, size = 1.5, color = "#f15a31")
```
:::

::: {.column width="40%"}

<br>

-   Describe the scatterplot
-   Write the equation of the regression line
-   Interpret the slope and intercept
-   Make a prediction
:::
:::::

## Q: Describe the scatterplot

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 70%
#| echo: false
 gf_jitter(dep_score ~ sedentary_hours, data = nhanes, xlab = "Sitting time (hrs/day)", ylab = "Depression risk score", color = "#003300", size = 1, width = 0.15, height = 0.3) # |>
#   gf_lm(dep_score ~ sedentary_hours, size = 1.5, color = "#f15a31")
```

::::: columns
::: {.column width="50%"}

:::{.small}
::: {.conceptest}
1. There is a strong, positive, linear association between sitting time and depression risk score.

2. There is a weak, positive, linear association between sitting time and depression risk score.
:::
:::

:::

::: {.column width="50%"}


:::{.small}
::: {.conceptest}
3. There is a weak, positive, non-linear association between sitting time and depression risk score.

4. There is no clear association between sitting time and depression risk score.
:::
:::

:::
:::::


## Q: Write the equation of the regression line

```{r}
#| echo: true


lm(dep_score ~ sedentary_hours, data = nhanes)
```

What is the equation of the regression line?

<br>


## Q: Interpret the slope in context

::: {.conceptest}
1. For every one-hour increase in sitting time, the model predicts an average decrease of 0.03 in depression risk score

1. For every one-hour increase in sitting time, the model predicts an average increase of 0.03 in depression risk score

1. If someone reports 0 hours of sitting time, we'd expect a depression risk score of 0.03

1. If sitting time increases by one hour, it will cause an increase of 0.03 in depression risk score every day
:::


## Q: Make a prediction

A participant reports sitting for 6 hours per day. What does the model predict for their depression risk score?


# Regression in Stat230

## Overview: The Normal Distribution

::::: columns
::: {.column width="50%"}
<br>

```{r echo=FALSE, normal, message=FALSE, fig.width=12}

# Set up gamma data
mu <- c(0)
sigma  <- c(1)
sigma.sq<-sigma^2
normals <- data.frame(
                    x = rep(seq(-5, 5, length = 500), 1),
                    mu = rep(mu, each = 500),
                    sigma = rep(sigma, each = 500))

normals <- normals %>%
    mutate(y = dnorm(x, mean = mu, sd = sigma))

ggplot(normals, aes(x = x, y = y)) +
  geom_line(linewidth = 2) +
  #facet_wrap(~ setting) +
  labs(x = "y", y = "f(y)")

```

:::

::: {.column width="50%"}

<br>
If Y is:

-   **numeric**
-   **unimodal**
-  and **symmetric**,

we might use a Normal model
:::
:::::

## $Y \sim N(\mu, \sigma^2)$

```{r echo=FALSE, normal-tuning, message=FALSE, fig.width=12}

# Set up gamma data
mu <- c(2,2,4)
sigma  <- c(.5,1,2)
sigma.sq<-sigma^2
normals <- data.frame(setting = factor(rep(1:3, each = 500)),
                    x = rep(seq(-2, 10, length = 500), 3),
                    mu = rep(mu, each = 500),
                    sigma = rep(sigma, each = 500))
normals <- normals %>%
    mutate(y = dnorm(x, mean = mu, sd = sigma))
levels(normals$setting) <- paste0("N(",mu,",",sigma.sq,")")
g <- ggplot(normals, aes(x = x, y = y)) +
  geom_line() +
  facet_wrap(~ setting) +
  labs(x = "y", y = "f(y)") +
  scale_y_continuous(breaks = c(0,2,4,6,8,10)) +
  lims(y = c(0,0.8))

g
```


## A deterministic model

- $Y_i = \beta_0 + \beta_1 x_i$

- No uncertainty!

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

nhanes |>
  ggplot(aes(x = sedentary_hours, y = dep_score)) +
  # geom_point(alpha = 0.6, color = "#003069") +
  geom_lm(color = "#f15a31") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

## Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

 nhanes_lm <- lm(dep_score ~ sedentary_hours, data = nhanes)
 nhanes_dsn <- data.frame(
   sedentary_hours = c(2, 6, 10, 14),
   yhat = predict(nhanes_lm, data.frame(sedentary_hours = c(2, 6, 10, 14))),
   resid_sd = rep(sigma(nhanes_lm), 4)
 )

# mod <- lm(Head ~ Mass, data = blue_jays_male)
#
# blue_jay_dsn <-
#   tibble(
#     Mass = c(60, 65, 70, 75, 80),
#     yhat = predict(mod, newdata = data.frame(Mass = c(60, 65, 70, 75, 80))),
#     resid_sd = sigma(mod)
#   )

nhanes |>
  ggplot(aes(x = sedentary_hours, y = dep_score)) +
  stat_slab(aes(x = sedentary_hours, ydist = "norm", arg1 = yhat, arg2 = resid_sd), data = nhanes_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  geom_lm(color = "#f15a31") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

## Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

predict_dep_score <- function(n, x) {
  rnorm(n, mean = predict(nhanes_lm, data.frame(sedentary_hours = x)), sd = sigma(nhanes_lm))
}

set.seed(1234)

mod_sims <- tibble(
  sedentary_hours = rep( c(2, 6, 10, 14), each = 20),
  dep_score = c(predict_dep_score(20, 2), predict_dep_score(20, 6), predict_dep_score(20, 10), predict_dep_score(20, 14))
)

p <- nhanes |>
  ggplot(aes(x = sedentary_hours, y = dep_score)) +
  stat_slab(aes(x = sedentary_hours, ydist = "norm", arg1 = yhat, arg2 = resid_sd),
            data = nhanes_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  geom_point(data = mod_sims, aes(sedentary_hours, dep_score), inherit.aes = FALSE, alpha = 0.4, color = "#003300") +
  geom_lm(color = "#f15a31") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
p
```

## Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 5
#| fig.align: center

nhanes |>
  ggplot(aes(x = sedentary_hours, y = dep_score)) +
  geom_jitter(alpha = 0.6, color = "#003300", width = 0.15, height = 0.3) +
  geom_lm(color = "#f15a31") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

## Simple Linear Regression Model

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br>

-   Linear relationship between $x$ and $y$
-   Errors are independent and identically distributed (iid)
-   Errors are normally distributed
-   Errors have mean 0
-   Variance of the errors doesn't depend on $x$

## Notation

Theoretical Model: 

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$, $\varepsilon \sim N(0, \sigma^2)$

<br> 

Mean function:

$\quad E(Y|X)=\mu\{Y|X\} = \beta_0 + \beta_1 x_i$

<br>

Fitted model equation:

$\quad \widehat{Y}_i = \widehat{\mu}\{Y|X\} = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$

## Sampling Variability

If we had a slightly different sample of people in the survey, our line would look a little bit different

```{r}
#| echo: false
#| fig.height: 2.5
#| fig.width: 6

set.seed(1234)

boots <- bootstraps(nhanes, times = 20, apparent = TRUE)

fit_lm_on_bootstrap <- function(split) {
    lm(dep_score ~ sedentary_hours, data = analysis(split))
}

boot_models <-
  boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))

boot_coefs <-
  boot_models %>%
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

boot_aug3 <-
  boot_models %>%
  slice(1:3) %>%
  mutate(augmented = map(model, augment)) %>%
  unnest(augmented)

labels <- c("Bootstrap01" = "Sample 1", "Bootstrap02" = "Sample 2", "Bootstrap03" = "Sample 3")

ggplot(boot_aug3, aes(sedentary_hours, dep_score)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = .fitted, group = id), col = "#00aeef") +
  facet_wrap(~id, labeller = as_labeller(labels)) +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

## How different?

```{r}
#| echo: false
#| fig.height: 3
#| fig.width: 5
hop <-
  ggplot(data = boot_coefs) +
  geom_abline(aes(slope = sedentary_hours, intercept = `(Intercept)`, group = id), color = "#00aeef", size = 0.5) +
  geom_point(data = nhanes, aes(sedentary_hours, dep_score), color = "grey60", inherit.aes = FALSE) +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score") +
  theme(
    strip.text = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    # plot.margin = margin(7, 1.5, 3.5, 1.5)
  ) +
  # panel_border() +
  transition_states(id, 0, 1) +
  shadow_mark(future = TRUE, color = "gray50", alpha = 1/20)

animate(hop)
```

## How different?

```{r}
#| echo: false
#| fig.width: 5
#| fig.height: 3
ggplot(data = boot_coefs) +
  geom_abline(aes(slope = sedentary_hours, intercept = `(Intercept)`, group = id), color = "#0072B2", size = 0.5, alpha = 0.2) +
  # theme_dviz_grid() +
  geom_point(data = nhanes, aes(sedentary_hours, dep_score), color = "grey60", inherit.aes = FALSE) +
  theme(
    # strip.text = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.margin = margin(7, 1.5, 3.5, 1.5)
  ) +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

## Rest of today

We're going to see how to make some graphs and fit SLR models in R! We'll focus on commands today, and talk about the maize server and .qmd files on Friday.

::: {.task .nonincremental}
- Open the SLR in R activity on Moodle and complete in your groups
- Make note of any questions
- Choose someone responsible for reporting out your answers
:::

<!-- For next time: include definition of residual. Removed sampling variability -->
