---
title: "03: Inference for SLR"
execute:
    echo: true
    warning: false
    message: false
---

```{r}
library(tidyverse)
library(broom)
library(ggformula)
```


# Part 2: In R[^1]

[^1]: Adapted from *Stat 2: Modeling with Regression and ANOVA*


## Load the data

Load data into your session (it should show up in the "Environment" pane)

```{r}
LeafWidth <- read_csv("https://aluby.github.io/stat230/data/LeafWidth.csv")
```


"Print" the data: 

```{r}
LeafWidth
```

## Exploratory Data Analysis

First, make a scatterplot of `Width` (response variable) against `Year` (explanatory variable). Overlay the linear regression line. Describe your findings.

::: callout-tip
## Solution

There's a weak, negative, linear relationship between Width and Year. 

```{r}
gf_point(Width ~ Year, data = LeafWidth) |>
    gf_lm()
```

:::

## Linear Regression Model

Find the linear regression equation using `lm()`. Use `tidy()` to print the coefficient table of the regression model. Make sure you get the same answer as the paper printout! 

::: callout-tip
## Solution

```{r}
model <- lm(Width ~ Year, data = LeafWidth)
tidy(model)
```

:::

## CI for Slope

Calculate a 95% confidence interval for the slope of the regression line using `confint`. 

::: callout-tip
## Solution

We are 95% confident that, for each year increases, Width decreases by between .008 and .026 mm on average. This supports the hypothesis that leaves are getting smaller over time (which could be related to temperature, but we can't actually tell that with this analysis). 

```{r}
confint(model)
```

:::



