---
title: "Day 03: Inference for SLR"
author: "Name: ________________________________"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(broom)
library(rsample)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")
bikes_lm <- lm(rides ~ temp_actual, data = bikes)
```

## Data for today

*Capital Bikeshare* is a bike sharing service in the Washington, D.C.
area. We're analyzing the relationship between number of rides taken
and temperature.

```{r}
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides",
         color = "#003300", size = 1) |>
  gf_lm(rides ~ temp_actual, linewidth = 1.2, color = "#f15a31")
```

## Last time

1.  $Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where
    $\varepsilon_i \overset{\mathrm{iid}}{\sim} N(0, \sigma^2)$
2.  Scatterplots + correlation
3.  Fit the regression line in R and look at a `summary()`
4.  Making confidence intervals using `confint`

**Today:** moving from *descriptive* to ___________________________

## Sampling variability

Every new sample of riders gives us a slightly different fitted line.
That variability in our ________________ (not the data) is what lets
us build confidence intervals and hypothesis tests.

```{r}
set.seed(1234)
boots <- bootstraps(bikes, times = 20, apparent = TRUE)
fit_lm_on_bootstrap <- function(split) lm(rides ~ temp_actual, data = analysis(split))
boot_models <- boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))
boot_coefs <- boot_models |>
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

ggplot(boot_coefs) +
  geom_abline(aes(slope = temp_actual, intercept = `(Intercept)`),
              color = "#00aeef", alpha = 0.6, linewidth = 0.5) +
  geom_point(data = bikes, aes(temp_actual, rides), color = "grey60") +
  labs(x = "Temperature", y = "No. rides", title = "20 bootstrap resamples")
```

## Sampling distribution of a coefficient

We use the **t distribution** to model the sampling distribution of a
regression coefficient.

-   Degrees of freedom: $df =$ ___________________
-   Mean: $= \widehat\beta_i$
-   Standard deviation: $SE(\widehat\beta_i) = \dfrac{\widehat\sigma}{\sqrt{\sum (x_i - \overline x)^2}}$

## Confidence interval (general form)

$$\text{statistic} \pm \text{critical value} \times \text{standard error}$$

$$\widehat\beta_i \pm t^* \cdot SE(\widehat\beta_i)$$

::: {.callout-note}
## Critical value

$t^*$ is the value that cuts off the right tail area that matches
your confidence level -- e.g. for a 95% CI, find $t^*$ so that the
two-sided tail area is 0.025 on each side, using `qt()`.
:::

## Hypothesis test (general form)

$$H_0: \beta_i = \text{null value} \quad \text{vs.} \quad H_a: \beta_i \; \begin{cases}\ne\\<\\>\end{cases} \text{null value}$$

$$\text{test statistic} = \frac{\text{estimate} - \text{null value}}{SE} = \frac{\widehat\beta_i - \text{null value}}{SE(\widehat\beta_i)}$$

Then find the appropriate tail area(s) under the t-distribution with
$df = n - 2$.

## Coefficient table

R's `summary()`/`tidy()` output gives us a coefficient table:

```{r}
kable_coef(bikes_lm, terms = c("Intercept", "temp_actual"))
```

::: {.callout-tip}
## Try it

Verify the test statistic for $H_0: \beta_1 = 0$ against
$H_A: \beta_1 \ne 0$ using the table above.
:::

## Example: 90% confidence interval for $\beta_1$

$$\widehat\beta_i \pm t^* \cdot SE(\widehat\beta_i)$$

```{r}
t_star <- qt(0.95, df = nrow(bikes) - 2)
ci90 <- confint(bikes_lm, level = 0.90)
slope_est <- coef(bikes_lm)[["temp_actual"]]
```

$t^* =$ `r round(t_star, 3)` (from `qt(0.95, df = nrow(bikes) - 2)`)

$$`r round(slope_est, 2)` \pm `r round(t_star, 3)` \cdot SE(\widehat\beta_1) \;\Longrightarrow\; (`r round(ci90[2,1], 2)`, \; `r round(ci90[2,2], 2)`)$$

**Interpretation:** We are 90% confident that for every 1°F increase
in temperature, the number of rides changes by between
`r round(ci90[2,1], 1)` and `r round(ci90[2,2], 1)`, on average.

::: {.callout-tip}
## General template (reusable for any CI)

We are ________% confident that for every 1 __________ increase in
______________________________, ___________________________________
also ________________ by between ____________ and ____________.
:::

## Your turn: on paper

::: {.callout-note}
Work the hypothesis test and confidence interval for this example by
hand before checking your answers against the worked example above.
:::

# Next: intro to R and RStudio
