---
title: "Day 10: Adding Categorical Predictors"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(broom)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

bikes <- bikes |> select(date, season, year, month, weekend, holiday,
                          temp_actual, temp_feel, humidity, windspeed, rides)

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt",
                   sep = "\t", header = TRUE) |>
  mutate(Smoke = factor(Smoke, levels = c(0, 1), labels = c("Nonsmoker", "Smoker")))

bikes_slr <- lm(rides ~ temp_actual, data = bikes)
bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)
```

## Recap: `bikes` MLR

Last time we added a second quantitative predictor, `windspeed`, to
our model for `rides`:

$$\widehat{\text{rides}} = `r round(coef(bikes_mlr)[1], 1)` `r round(coef(bikes_mlr)[2], 2)` \cdot \text{temp\_actual} `r round(coef(bikes_mlr)[3], 2)` \cdot \text{windspeed}$$

```{r}
kable_coef(bikes_mlr, terms = c("Intercept", "temp_actual", "windspeed"))
```

$R^2 =$ `r round(summary(bikes_mlr)$r.squared, 3)`

## Today's plan: drop `windspeed`, add `weekend`

```{r}
#| fig-width: 8
#| fig-height: 3
p1 <- gf_point(rides ~ temp_actual, color = ~weekend, data = bikes) +
  theme(legend.position = "none")
p2 <- gf_boxplot(rides ~ weekend, fill = ~weekend, data = bikes) +
  theme(legend.position = "none")

p1 + p2 + plot_layout(widths = c(3, 1))
```

## A new example: lung function and smoking

**Goal:** investigate the association between smoking and lung
capacity using data from 345 adolescents ages 10-19.

**Wrinkle:** lung function naturally increases during adolescence, but
smoking may slow its progression -- so we need to account for *both*
age and smoking status.

| Variable | Description |
|:---|:---|
| `FEV` | forced expiratory volume (liters/second) |
| `Age` | age in years |
| `Smoke` | `Smoker` or `Nonsmoker` |

::: {.callout-note}
Source: Rosner (2006)
:::

## EDA

```{r}
#| fig-width: 5
#| fig-height: 3
ggplot(fev) +
  geom_boxplot(aes(x = Smoke, y = FEV, fill = Smoke), alpha = 0.6) +
  scale_fill_manual(values = c("#f15a31", "#000036")) +
  theme(legend.position = "none") +
  coord_flip() +
  labs(x = "", y = "FEV (L/s)")
```

```{r}
#| fig-width: 6
#| fig-height: 3
ggplot(fev) +
  geom_point(aes(x = Age, y = FEV, color = Smoke, shape = Smoke), alpha = 0.5) +
  facet_wrap(vars(Smoke)) +
  scale_color_manual(values = c("#f15a31", "#000036")) +
  labs(y = "FEV (L/s)")
```

Notice smokers skew *older* -- any raw FEV difference between smokers
and nonsmokers is tangled up with age. This is exactly why we need a
model that controls for both variables at once.

## Indicator variables

Regression requires a **numeric** representation of every variable,
so a categorical predictor gets converted to indicator (0/1) columns
-- e.g. `Smoker` = ___, `Nonsmoker` = ___.

::: {.callout-tip}
## Group task

For each regression model on the handout, sketch the fitted model on
the whiteboard. Each fitted model will have two lines: one for
smokers, one for nonsmokers.
:::

::: callout-important
## Stop here

Let Amanda know when your group finishes!
:::

## Three ways to combine a line + a group

We'll illustrate with `rides ~ temp_actual + weekend` on the bikes
data, since it's easy to see two clear groups.

### Parallel lines model

Same rate of increase in both groups, but a different baseline.

```{r}
#| fig-height: 3
mod_parallel <- lm(rides ~ temp_actual + weekend, data = bikes)
bikes_aug <- augment(mod_parallel, data = bikes)
p1 |> gf_line(.fitted ~ temp_actual, color = ~weekend, data = bikes_aug, linewidth = 1)
```

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{temp} + \beta_2 \texttt{weekend}$$

-   $\beta_0$: intercept for weekdays
-   $\beta_0 + \beta_2$: intercept for weekends
-   $\beta_1$: slope for ___________________________

### Different slopes model

Same baseline, but temperature affects ridership differently on
weekends vs. weekdays.

```{r}
#| fig-height: 3
mod_diff_slope <- lm(rides ~ temp_actual:weekend, data = bikes)
bikes_aug <- augment(mod_diff_slope, data = bikes)
p1 |> gf_line(.fitted ~ temp_actual, color = ~weekend, data = bikes_aug, linewidth = 1)
```

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{temp} + \beta_2 \texttt{temp} \times \texttt{weekend}$$

-   $\beta_0$: intercept for ___________________________
-   $\beta_1$: slope for weekdays
-   $\beta_1 + \beta_2$: slope for weekends

### Separate lines model

Both the baseline *and* the rate of increase differ by group.

```{r}
#| fig-height: 3
mod_separate <- lm(rides ~ temp_actual + weekend + temp_actual * weekend, data = bikes)
bikes_aug <- augment(mod_separate, data = bikes)
p1 |> gf_line(.fitted ~ temp_actual, color = ~weekend, data = bikes_aug, linewidth = 1)
```

$$\mu(y|x) = \beta_0 + \beta_1 \texttt{temp} + \beta_2 \texttt{weekend} + \beta_3 \texttt{temp} \times \texttt{weekend}$$

-   $\beta_0$: intercept for weekdays
-   $\beta_0 + \beta_2$: intercept for weekends
-   $\beta_1$: slope for weekdays
-   $\beta_1 + \beta_3$: slope for ___________________________

::: {.callout-note}
A model with an **interaction term** ($x \times \text{group}$) is
exactly the separate-lines model: it lets both the intercept and
slope differ by group.
:::

# More than two categories

What if we want to incorporate `season` (fall, spring, summer,
winter)?

## A bad idea

Convert the column to a single number:
Fall $\to 1$, Winter $\to 2$, Spring $\to 3$, Summer $\to 4$, then fit
$\mu(\text{rides} \mid \text{season}) = \beta_0 + \beta_1 \, \texttt{season}$.

This is wrong -- it forces the seasons onto an arbitrary numeric scale
and assumes, e.g., that the "jump" from fall to winter is the same
size and direction as the jump from spring to summer.

## A good idea: indicator (dummy) variables

For a categorical variable with $k$ levels, create $k - 1$ indicator
variables (one level is left out as the **reference level**).

```{r}
season_ex <- tibble(
  season = factor(c("winter", "fall", "spring", "spring", "fall", "summer"),
                   levels = c("fall", "winter", "spring", "summer"))
)
cbind(season = as.character(season_ex$season),
      model.matrix(~season, data = season_ex)[, -1]) |>
  knitr::kable()
```

Here `fall` is the reference level -- it's represented by all zeros --
and each other season gets its own 0/1 column.

::: {.callout-tip}
## Group task

For each sketched regression model on the handout, write the mean
function that matches it (line type represents `season`).
:::

::: callout-important
## Stop here

Let Amanda know when your group finishes!
:::
