---
title: "Day 11: Inference for MLR"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(broom)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

bikes <- bikes |> select(date, season, year, month, weekend, holiday,
                          temp_actual, temp_feel, humidity, windspeed, rides)

bikes_slr <- lm(rides ~ temp_actual, data = bikes)
bikes_mlr <- lm(rides ~ temp_actual * weekend, data = bikes)
bikes_mlr2 <- lm(rides ~ temp_actual + weekend, data = bikes)
```

## `bikes` data

**Goal:** Predict ______________ based on `temp_actual` and `weekend`


```{r}
kable_coef(bikes_mlr, terms = c("Intercept", "temp_actual", "weekendTRUE",
                                 "temp_actual:weekendTRUE"))
```

**Warm up:**

Write out the theoretical model from the table above

\vspace{1in}

Write out the fitted model for weekends and weekdays separately

\vspace{2in}


## Do we *need* a different-slopes model?

```{r}
#| fig-height: 2
#| fig-width: 4 
#| fig-align: left
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes, size = .5) |> gf_lm()
```

**Answer based on the graph:**

\vspace{.5in}

**Answer based on the table:**

\vspace{.5in}

## T-test for a single coefficient

**Hypotheses:** $H_0: \beta_j = \#$ vs. $H_a: \beta_j \ne, <, \text{or} > \#$

**Test statistic:** $t = \dfrac{\widehat\beta_j - \#}{SE(\widehat\beta_j)}$

**Reference distribution:** $t$ distribution with $df = n - (k+1)$,
where $k$ = number of predictors in the model

**p-value:** area in the tail(s) specified by $H_a$

```{r}
#| fig-height: 2
#| fig-width: 7.5
p1 <- ggplot(data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = 0.4) +
  ggtitle(expression(H[a]: beta[j] < 0)) +
  theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())

p2 <- ggplot(data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3.5, -2), geom = "area", alpha = 0.4) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = 0.4) +
  ggtitle(expression(H[a]: beta[j] != 0)) +
  theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())

p3 <- ggplot(data.frame(x = c(-3.5, 3.5)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 5)) +
  stat_function(fun = dt, args = list(df = 5), xlim = c(2, 3.5), geom = "area", alpha = 0.4) +
  ggtitle(expression(H[a]: beta[j] > 0)) +
  theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())

p1 + p2 + p3
```

**CI for a single coefficient:**

$\widehat\beta_j \pm t^*_{n-(k+1)} \cdot SE(\widehat\beta_j)$

\vspace{1in}

::: {.callout-caution}
## Caution!

$\widehat \beta$'s and p-values are highly ___________________ on each other! ___________________ $\beta$'s might become ___________________ after removing or adding another term
:::

Once we drop the interaction between `weekend` and `temp_actual`, the simpler
parallel-lines model shows `weekend` is clearly still useful:

```{r}
kable_coef(bikes_mlr2, terms = c("Intercept", "temp_actual", "weekendTRUE"))
```

## Explained variability

```{r}
#| fig-height: 2.5
#| fig-width: 7
slr_aug <- augment(bikes_slr) |> mutate(mean.y = mean(rides))

p1 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Total") |>
  gf_segment(mean.y + rides ~ temp_actual + temp_actual, color = "gray70", alpha = 0.4) |>
  gf_hline(yintercept = mean(slr_aug$rides), color = "#f15a31")

p2 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Error") |>
  gf_segment(.fitted + rides ~ temp_actual + temp_actual, color = "#003069", alpha = 0.4) |>
  gf_lm(color = "#f15a31")

p1 + p2
```

**Idea:** does a larger model explain a *significantly* bigger share
of the total variability than a smaller model?

## Can we just compare $R^2$ values?

```{r}
bikes_mlr_3var <- lm(rides ~ temp_actual + windspeed + weekend, data = bikes)
bikes_mlr_4var <- lm(rides ~ temp_actual + windspeed * weekend, data = bikes)

tibble(
  Model = c("temp only", "temp + weekend", "temp + windspeed + weekend",
            "temp + windspeed*weekend"),
  R2 = c(summary(bikes_slr)$r.squared, summary(bikes_mlr2)$r.squared,
         summary(bikes_mlr_3var)$r.squared, summary(bikes_mlr_4var)$r.squared),
  Adj_R2 = c(summary(bikes_slr)$adj.r.squared, summary(bikes_mlr2)$adj.r.squared,
             summary(bikes_mlr_3var)$adj.r.squared, summary(bikes_mlr_4var)$adj.r.squared)
) |>
  knitr::kable(digits = 4, col.names = c("Model", "$R^2$", "Adjusted $R^2$"))
```

$$R^2 = 1 - \frac{SSE}{SST} \hspace{2cm} R^2_{adj} = 1 - \frac{SSE/(n-k)}{SST/(n-1)} = 1 - \frac{MSE}{MST}$$

-   $R^2$ **always** increases when predictors are added, even useless
    ones
-   $R^2_{adj}$ **could** ___________________________ if unnecessary
    predictors are added
-   Better to use $R^2_{adj}$ for comparing multiple regression models

## Formal test for comparing models

| | Model |
|:---|:---|
| Full | $\mu\{Y|X\} = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2$ |
| Reduced | $\mu\{Y|X\} = \beta_0 + \beta_1 x_1$ |

**Hypotheses:** $H_0: \beta_2 = \beta_3 = 0$ vs. $H_a:$ at least one
$\beta_j \ne 0$, $j = 2,3$

## Comparing models with an F-test

$$F = \frac{(R^2_{\text{full}} - R^2_{\text{reduced}})/d}{(1 - R^2_{\text{full}})/df_{\text{full}}} = \frac{(SSR_{\text{full}} - SSR_{\text{reduced}})/d}{MSE_{\text{full}}} = \frac{(SSE_{\text{reduced}} - SSE_{\text{full}})/d}{MSE_{\text{full}}}$$

where

-   $d = df_{\text{full}} - df_{\text{reduced}} =$ \# of $\beta$'s
    being tested
-   $df_i = n - (p+1) =$ error d.f. for model $i$
-   $MSE_{\text{full}} = \dfrac{SSE_{\text{full}}}{df_{\text{full}}}$

The $F$ statistic follows an $F$ distribution with $d$ and $n-p-1$
degrees of freedom. Like the $\chi^2$, it's right-skewed and only
takes positive values, so we only ever look at the **upper tail** for
a p-value:

```{r}
#| fig-height: 2.5
#| fig-width: 5
#| fig-align: center
ggplot(data.frame(x = c(0.00001, 5.5)), aes(x)) +
  stat_function(fun = df, args = list(df1 = 4, df2 = 23), color = "#003300", linewidth = 1) +
  stat_function(fun = df, args = list(df1 = 4, df2 = 23), xlim = c(3, 5.5),
                geom = "area", fill = "#003300", alpha = 0.4) +
  labs(x = "F", y = "Density") +
  annotate("text", x = 4.6, y = 0.15, label = "p-value", color = "#003300", size = 5)
```

Upper-tail area in R: `1 - pf(stat, df1, df2)`

## Putting it all together

```{r}
full_table <- glance(bikes_mlr) |> select(r.squared, df, df.residual, nobs) |>
  mutate(model = "Full: temp*weekend")
reduced_table <- glance(bikes_slr) |> select(r.squared, df, df.residual, nobs) |>
  mutate(model = "Reduced: temp only")

bind_rows(full_table, reduced_table) |>
  select(model, everything()) |>
  knitr::kable(digits = 3)
```

```{r}
d <- 2
mse_full <- summary(bikes_mlr)$sigma^2
f_stat <- ((summary(bikes_mlr)$r.squared - summary(bikes_slr)$r.squared) / d) /
  ((1 - summary(bikes_mlr)$r.squared) / bikes_mlr$df.residual)
p_val <- 1 - pf(f_stat, d, bikes_mlr$df.residual)
```

$$F = \frac{(R^2_{\text{full}} - R^2_{\text{reduced}})/d}{(1 - R^2_{\text{full}})/df_{\text{full}}} = \frac{(`r round(summary(bikes_mlr)$r.squared,3)` - `r round(summary(bikes_slr)$r.squared,3)`)/2}{(1 - `r round(summary(bikes_mlr)$r.squared,3)`)/`r bikes_mlr$df.residual`} \approx `r round(f_stat,1)`$$

p-value = `r format(p_val, scientific = TRUE, digits = 3)` -- extremely
small, so we ___________________________ $H_0$: adding `weekend`
(and its interaction with `temp_actual`) meaningfully improves the
model.

## Doing it directly in R with `anova()`

`anova(reduced, full)` takes the **reduced** model first, then the
**full** model, and runs exactly the F-test above -- no need to
compute $R^2$'s by hand.

```{r}
full <- lm(rides ~ temp_actual * weekend, data = bikes)
reduced <- lm(rides ~ temp_actual, data = bikes)
anova(reduced, full)
```
