---
title: "Day 11: Inference for MLR"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(broom)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

bikes <- bikes |> select(date, season, year, month, weekend, holiday,
                          temp_actual, temp_feel, humidity, windspeed, rides)

bikes_slr <- lm(rides ~ temp_actual, data = bikes)
bikes_mlr <- lm(rides ~ temp_actual * weekend, data = bikes)
bikes_mlr2 <- lm(rides ~ temp_actual + weekend, data = bikes)
```

## `bikes` data

**Goal:** Predict ______________ based on `temp_actual` and `weekend`


```{r}
kable_coef(bikes_mlr, terms = c("Intercept", "temp_actual", "weekendTRUE",
                                 "temp_actual:weekendTRUE"))
```

**Warm up:**

Write out the theoretical model from the table above

\vspace{1in}

Write out the fitted model for weekends and weekdays separately

\vspace{2in}


## Do we *need* a different-slopes model?

```{r}
#| fig-height: 2
#| fig-width: 4 
#| fig-align: left
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes, size = .5) |> gf_lm()
```

**Answer based on the graph:**

\vspace{.5in}

**Answer based on the table:**

\vspace{.5in}

## T-test for a single coefficient

**Hypotheses:** $H_0: \beta_j = \#$ vs. $H_a: \beta_j \ne, <, \text{or} > \#$

**Test statistic:** $t = \dfrac{\widehat\beta_j - \#}{SE(\widehat\beta_j)}$

**Reference distribution:** $t$ distribution with $df = n - (p+1)$,
where $p$ = number of predictors in the model (an interaction counts as
one)

**p-value:** area in the tail(s) specified by $H_a$

**CI for a single coefficient:**

$\widehat\beta_j \pm t^*_{n-(p+1)} \cdot SE(\widehat\beta_j)$

**Example:** Use the coefficient table at the top of the notes and `qt(0.975, df = 496)` to calculate a 95% confidence interval for the slope of `temp_actual` (the slope for weekdays).

\vspace{1in}

Do we need `weekend` at all? 

**Answer:** 

\vspace{.75in}

Now drop only the interaction between `weekend` and `temp_actual` (the
parallel-lines model):

```{r}
kable_coef(bikes_mlr2, terms = c("Intercept", "temp_actual", "weekendTRUE"))
```

What changed about `weekend`?

\vspace{.75in}

::: {.callout-caution}
## Caution!

$\widehat \beta$'s and p-values are highly ___________________ on each other! ___________________ $\beta$'s might become ___________________ after removing or adding another term
:::

A t-test can only ask about **one** coefficient at a time, given everything else in the model. We need a test that asks about several coefficients at once

## Explained variability

**Idea:** does the **full** model reduce the error (SSE) by *more than chance alone would*?

```{r}
#| include: false
bikes_noise <- lm(rides ~ temp_actual + n1 + n2 + n3 + n4 + n5,
                  data = {
                    set.seed(230)
                    bikes |> mutate(n1 = rnorm(n()), n2 = rnorm(n()), n3 = rnorm(n()),
                                    n4 = rnorm(n()), n5 = rnorm(n()))
                  })
bikes_season <- lm(rides ~ temp_actual + season, data = bikes)
bikes_season_w <- lm(rides ~ temp_actual + season_w,
                     data = bikes |> mutate(season_w = relevel(season, ref = "winter")))

a_full <- anova(bikes_slr, bikes_mlr)
a_slopes <- anova(bikes_mlr2, bikes_mlr)
a_shift <- anova(bikes_slr, bikes_mlr2)
d_full <- a_full$Df[2]
f_full <- a_full$F[2]
f_full_p <- a_full$`Pr(>F)`[2]
sci <- function(p) format(p, digits = 2, scientific = TRUE)
```

```{r}
#| fig-height: 2.5
#| fig-width: 7
slr_aug <- augment(bikes_slr) |> mutate(mean.y = mean(rides))

p1 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Total", size = .1) |>
  gf_segment(mean.y + rides ~ temp_actual + temp_actual, color = "gray70", alpha = 0.4) |>
  gf_hline(yintercept = mean(slr_aug$rides), color = "#f15a31")

p2 <- gf_point(rides ~ temp_actual, data = slr_aug, title = "SS Error", size = .1) |>
  gf_segment(.fitted + rides ~ temp_actual + temp_actual, color = "#003069", alpha = 0.4) |>
  gf_lm(color = "#f15a31")

p1 + p2
```


## Comparing nested models: the Extra Sum of Squares F Test

Full Model: $\mu\{Y|X\} = \beta_0 + \beta_1 \text{temp} + \beta_2 \text{weekend} + \beta_3 \text{temp} \times \text{weekend}$ 

Reduced Model: $\mu\{Y|X\} = \beta_0 + \beta_1 \text{temp}$ 

The reduced model is the full model with some $\beta$'s set to 0, so we say the models are **nested**.

- $H_0$: ___________________ 
- $H_a:$ ___________________

$$F = \frac{(\text{SSE}_{\text{reduced}} - \text{SSE}_{\text{full}}) / d}{\text{MSE}_{\text{full}}} = \frac{\text{how much error drops per } \beta}{\text{estimate of } \sigma^2}$$

-   If $H_0$ is true, both parts estimate $\sigma^2$, so
    $F \approx$ ______
-   If the extra terms really help, $F$ is much ___________________ than 1

where 

  - $d = \text{df}_{\text{reduced}} - \text{df}_{\text{full}}$ is the number of $\beta$'s being tested
  - $\text{df}_i = n - (p_i + 1)$ is the error d.f. for model $i$
  - $\text{MSE}_{\text{full}} = \text{SSE}_{\text{full}} / \text{df}_{\text{full}}$.

The $F$ statistic follows an $F$ distribution with $d$ and
$\text{df}_{\text{full}}$ degrees of freedom. It is right-skewed, and only large values of $F$ are evidence against $H_0$, so the p-value is the
**upper tail**:

```{r}
#| fig-height: 2
#| fig-width: 3
#| fig-align: left
ggplot(data.frame(x = c(0.00001, 5.5)), aes(x)) +
  stat_function(fun = df, args = list(df1 = 4, df2 = 23), color = "#003300", linewidth = 1) +
  stat_function(fun = df, args = list(df1 = 4, df2 = 23), xlim = c(3, 5.5),
                geom = "area", fill = "#003300", alpha = 0.4) +
  labs(x = "F", y = "Density") +
  annotate("text", x = 4.6, y = 0.15, label = "p-value", color = "#003300", size = 5)
```

Upper-tail area in R: `pf(stat, df1, df2, lower.tail = FALSE)`



```{r}
tibble(
  Model = c("Reduced: temp", "Full: temp * weekend"),
  SSE = c(deviance(bikes_slr), deviance(bikes_mlr)),
  df = c(df.residual(bikes_slr), df.residual(bikes_mlr))
) |>
  knitr::kable(format.args = list(big.mark = ",", scientific = FALSE), digits = 0)
```

**Example:** Calculate the $F$ statistic.

\vspace{1.25in}


```{r}
f_stat <- 15.26
```

```{r}
#| echo: true
#| collapse: true
pf(f_stat, df1 = 2, df2 = 496, lower.tail = FALSE)
```


&rarr; We ___________________________ $H_0$ and conclude: 

## Extra Sum of Squares F test in R

```{r}
#| echo: true
full <- lm(rides ~ temp_actual * weekend, data = bikes)
reduced <- lm(rides ~ temp_actual, data = bikes)
anova(reduced, full)
```

```
Analysis of Variance Table

Model 1: rides ~ temp_actual
Model 2: rides ~ temp_actual * weekend
  Res.Df       RSS Df    Sum of Sq      F               Pr(>F)    
1    df1      SSE1                                  
2    df2      SSE2  d    SSE1-SSE2  (SumSq/d)/(SSE2/df2) p-val
```

## Resolving the puzzle

Earlier, the t-tests in the full model said neither `weekend` term was
useful, but `weekend` was highly significant once the interaction was
dropped. The F-test lets us ask each question directly:

```{r}
tibble(
  Question = c("Different slopes needed?", "Weekend shift needed (no interaction)?",
               "Weekend needed at all?"),
  Reduced = c("temp + weekend", "temp", "temp"),
  Full = c("temp * weekend", "temp + weekend", "temp * weekend"),
  d = c(a_slopes$Df[2], a_shift$Df[2], a_full$Df[2]),
  F = as.character(signif(c(a_slopes$F[2], a_shift$F[2], a_full$F[2]), 3)),
  `p-value` = fmt_p(c(a_slopes$`Pr(>F)`[2], a_shift$`Pr(>F)`[2], a_full$`Pr(>F)`[2]))
) |>
  knitr::kable()
```

-   One tool answers all three questions
-   When $d = 1$, the F-test and the t-test are the same test:
    $F = $ ______
-   The t-tests couldn't answer the *joint* question (row __)

## `anova()` on a single model

```{r}
#| echo: true
anova(bikes_mlr)
```

-   Each row adds one term to the model, in order, and tests whether it
    improves the fit ___________________________ the terms above it
-   The **last row** is the extra sums of squares F-test for the last term
    (here, the interaction) and matches
    `anova(bikes_mlr2, bikes_mlr)`


## Reading R output

```
lm(formula = rides ~ temp_actual * weekend, data = bikes)

Coefficients:
                         Estimate Std. Error t value Pr(>|t|)    
(Intercept)             -1843.867    415.700  -4.436 1.13e-05 ***
temp_actual                87.173      6.439  13.538  < 2e-16 ***
weekendTRUE              -714.938    795.698  -0.899    0.369    
temp_actual:weekendTRUE     0.411     12.539   0.033    0.974    

Residual standard error: 1253 on 496 degrees of freedom
Multiple R-squared:  0.3709,	Adjusted R-squared:  0.3671 
F-statistic: 97.47 on 3 and 496 DF,  p-value: < 2.2e-16
```

The **F-statistic** at the bottom is the same kind of test: the full
model vs. an *intercept-only* model.

$H_0$: ___________________________________ (none of the predictors
help)



## Adjusted $R^2$


$$R^2 = 1 - \frac{SSE}{SST} \hspace{2cm} R^2_{adj} = 1 - \frac{SSE/(n-p-1)}{SST/(n-1)} = 1 - \frac{MSE}{MST}$$

where $p$ is the number of predictors in the model.

**Example:** Look at the adjusted $R^2$ column. Which model has the
largest adjusted $R^2$? Did adding the interaction help?

\vspace{.75in}

```{r}
tibble(
  Model = c("temp", "temp + weekend", "temp * weekend", "temp + 5 random noise variables"),
  R2 = c(summary(bikes_slr)$r.squared, summary(bikes_mlr2)$r.squared,
         summary(bikes_mlr)$r.squared, summary(bikes_noise)$r.squared),
  Adj_R2 = c(summary(bikes_slr)$adj.r.squared, summary(bikes_mlr2)$adj.r.squared,
             summary(bikes_mlr)$adj.r.squared, summary(bikes_noise)$adj.r.squared)
) |>
  knitr::kable(digits = 4, col.names = c("Model", "$R^2$", "Adjusted $R^2$"))
```

-   $R^2$ **never** goes ___________________ when predictors are added,
    even for pure ___________________
-   $R^2_{adj}$ **could** ___________________________ if unnecessary
    predictors are added
-   Better to use $R^2_{adj}$ than $R^2$ to *compare* models, but neither
    is a formal test

## Takeaways

-   A t-test asks about **one** coefficient, *given* everything else in
    the model
-   To test **several** coefficients at once, compare nested models with
    an **extra sums of squares F-test**
-   $F = \dfrac{(\text{SSE}_{\text{reduced}} - \text{SSE}_{\text{full}}) / d}{\text{MSE}_{\text{full}}}$:
    error reduction per extra $\beta$, relative to the noise
-   $R^2$ always goes up; the F-test tells us whether the improvement is
    *real*
