---
title: "Day 04: Inference for Prediction"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(ggdist)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

nhanes <- read_csv(here::here("data/nhanes-sitting-depression.csv")) |>
  mutate(dep_score = log(phq9 + 1))

bikes_lm <- lm(rides ~ temp_actual, data = bikes)

nhanes_lm <- lm(dep_score ~ sedentary_hours, data = nhanes)

```

## In groups: warm-up questions 

Is education level associated with income? Researchers collected education level (in years) and income (in thousands of dollars) for a random sample of 32 employees working for the city of Riverview.

```{r}
#| include: false
city <- readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/riverview.csv")
lm.a = lm(income ~ 1 + education, data = city)
```

The researchers fit a simple linear regression of income on education and obtained the following output:

```{r}
#| echo: false
#| tbl-align: center
broom::tidy(lm.a) |>
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
knitr::kable(format = "pipe", digits = 3, booktabs = TRUE)
```

**Q1.** Write the fitted regression equation.

\vspace{1in}

**Q2.** Interpret the slope coefficient in the context of the problem. (Don't forget to specify units.)

\vspace{1in}

**Q3.** Interpret the intercept in the context of the problem. (Don't forget to specify units.)

\vspace{1in}

A new researcher joined the team and decided that education should be standardized (to have mean 0 and SD 1) before fitting the regression model. The output from this regression is shown below:

```{r}
#| echo: false
lm.b <- lm(income ~ 1 + scale(education), data = city)
broom::tidy(lm.b) |>
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
knitr::kable(format = "pipe", digits = 3, booktabs = TRUE)
```

**Q4.** Write the fitted regression equation for this new model.

\vspace{1in}

**Q5.** Interpret the slope coefficient in the context of the problem. (Don't forget to specify units.)

\vspace{1in}

**Q6.** Interpret the intercept in the context of the problem. (Don't forget to specify units.)

\vspace{1in}

**Q7.** Compare the two models. What's the same? What's different?

\vspace{1in}

::: callout-important
## Stop here

Let Amanda know when your group finishes! 
:::

## Two types of predictions

1. Predicting the **mean response** at a specific value of $x$ (called a _________________________ interval)
    - e.g., the average leaf width in 2010

2. Predicting the response for a **specific future observation** (called a _________________________ interval)
    - e.g., the leaf width of *one specific leaf* in 2010


**Other Examples:** 

\vspace{1in}


## Inference for prediction

Best point estimate: $\widehat y_0 = \widehat\beta_0 + \widehat\beta_1 x_0$

General interval formula:
$$\text{estimate} \pm t^* \times SE$$

- $\widehat y_0$ is our estimate
- use a $t$-distribution with $df = n-2$ to find $t^*$

We'll need a *different* SE depending on whether we're building:

- a confidence interval for the mean response $\widehat\mu(Y \mid X_0)$, or
- a prediction interval for an individual response $Y \mid X_0$

## Standard errors

$$SE\big(\widehat\mu(Y\mid X)\big) = \widehat\sigma \sqrt{\frac{1}{n} + \frac{(x_0-\overline x)^2}{\sum_{i=1}^n (x_i-\overline x)^2}}$$

$$SE(\widehat y) = \widehat\sigma \sqrt{1 + \frac{1}{n} + \frac{(x_0-\overline x)^2}{\sum_{i=1}^n (x_i-\overline x)^2}}$$

- The interval for _________ will be wider

- As $x_0$ gets farther from $\overline{x}$, both SE's get _______________

$$SE(\widehat y) = \sqrt{\underbrace{\widehat\sigma^2}_{\text{variability in } \rule{.5cm}{0.3pt}} + \underbrace{\frac{\widehat\sigma^2}{n} + \frac{\widehat\sigma^2(x_0-\overline x)^2}{\sum_{i=1}^n (x_i-\overline x)^2}}_{\text{variability in } \rule{1cm}{0.3pt}}}$$


```{r}
#| fig-width: 8
#| fig-height: 3
set.seed(1234)
boots <- bootstraps(bikes, times = 20, apparent = TRUE)
fit_lm_on_bootstrap <- function(split) lm(rides ~ temp_actual, data = analysis(split))
boot_models <- boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap), coef_info = map(model, tidy))
boot_coefs <- boot_models |>
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

bike_dsn <- tibble(
  temp_actual = c(50, 60, 70, 80),
  yhat = predict(bikes_lm, data.frame(temp_actual = c(50, 60, 70, 80))),
  resid_sd = sigma(bikes_lm)
)

p_beta <- ggplot(boot_coefs) +
  geom_abline(aes(slope = temp_actual, intercept = `(Intercept)`),
              color = "#0072B2", linewidth = 0.5, alpha = 0.4) +
  geom_point(data = bikes, aes(temp_actual, rides), color = "grey60", size = .25) +
  labs(x = "Temperature", y = "No. rides", title = "Variability in the fitted line")

p_eps <- ggplot(data = bikes, aes(x = temp_actual, y = rides)) +
  #geom_point(alpha = 0.4, size = 1.2, color = "#003300") +
  geom_lm(color = "#f15a31") +
  stat_slab(aes(x = temp_actual, ydist = "norm", arg1 = yhat, arg2 = resid_sd),
            data = bike_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  labs(x = "Temperature", y = "No. rides", title = "Variability around the line")

p_beta + p_eps
```

## R commands

`predict()` quickly calculates $\widehat y$ for a given $x$ (or a
vector of $x$'s):

```{r}
#| echo: true
#| collapse: true
predict(nhanes_lm, newdata = data.frame(sedentary_hours = 10))
predict(nhanes_lm, newdata = data.frame(sedentary_hours = c(8, 10, 12)))
```

The `interval` argument specifies which type of interval you want:

```{r}
#| echo: true
#| collapse: true
predict(nhanes_lm, newdata = data.frame(sedentary_hours = 10),
 interval = "confidence")
predict(nhanes_lm, newdata = data.frame(sedentary_hours = 10),
 interval = "prediction")
```

The `level` argument sets the confidence level (default 0.95):

```{r}
#| echo: true
#| collapse: true
predict(nhanes_lm, newdata = data.frame(sedentary_hours = 10), 
interval = "prediction", level = .99)
```

Adding `se.fit = TRUE` returns the standard error needed for "by hand" calculations:

```{r}
#| echo: true
#| collapse: true
predict(nhanes_lm, newdata = data.frame(sedentary_hours = 10), 
se.fit = TRUE)
```

## In groups: prediction and confidence intervals

::: {.callout-tip}
The R code for the following questions is found at <br>

<https://aluby.github.io/stat230/activities/04R-prediction>

The URL is also posted on Moodle.
:::

Now, we'll explore the `city` data using R.


**Q1.** Make a plot of `income` against `education`  and overlay the linear regression equation. Does it look like what you expected?

\vspace{1in}

**Q2.** Make a plot of `income` against `scale(education)` and overlay the linear regression equation. Does it look like what you expected?

\vspace{1in}

**Q3.** Find the linear regression equation from Q1. Confirm that it matches the table at the beginning of this worksheet.

**Q4.** The first person in this dataset has `education = 8`. Calculate the expected income of this person using the first regression equation.

\vspace{1in}

**Q5.** If we want to predict the income of **this** person, should we use a confidence interval or a prediction interval?

\vspace{1in}

**Q6.** Use R to construct the appropriate 89% interval for this person's income. Record this interval below.

\vspace{1in}

**Q7.** Interpret the interval in context.

\vspace{1in}

**Q8.** Run the code to produce a scatterplot, regression line, and both types of intervals. Which is the prediction interval and which is the confidence interval? How can you tell?

\vspace{1in}

