---
title: "Day 05: Evaluating SLR models"
author: "Name: ________________________________"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(broom)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

set.seed(040726)
bike_sample <- bikes |> slice_sample(n = 50)
bike_sm_lm <- lm(rides ~ temp_actual, data = bike_sample)
```

## Data for today

We return to the Capital Bikeshare data, this time working with a
small random sample of 50 days so that patterns are easier to see by
eye.

```{r}
gf_point(rides ~ temp_actual, data = bike_sample, xlab = "Temperature (F)",
         ylab = "No. rides", color = "#003300") |>
  gf_lm(color = "#f15a31")
```

```{r}
kable_coef(bike_sm_lm, terms = c("Intercept", "temp_actual"))
```

## Two definitions of a "good" model

1.  The model is appropriate, so our inferences (p-values, CIs) are
    ___________________________
2.  The x variable does a good job ___________________________ the y
    variable

Theoretical model: $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where
$\varepsilon_i \overset{\mathrm{iid}}{\sim} N(0, \sigma^2)$

Fitted model: $\widehat y_i = \widehat\beta_0 + \widehat\beta_1 x_i$,
so that $y_i = \widehat y_i + \widehat\varepsilon_i$

## Conditions for inference: LINE

For our p-values and confidence intervals to be trustworthy, the
theoretical model above has to be a reasonable description of the
data. Check:

-   **L**inear: ___________________________
-   **I**ndependent and identically distributed errors:
    ___________________________
-   **N**ormally distributed errors: ___________________________
-   **E**qual variance of the errors: ___________________________

::: {.callout-note}
These conditions are about the *errors*, not about the raw x or y
variables themselves.
:::

## Residuals

**Definition:** $e_i = \widehat\varepsilon_i = y_i - \widehat y_i$

**Properties:**

-   sum to zero $\Longrightarrow$ mean is ___
-   **uncorrelated** with $x$ and $\widehat y$
-   (should be) normally distributed
-   $SD(e_i) = \widehat\sigma \sqrt{1 - \dfrac1n - \dfrac{(x_i - \overline x)^2}{\sum (x_i - \overline x)^2}}$

## Residual plots

For SLR, plotting residuals against $x$ or against the fitted values
$\widehat y$ shows the same shape (since $\widehat y$ is a linear
function of $x$) -- either is a valid way to check linearity and
equal variance.

```{r}
#| fig-width: 8
#| fig-height: 3
p1 <- ggplot(augment(bike_sm_lm), aes(x = temp_actual, y = .resid)) +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linewidth = 1) +
  labs(x = "Temperature", y = "Residual")

p2 <- ggplot(augment(bike_sm_lm), aes(x = .fitted, y = .resid)) +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linewidth = 1) +
  labs(x = "Fitted values", y = "Residual")

p1 + p2
```

**A "good" residual plot** shows no pattern: residuals bounce randomly
around zero with roughly constant spread across the range of fitted
values.

```{r}
#| fig-height: 3
#| fig-width: 4.5
set.seed(72384)
x <- rnorm(100, mean = 10, sd = 3.5)
y <- 1 + 1 * x + rnorm(100, mean = 0, sd = 10)
good_mod <- lm(y ~ x)

gf_point(residuals(good_mod) ~ fitted(good_mod),
         ylab = "Residual", xlab = "Fitted values") |>
  gf_hline(yintercept = 0, color = "#f15a31", linetype = 2)
```

::: {.callout-tip}
## Group task

On the whiteboards, sketch a plot of $y$ vs. $x$ *and* the
corresponding residual plot that would indicate a violation of

1.  the linearity condition
2.  the equal variance condition
:::

## Assessing normality

Use a histogram or normal Q-Q plot of the residuals. A "good" Q-Q plot
hugs the diagonal reference line closely.

```{r}
#| fig-width: 8
#| fig-height: 3
p1 <- gf_histogram(~ .resid, data = augment(bike_sm_lm), color = "white",
                    bins = 10, xlab = "Residual", ylab = "Count")

p2 <- gf_qq(~ .resid, data = augment(bike_sm_lm),
            ylab = "Residual", xlab = "N(0,1) quantiles") |>
  gf_qqline()

p1 + p2
```

::: {.callout-note}
Normality is **harder to judge at small sample sizes** -- even true
random draws from a normal distribution can look lumpy or have a
wandering Q-Q plot when $n$ is small. Don't over-interpret minor
wiggles with only 30-50 points.
:::

## Assessing independence

Plot residuals against whatever variable might induce dependence
(time, location, subject, batch, ...). Using the *full* bikes dataset
ordered by date reveals a clear pattern -- a strong hint that
day-to-day ridership is **not** independent once we condition only on
temperature:

```{r}
#| fig-height: 3
bikes_full_lm <- lm(rides ~ temp_actual, data = bikes)

bikes |>
  mutate(.resid = residuals(bikes_full_lm)) |>
  ggplot(aes(x = date, y = .resid)) +
  geom_line(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linetype = 2) +
  labs(x = "Date", y = "Residual")
```

## What happens if the conditions aren't met?

-   **Linearity:** if nonlinear, everything breaks!
-   **Independence:** estimates are still unbiased, but the SEs are
    typically too **small** (we have less information than $n$
    independent points suggests)
-   **Normality:** estimates and SEs are still fine, but **prediction
    intervals** (which rely on individual points being normal around
    the mean) are wrong
-   **Equal variance:** estimates are still unbiased, but SEs are
    wrong -- and we don't know in which direction

## What do we do if assumptions are violated?

1.  Change our assumptions (hard -- needs more advanced methods)
2.  Transform $y$, $x$, or both
3.  Change the type of inference (e.g. bootstrap)

# Is x a "good" predictor of y?

## Partitioning variability

```{r}
#| fig-width: 8
#| fig-height: 3
slr_aug <- augment(bike_sm_lm) |> mutate(mean.y = mean(rides))

p1 <- ggplot(slr_aug, aes(x = temp_actual, y = rides)) +
  geom_point(color = "#003300") +
  geom_segment(aes(xend = temp_actual, yend = mean.y), color = "gray70", alpha = 0.6) +
  geom_hline(yintercept = mean(slr_aug$rides), color = "#f15a31") +
  labs(title = "SS Total")

p2 <- ggplot(slr_aug, aes(x = temp_actual, y = rides)) +
  geom_point(color = "#003300") +
  geom_segment(aes(xend = temp_actual, yend = .fitted), color = "#003069", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "#f15a31") +
  labs(title = "SS Error")

p1 + p2
```

$$\underbrace{\sum (y_i - \overline y)^2}_{SSTO} = \underbrace{\sum (\widehat y_i - \overline y)^2}_{SSR} + \underbrace{\sum e_i^2}_{SSE}$$

$$\text{total variability} = \text{variability of } \widehat Y\text{'s} \; + \; \text{variability of residuals (}\underline{\hspace{2cm}}\text{)}$$

$$R^2 = \frac{SSR}{SSTO} = \frac{\text{variability of } \widehat Y}{\text{variability of } Y}$$

::: {.callout-note}
## $R^2$

$R^2$ gives the proportion of total variability in $Y$ explained by
the SLR model.
:::

For our small bike sample, $R^2 =$ `r round(summary(bike_sm_lm)$r.squared, 3)`
-- about `r round(summary(bike_sm_lm)$r.squared, 3) * 100`% of the
variability in rides is explained by temperature.

## Properties of $R^2$

-   Between ___ and ___
-   Big when $\sigma$ is small relative to the overall variability of
    $y$ (i.e. a big proportion of variability explained by the model)
-   For SLR, $R^2 =$ ___ (where $r$ is the correlation coefficient
    between $x$ and $y$)

::: {.callout-note}
## Quick notation reference

| Symbol | Meaning |
|:---|:---|
| $r$ | sample correlation coefficient |
| $\rho$ | population correlation coefficient |
| $R^2$ | coefficient of determination |
| $p$ | a population proportion |
| $\widehat p$ | a sample proportion |
| p-value | $P(\text{test statistic} \mid H_0)$ |
| $P(A)$ | probability of event $A$ |
:::

## $R^2$ dangers: high $R^2$ doesn't mean "good fit"

```{r}
#| fig-width: 8
#| fig-height: 3.2
set.seed(040726)
fake_data <- tibble(
  x = c(runif(50, 0, 200), 400),
  y = c(.5 * x[1:50]^2 + rnorm(50, 0, 50), 30000)
)
fake_lm <- lm(y ~ x, data = fake_data)

p1 <- ggplot(fake_data, aes(x = x, y = y)) +
  geom_point(color = "#003300") +
  geom_smooth(method = "lm", se = FALSE, color = "#f15a31") +
  labs(title = paste0("R² = ", round(summary(fake_lm)$r.squared, 3)))

p2 <- ggplot(augment(fake_lm), aes(x = .fitted, y = .resid)) +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linetype = 2) +
  labs(x = "Fitted values", y = "Residual")

p1 + p2
```

A single high-leverage point drove $R^2$ up to
`r round(summary(fake_lm)$r.squared, 3)` even though the residual plot
shows an obvious curve -- the linear model is a bad fit! **Always check
the residual plot, not just $R^2$.**

::: {.callout-tip}
## Your turn

Would you trust this model just based on its $R^2$? What does the
residual plot tell you that $R^2$ alone can't?
:::
