---
title: "Day 02: Simple Linear Regression (SLR)"
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
library(rsample)
library(broom)
library(ggdist)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

nhanes <- read_csv(here::here("data/nhanes-sitting-depression.csv")) 

median_hrs <- median(nhanes$sedentary_hours)

nhanes <- nhanes |>
  mutate(dep_score = log(phq9 + 1),
          sitting_group = if_else(sedentary_hours > median_hrs, "High", "Low"))


nhanes_lm <- lm(dep_score ~ sedentary_hours, data = nhanes)
```

## Last time

The *National Health and Nutrition Examination Survey* (NHANES) is an
ongoing CDC survey that collects health and behavioral data from a
nationally representative sample of the U.S. population. participants aged 18-25 were classified into two groups: those who reported **more** sitting time than average ("High"), and those who reported **less** ("Low"). Their depression symptoms were measured with the PHQ-9, a standard 9-item screening scale (0-27, higher = more symptoms). This was translated into a **depression risk score**, with a larger number indicating a higher risk. 

Last time, we used the sample statistics and our trusty intro stat skills to show that there was **no significant difference** between the two groups.

```{r}
#| fig-width: 4
#| fig-height: 2.5
#| out-width: "50%"
#| fig-align: "left"


gf_boxplot(dep_score ~ sitting_group, data = nhanes)
```

A closer look: 

```{r}
#| fig-width: 8
#| fig-height: 2.5
#| echo: false

p1 <- gf_histogram(~dep_score, data = nhanes, xlab = "Depression risk score", fill = "#003300", col = "white", bins = 25)

p2 <- gf_histogram(~sedentary_hours, data = nhanes, xlab = "Hours spent sedentary/sitting", fill = "#003300", col = "white", bins = 15) |>
  gf_vline(xintercept = median_hrs, col = "darkred", linetype = "dashed")

p1 + p2
```

- Response variable: ___________

- Predictor variable: __________



# Regression, Intro Stat version

The main linear regression skills we focus on in intro stat are:

-   Describe the scatterplot
-   Write the equation of the regression line
-   Interpret the slope and intercept
-   Make a prediction

```{r}
#| fig.height: 2
#| fig.width: 4
#| fig.align: center
#| echo: false
gf_jitter(dep_score ~ sedentary_hours, data = nhanes, xlab = "Sitting time (hrs/day)", ylab = "Depression risk score", color = "#003300", size = 1, width = 0.15, height = 0.3) |>
  gf_lm(dep_score ~ sedentary_hours, size = 1.5, color = "#f15a31")
```

**Example:** describe the scatterplot

There is a ___________________ (strength), ___________________
(direction), ___________________ (form) association between
sitting time and depression risk score.

```{r}
#| echo: true
#| eval: false
lm(dep_score ~ sedentary_hours, data = nhanes)
```

```{r}
nhanes_lm$coef
```

**Example:** Write out the regression equation


\vspace{.75in}

**Example:** Interpreting the slope:

\vspace{.75in}

::: {.callout-note}
## Correlation vs Causation
The slope describes an *association* -- an increase in sitting time
doesn't *cause* an increase in depression symptoms just because the
model says so. Watch for language that implies causation from an
observational study.
:::

**Example:** make a prediction. A participant reports sitting for 6 hours per day. The model predicts:

\vspace{.5in}

# Regression, Stat230 Version

## Recap: the Normal distribution

If $Y$ is **numeric**, **unimodal**, and **symmetric**, we might use a
Normal model: $Y \sim N(\mu, \sigma^2)$.

```{r}
#| fig-width: 8
#| fig-height: 2
mu <- c(2, 2, 4); sigma <- c(.5, 1, 2)
normals2 <- tibble(
  setting = factor(rep(1:3, each = 500)),
  x = rep(seq(-2, 10, length = 500), 3),
  mu = rep(mu, each = 500),
  sigma = rep(sigma, each = 500)
) |> mutate(y = dnorm(x, mean = mu, sd = sigma))
levels(normals2$setting) <- paste0("N(", mu, ",", sigma^2, ")")

ggplot(normals2, aes(x, y)) +
  geom_line(color = "#003300") +
  facet_wrap(~setting) +
  labs(x = "y", y = "f(y)") +
  lims(y = c(0, 0.8))
```

## The Simple Linear Regression model

$$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i \quad\text{where}\quad \varepsilon_i \overset{\mathrm{iid}}{\sim} N(0, \sigma^2)$$


```{r}
#| fig-width: 5
#| fig-height: 2.5
nhanes_dsn <- tibble(
  sedentary_hours = c(2, 6, 10, 14),
  yhat = predict(nhanes_lm, data.frame(sedentary_hours = c(2, 6, 10, 14))),
  resid_sd = sigma(nhanes_lm)
)

ggplot(data = nhanes, aes(x = sedentary_hours, y = dep_score)) +
  geom_jitter(alpha = 0.4, size = 1.5, color = "#003300", width = 0.15, height = 0.3) +
  geom_lm(color = "#f15a31") +
  stat_slab(aes(x = sedentary_hours, ydist = "norm", arg1 = yhat, arg2 = resid_sd),
            data = nhanes_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

1.   Linear relationship between $x$ and $y$
2.   Errors are independent and identically distributed (iid)
3.   Errors are normally distributed
4.   Errors have mean 0
5.   Variance of the errors doesn't depend on $x$

## Notation

Mean function: $\;E(Y \mid X) = \mu\{Y \mid X\} = \beta_0 + \beta_1 x_i$

Fitted model equation:
$\;\widehat Y_i = \widehat\mu\{Y \mid X\} = \widehat\beta_0 + \widehat\beta_1 x_i$

## Sampling Variability (preview for next class!) 

```{r}
#| fig-width: 8
#| fig-height: 2
set.seed(1234)
boots <- bootstraps(nhanes, times = 20, apparent = TRUE)
fit_lm_on_bootstrap <- function(split) lm(dep_score ~ sedentary_hours, data = analysis(split))
boot_models <- boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap), coef_info = map(model, tidy))
boot_coefs <- boot_models |>
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

boot_aug3 <- boot_models |>
  slice(1:3) |>
  mutate(augmented = map(model, augment)) |>
  unnest(augmented)
labels <- c("Bootstrap01" = "Sample 1", "Bootstrap02" = "Sample 2", "Bootstrap03" = "Sample 3")

ggplot(boot_aug3, aes(sedentary_hours, dep_score)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = .fitted, group = id), col = "#00aeef") +
  facet_wrap(~id, labeller = as_labeller(labels)) +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

# R Activity

Open the SLR in R activity on Moodle and complete it in your groups.
Make note of any questions, and choose someone responsible for
reporting out your group's answers.


All of the R commands you need are in the file, but here's a reference list of some of the functions you'll learn: 

| Function | Purpose |
|----|-----------|
| `read.csv("file_or_url")` | Load a CSV data set from a file or the web |
| `glimpse(data)` | Get a quick overview of a data set  |
| `mean(x)` | Calculate the mean  |
| `sd(x)` | Calculate the standard deviation  |
| `gf_histogram(~x, data = data, bins = n)` | Create a histogram |
| `gf_boxplot(~x, data = data)` | Create a boxplot |
| `gf_point(y ~ x, data = data)` | Create a scatterplot |
| `gf_lm()` | Add a SLR line to a scatterplot |
| `cor(x, y)` | Calculate the correlation  |
| `lm(y ~ x, data = data)` | Fit a SLR model |
| `summary(model)` | Get detailed info about a fitted model |
| `confint(model, level = 0.95)` | Confidence interval for SLR coefficients |
| `predict(model, newdata = new_data)` | Make predictions using a fitted model |
