---
title: "Day 02: Simple Linear Regression (SLR)"
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
library(rsample)
library(broom)
library(ggdist)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

nhanes <- read_csv(here::here("data/nhanes-sitting-depression.csv")) |>
  mutate(dep_score = log(phq9 + 1))
nhanes_lm <- lm(dep_score ~ sedentary_hours, data = nhanes)
```

## Data for today

The *National Health and Nutrition Examination Survey* (NHANES) is an
ongoing CDC survey that collects health and behavioral data from a
nationally representative sample of the U.S. population. Today we
return to the sitting time and depression data from last class -- this
time looking at the actual numeric variables instead of the
"High"/"Low" groups. We'll summarize each participant's responses to
a 9-item depression screening survey as a **depression risk score**
(higher = more risk).

```{r}
#| fig-width: 8
#| fig-height: 3


p1 <- gf_histogram(~dep_score, data = nhanes, xlab = "Depression risk score", fill = "#003300", col = "white", bins = 25)

p2 <- gf_jitter(dep_score ~ sedentary_hours, data = nhanes, xlab = "Sitting time (hrs/day)", ylab = "Depression risk score",
         color = "#003300", size = 1, width = 0.15, height = 0.3) |>
  gf_lm(linewidth = 1.2, color = "#f15a31")

p1 + p2
```

The data also includes each participant's age and sex. For our first
analysis, we'll investigate the relationship between **sitting time**
and **depression symptoms**.

## Regression, Intro Stat version

**Q: describe the scatterplot**

There is a ___________________ (strength), ___________________
(direction), ___________________ (form) association between
sitting time and depression risk score.

**Q: Write out the regression equation**

```{r}
#| echo: true
lm(dep_score ~ sedentary_hours, data = nhanes)
```

\vspace{1in}

**Q: Interpreting the slope:** 

\vspace{1in}

::: {.callout-note}
The slope describes an *association* -- an increase in sitting time
doesn't *cause* an increase in depression symptoms just because the
model says so. Watch for language that implies causation from an
observational study.
:::

**Q: make a prediction**

A participant reports sitting for 6 hours per day. The model predicts:

\vspace{1in}

## Regression, Stat230 Version

## Recap: the Normal distribution

If $Y$ is **numeric**, **unimodal**, and **symmetric**, we might use a
Normal model: $Y \sim N(\mu, \sigma^2)$.

```{r}
#| fig-width: 6
#| fig-height: 2.6
mu <- c(2, 2, 4); sigma <- c(.5, 1, 2)
normals2 <- tibble(
  setting = factor(rep(1:3, each = 500)),
  x = rep(seq(-2, 10, length = 500), 3),
  mu = rep(mu, each = 500),
  sigma = rep(sigma, each = 500)
) |> mutate(y = dnorm(x, mean = mu, sd = sigma))
levels(normals2$setting) <- paste0("N(", mu, ",", sigma^2, ")")

ggplot(normals2, aes(x, y)) +
  geom_line(color = "#003300") +
  facet_wrap(~setting) +
  labs(x = "y", y = "f(y)") +
  lims(y = c(0, 0.8))
```

## The Simple Linear Regression model

$$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i \quad\text{where}\quad \varepsilon_i \overset{\mathrm{iid}}{\sim} N(0, \sigma^2)$$

A **deterministic** model $Y_i = \beta_0 + \beta_1 x_i$ (no error
term) would place every point exactly on the line -- no uncertainty.
Real data doesn't look like that. Instead, at every $x$ value we
imagine a full Normal *distribution* of possible $y$'s, centered on
the line, all sharing the same spread $\sigma$:

```{r}
#| fig-width: 6
#| fig-height: 3
nhanes_dsn <- tibble(
  sedentary_hours = c(2, 6, 10, 14),
  yhat = predict(nhanes_lm, data.frame(sedentary_hours = c(2, 6, 10, 14))),
  resid_sd = sigma(nhanes_lm)
)

ggplot(data = nhanes, aes(x = sedentary_hours, y = dep_score)) +
  geom_jitter(alpha = 0.4, size = 1.5, color = "#003300", width = 0.15, height = 0.3) +
  geom_lm(color = "#f15a31") +
  stat_slab(aes(x = sedentary_hours, ydist = "norm", arg1 = yhat, arg2 = resid_sd),
            data = nhanes_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

**SLR assumptions** (mnemonic: think "LINE"):

- **L**inear relationship between $x$ and $y$
- Errors are **I**ndependent and identically distributed
- Errors are **N**ormally distributed
- **E**qual variance of errors across all $x$ (doesn't depend on $x$)
- (and errors have mean 0)

## Notation

Mean function: $\;E(Y \mid X) = \mu\{Y \mid X\} = \beta_0 + \beta_1 x_i$

Fitted model equation:
$\;\widehat Y_i = \widehat\mu\{Y \mid X\} = \widehat\beta_0 + \widehat\beta_1 x_i$

## Uncertainty in the parameters

Every new sample of participants gives a slightly different fitted
line. Below, each panel refits the model on a different bootstrap
resample of the NHANES data:

```{r}
#| fig-width: 7
#| fig-height: 2.8
set.seed(1234)
boots <- bootstraps(nhanes, times = 20, apparent = TRUE)
fit_lm_on_bootstrap <- function(split) lm(dep_score ~ sedentary_hours, data = analysis(split))
boot_models <- boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap), coef_info = map(model, tidy))
boot_coefs <- boot_models |>
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

boot_aug3 <- boot_models |>
  slice(1:3) |>
  mutate(augmented = map(model, augment)) |>
  unnest(augmented)
labels <- c("Bootstrap01" = "Sample 1", "Bootstrap02" = "Sample 2", "Bootstrap03" = "Sample 3")

ggplot(boot_aug3, aes(sedentary_hours, dep_score)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = .fitted, group = id), col = "#00aeef") +
  facet_wrap(~id, labeller = as_labeller(labels)) +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score")
```

Overlaying all 20 resampled lines on one plot shows that variability
directly -- that spread in the fitted ___________________ (not the
raw data) is exactly what will let us build confidence intervals and
run hypothesis tests next class:

```{r}
#| fig-width: 6
#| fig-height: 3
ggplot(data = boot_coefs) +
  geom_abline(aes(slope = sedentary_hours, intercept = `(Intercept)`),
              color = "#0072B2", linewidth = 0.5, alpha = 0.3) +
  geom_point(data = nhanes, aes(sedentary_hours, dep_score), color = "grey60") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score", title = "20 bootstrap resamples")
```

## Rest of today

::: {.callout-tip}
## Group task

Open the SLR in R activity on Moodle and complete it in your groups.
Make note of any questions, and choose someone responsible for
reporting out your group's answers.
:::
