---
title: "Day 03: Inference for SLR"
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
library(broom)
library(rsample)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")
nhanes <- read_csv(here::here("data/nhanes-sitting-depression.csv")) |>
  mutate(dep_score = log(phq9 + 1))
nhanes_lm <- lm(dep_score ~ sedentary_hours, data = nhanes)
```

## Data for today

We're returning to the NHANES sitting time and depression risk score
data. We're analyzing the relationship between depression risk score
and sitting time.

::: callout-note
## Theoretical Model:

$(x_i, Y_i)$ are the data points; \underline{\hspace{1.5cm}} and \underline{\hspace{1.5cm}} are (unknowable) population parameters


 $\underline{\hspace{1.5cm}} = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i\overset{iid}{\sim} N (0, \underline{\hspace{1.5cm}})$
:::

::: callout-note
## Fitted Model:

$\widehat{Y}_i$ are predictions; \underline{\hspace{1.5cm}} and \underline{\hspace{1.5cm}} are estimated from the sample


 $\underline{\hspace{1.5cm}} = \widehat \beta_0 + \widehat \beta_1 x_i$
:::

```{r}
#| echo: true
summary(nhanes_lm)
```



## Sampling variability

Every new sample gives us a slightly different fitted line.

```{r}
#| fig-width: 8
#| fig-height: 3
set.seed(1234)
boots <- bootstraps(nhanes, times = 20, apparent = TRUE)
fit_lm_on_bootstrap <- function(split) lm(dep_score ~ sedentary_hours, data = analysis(split))
boot_models <- boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))
boot_coefs <- boot_models |>
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

boot_slrs <- ggplot(boot_coefs) +
  geom_abline(aes(slope = sedentary_hours, intercept = `(Intercept)`),
              color = "#00aeef", alpha = 0.6, linewidth = 0.5) +
  geom_point(data = nhanes, aes(sedentary_hours, dep_score), color = "grey60") +
  labs(x = "Sitting time (hrs/day)", y = "Depression risk score", title = "SLRs from many samples")

boots1k <- bootstraps(nhanes, times = 1000, apparent = TRUE)


boot_models_1k <-
  boots1k |>
  mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))

boot_coefs <- 
  boot_models_1k %>% 
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  rename(
    intercept = `(Intercept)`,
    slope = sedentary_hours
  )

slope_dist <- ggplot(boot_coefs) +
   ggdist::stat_dotsinterval(aes(x = slope), quantiles = 500, scale = 0.6, position = position_nudge(y = 0.175)) +
  labs(y = "Density", x = "Slopes")

intercept_dist <- ggplot(boot_coefs) +
  ggdist::stat_dotsinterval(aes(x = intercept), quantiles = 500, scale = 0.6, position = position_nudge(y = 0.175))  +
  labs(y = "Density", x = "Intercepts")

boot_slrs + (slope_dist / intercept_dist)

```

::: callout-tip
## Sampling Distribution

If we obtained many different random samples, what would the distribution of their ___________________ ($\widehat \beta_0$ and $\widehat\beta_1$) look like?
:::

We use the **t distribution** to model the sampling distribution of a
regression coefficient.

-   Degrees of freedom: $df =$ ___________________
-   Mean: $= \widehat\beta_i$
-   Standard deviation: $SE(\widehat\beta_i) = \dfrac{\widehat\sigma}{\sqrt{\sum (x_i - \overline x)^2}}$

## Confidence interval

General Form:

$$\text{statistic} \pm \text{critical value} \times \text{standard error}$$

Form for regression coefficient: 

$$\widehat\beta_i \pm t^* \cdot SE(\widehat\beta_i)$$

::: {.callout-note}
## Critical value

$t^*$ is the value that cuts off the right tail area that matches your confidence level. For a 95% CI, find $t^*$ so that the two-sided tail area is 0.025 on each side, using `qt(.025, df = n-2)`.
:::

## Hypothesis test 

$$H_0: \beta_i = \text{null value} \quad \text{vs.} \quad H_a: \beta_i \; \begin{cases}\ne\\<\\>\end{cases} \text{null value}$$

$$\text{test statistic} = \frac{\text{estimate} - \text{null value}}{SE} = \frac{\widehat\beta_i - \text{null value}}{SE(\widehat\beta_i)}$$

Then find the appropriate tail area(s) under the t-distribution with
$df = n - 2$.

```{r}
#| echo: true 
tidy(nhanes_lm)
```

::: {.callout-tip}
## Try it

Verify the test statistic for $H_0: \beta_1 = 0$ against
$H_A: \beta_1 \ne 0$ using the table above.

\vspace{1in}
:::

**Example:** 90% confidence interval for $\beta_1$

$$\widehat\beta_i \pm t^* \cdot SE(\widehat\beta_i)$$

```{r}
t_star <- qt(0.95, df = nrow(nhanes) - 2)
ci90 <- confint(nhanes_lm, level = 0.90)
slope_est <- coef(nhanes_lm)[["sedentary_hours"]]
```

$t^* =$ 

```{r}
#| echo: true
qt(0.95, df = 659 - 2)
```


\vspace{2in}

**Interpretation:**

$~$

::: {.callout-tip}
## General template (reusable for any CI)

- We are _______% confident that for every 1 _________ increase in ____________________, 

- _____________________________________________________  also ________________ 

- by between ____________ and _____________

:::

$~$

**Discussion:** Recall from Day01: we found no significant difference between the "High" and "Low" sitting groups. How does this result compare? Which do you trust more? 


