---
title: "Day 12: Inference for linear combinations"
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
library(broom)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt",
                   sep = "\t", header = TRUE) |>
  mutate(Smoke = as.logical(Smoke))

fev_lm <- lm(FEV ~ Age * Smoke, data = fev)
V <- vcov(fev_lm)
b <- coef(fev_lm)
df_resid <- fev_lm$df.residual
```

We're interested in modeling **lung capacity** (FEV) on **age** and **smoking status** for a sample of adolescents. As kids get older, FEV generally increases, but we also think FEV decreases for smokers. Since smokers are likely to be older, we need a model that accounts for both.

```{r}
#| fig-height: 2.5
#| fig-width: 4.5
#| fig-align: left
ggplot(fev, aes(x = Age, y = FEV, color = Smoke)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("#f15a31", "#000036")) +
  labs(y = "FEV (L/s)") +
  theme(legend.position = "top")
```

Theoretical mean model: 

$$\mu\{FEV \mid Age, Smoke\} = \beta_0 + \beta_1\, Age + \beta_2\, Smoke + \beta_3\, Age \times Smoke$$

```{r}
kable_coef(fev_lm, terms = c("Intercept", "Age", "SmokeTRUE", "Age:SmokeTRUE"))
```

**Example:** Models for each group

Nonsmokers ($Smoke = 0$): $\mu\{FEV \mid Age\} =$ \vspace{.1in}

\hspace{.5in} $\hat y=$ \vspace{.1in}

Smokers ($Smoke = 1$): $\mu\{FEV \mid Age\} =$ \vspace{.1in}

\hspace{.5in} $\hat y=$ \vspace{.1in}

## Reading the table: what is significantly different than 0?

- The intercept for non-smokers is ___________
- The intercept offset for smokers is ___________
- The slope for non-smokers is ___________
- The *difference* in slopes between smokers and non-smokers is __________
- The slope for smokers is ____________

The slope for smokers is $\beta_1 + \beta_3$. The table gives us $SE(\widehat\beta_1)$ and $SE(\widehat\beta_3)$, but
**not** the standard error of their sum.

**Why can't we just add $SE(\widehat\beta_1) + SE(\widehat\beta_3)$?**

\vspace{1.25in}

**Why can't we use the p-value for `Age:SmokeTRUE` to test whether the smoker slope is zero?**

\vspace{1.25in}


::: callout-tip
## Definition

A **linear combination** of coefficients has the form

$$\gamma = c_1 \beta_i + c_2 \beta_j$$

where $c_1$ and $c_2$ are constants

**Estimate:** $\widehat\gamma =$ _________________

**Standard error:**

$$SE(\widehat\gamma) = \sqrt{c_1^2 \{SE(\widehat\beta_i)\}^2 + c_2^2 \{SE(\widehat\beta_j)\}^2 + 2 c_1 c_2 \, Cov(\widehat\beta_i, \widehat\beta_j)}$$

::: 

The extra term is there because $\widehat\beta_i$ and $\widehat\beta_j$ come
from the *same* fitted model, so their estimation errors are not
independent. $Cov(\widehat\beta_i, \widehat\beta_j)$ measures how they move together.


## The covariance matrix of the coefficients

```{r}
#| echo: true
vcov(fev_lm) |> round(4)
```

The **diagonal** entries are the variances, $V(\widehat \beta_j) = \{SE(\widehat\beta_j)\}^2$.

  - Check: $SE(\widehat\beta_1) =$ ________ , so $\{SE(\widehat\beta_1)\}^2 =$ ________

The **off-diagonal** entries are the covariances, $Cov(\widehat\beta_i, \widehat\beta_j)$.
-   The matrix is symmetric: $Cov(\widehat\beta_i, \widehat\beta_j) = Cov(\widehat\beta_j, \widehat\beta_i)$.

## Example 1: CI for the smoker slope

**Target:** $\gamma = \beta_1 + \beta_3$. So $c_1 =$_____ and $c_2=$_________

**Estimate:**

\vspace{.5in}

**Standard error:** pull the pieces out of the covariance matrix.

| piece | value |
|:--|:--|
| $\{SE(\widehat\beta_1)\}^2$ |  |
| $\{SE(\widehat\beta_3)\}^2$ |   |
| $Cov(\widehat\beta_1, \widehat\beta_3)$ |   |

\vspace{1in}

**90% CI:** $\widehat\gamma \pm t^*_{n-(p+1)} \cdot SE(\widehat\gamma)$, where $n = 345$, $p = 3$,
so $df = 341$ and $t^* \approx 1.649$.

\vspace{1in}

**Interpretation:**

\vspace{.75in}

::: {.callout-note}
## What if we ignored the covariance?

Using only $\sqrt{\{SE(\widehat\beta_1)\}^2 + \{SE(\widehat\beta_3)\}^2}$ would
give `r round(sqrt(V[2,2] + V[4,4]), 3)`. The correct SE is
`r round(sqrt(V[2,2] + V[4,4] + 2 * V[2,4]), 3)`. Here the covariance is
**negative**, so ignoring it makes the interval too wide.
:::

## Testing a linear combination

**Hypotheses:** $H_0: \gamma = \gamma_0$ vs. $H_a: \gamma \ne, <, \text{or} > \gamma_0$

**Test statistic:** $t = \dfrac{\widehat\gamma - \gamma_0}{SE(\widehat\gamma)}$

**Reference distribution:** $t$ distribution with $df = n - (p+1)$

**p-value:** area in the tail(s) specified by $H_a$

Same four steps as a test for a single coefficient. Only the estimate
and the SE change.

**Example:** Is the smoker slope different from zero? Using
$\widehat\gamma =$ `r round(b[2] + b[4], 4)` and
$SE(\widehat\gamma) =$ `r round(sqrt(V[2,2] + V[4,4] + 2 * V[2,4]), 4)`:

\vspace{1in}

## Your turn: smokers vs. nonsmokers at age 15

At age 15, how different is the mean FEV of smokers and nonsmokers?

**0.** Look back and the graph and identify what this question is asking about. What do you think you will find?

\vspace{1in}

**1.** Write the theoretical mean function for each group evaluated at $Age = 15$,
and subtract (smoker $-$ nonsmoker) to find $\gamma$.

\vspace{1.25in}

**2.** The difference is $\gamma =$___________________________, so $c_2 =$ ______ $\qquad c_3 =$ ______

**3.** Find $\hat\gamma$

\vspace{1in}

**3.** Write out $SE(\widehat\gamma)$ using the values from the
covariance matrix.

\vspace{2in}

**4.** Construct a 95% CI for $\gamma$ ($df = 341$, $t^* \approx 1.967$).

\vspace{1.5in}

**5.** Test $H_0: \gamma = 0$ vs. $H_a: \gamma \ne 0$.

\vspace{1.5in}

**6.** Interpret your results in context.

\vspace{1in}

## A shortcut: re-center the model

If we measure age as "years since
age 15", the coefficient on `Smoke` *is* the smoker-minus-nonsmoker
difference at age 15, and the standard table gives us its SE directly:

```{r}
#| echo: true
fev <- mutate(fev, Age15 = Age - 15)
lm(FEV ~ Age15 * Smoke, data = fev) |> tidy()
```

Compare the `SmokeTRUE` row to the output from the previous section. 

## Summary

| To do inference for... | Use | 
|:--|:--|:--|
| a single coefficient $\beta_j$ | $t$-test / CI from the coefficient table | 
| a linear combination $c_1\beta_i + c_2\beta_j$ | $t$-test / CI with $SE = \sqrt{c_1^2 \{SE(\widehat\beta_i)\}^2 + c_2^2 \{SE(\widehat\beta_j)\}^2 + 2 c_1 c_2 \, Cov(\widehat\beta_i, \widehat\beta_j)}$ | 
| several coefficients at once ($\beta_2 = \beta_3 = 0$) | extra sums of squares $F$-test | 

-   Never add standard errors. Use the covariance matrix.
-   The reference distribution is always $t_{n-(p+1)}$ for a single
    linear combination.
-   A linear combination is *one* number, so we use a $t$ procedure. When
    $d > 1$ coefficients are tested at the same time, that's an $F$-test.

## Your turn: in R

The R activity is posted on moodle. Here are some functions that might be helpful: 

| Function | Purpose |
|----|-----------|
| `coef()` | Extract the estimated coefficients from a fitted model (as a named vector) |
| `vcov()` | Extract the covariance matrix of the estimated coefficients from a fitted model |
| `x[i]`, `m[i, j]` | Pull off element `i` of a vector, or the entry in row `i` and column `j` of a matrix |
| `sqrt()` | Take the square root (e.g., to turn a variance into a standard error) |