---
title: "Day 09: Intro to Multiple Linear Regression"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

bikes_slr <- lm(rides ~ temp_actual, data = bikes)
bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)

bikes <- bikes |>
  mutate(
    temp_centered = temp_actual - mean(temp_actual, na.rm = TRUE),
    wind_centered = windspeed - mean(windspeed, na.rm = TRUE),
    temp_std = (temp_actual - mean(temp_actual, na.rm = TRUE)) / sd(temp_actual, na.rm = TRUE),
    wind_std = (windspeed - mean(windspeed, na.rm = TRUE)) / sd(windspeed, na.rm = TRUE)
  )
bikes_mlr_c <- lm(rides ~ temp_centered + wind_centered, data = bikes)
bikes_mlr_std <- lm(rides ~ temp_std + wind_std, data = bikes)
```

## Back to `bikes`: one predictor

$$\widehat{\text{rides}} = `r round(coef(bikes_slr)[1],2)` + `r round(coef(bikes_slr)[2],2)`(\text{temp\_actual})$$

```{r}
#| fig-width: 6
#| fig-height: 3
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides",
         color = "#003300", size = 1) |>
  gf_lm(linewidth = 1.2, color = "#f15a31")
```

$R^2 =$ `r round(summary(bikes_slr)$r.squared, 3)` -- temperature alone
leaves a lot of variation in ridership unexplained. What if we add a
second predictor?

## Pairs plot and correlation matrix

```{r}
#| fig-width: 6
#| fig-height: 4.5
bikes |>
  select(temp_actual, windspeed, rides) |>
  GGally::ggpairs(progress = FALSE)
```

```{r}
#| echo: true
bikes |>
  select(temp_actual, windspeed, rides) |>
  cor() |> round(3)
```

## Multiple Linear Regression (MLR)

$$Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots + \beta_p x_{ip} + \varepsilon_i, \quad \varepsilon_i \overset{\mathrm{iid}}{\sim} N(0, \sigma^2)$$

This model requires:

1. $\mu\{Y \mid x_1, x_2, \ldots, x_p\}$ is a ___________________
   function of the predictors
2. The error terms are ___________________
3. The error terms follow a Normal distribution with mean 0 and
   variance $\sigma^2$

## Fitting the model

**Procedure:** least squares estimation chooses the coefficients to
minimize

$$\text{SSE} = \sum_{i=1}^n \left(Y_i - \widehat{Y}_i\right)^2, \qquad \text{then} \quad \widehat\sigma = \sqrt{\frac{\text{SSE}}{n - (p+1)}}$$

```{r}
#| echo: true
bikes_mlr <- lm(rides ~ temp_actual + windspeed, data = bikes)
coef(bikes_mlr)
```

## Interpreting $\widehat\beta_0$

```{r}
coef(bikes_mlr) |> round(3)
```

$\widehat\beta_0$ is the expected value of the response when
___________________. Here: the expected number of rides is about
`r round(coef(bikes_mlr)[1], 1)` when temperature = 0 and windspeed =
0 -- not a meaningful day in this dataset!

## Interpreting $\widehat\beta_j$

$\widehat\beta_j$ is the expected change in the response for a
1-unit increase in $x_j$, when ___________________________________.

- We expect `r round(coef(bikes_mlr)[2],2)` *additional* rides for
  every 1 degree increase in actual temperature, holding windspeed
  constant.
- We expect `r round(abs(coef(bikes_mlr)[3]),2)` *fewer* rides for
  every 1 MPH increase in windspeed, holding temperature constant.

## Geometric interpretation: the MLR "line" is a plane

With two predictors, the fitted model is no longer a line but a
**plane** through 3-dimensional space:

```{r}
#| fig-width: 6
#| fig-height: 5
library(scatterplot3d)
s3d <- scatterplot3d(bikes$temp_actual, bikes$windspeed, bikes$rides,
                      angle = 55, pch = 16, color = "#003300",
                      xlab = "Temperature", ylab = "Windspeed", zlab = "No. rides")
s3d$plane3d(bikes_mlr, draw_polygon = TRUE, polygon_args = list(col = adjustcolor("#f15a31", alpha.f = 0.3)))
```

## Centering predictor variables

*Idea:* recenter each predictor at its own mean, so an intercept of
"all predictors = 0" becomes an intercept of "all predictors at their
___________________" -- a value that actually occurs in the data.

```{r}
#| echo: true
bikes <- bikes |>
  mutate(
    temp_centered = temp_actual - mean(temp_actual, na.rm = TRUE),
    wind_centered = windspeed - mean(windspeed, na.rm = TRUE)
  )
```

Centering shifts the mean to 0 but leaves the spread alone:

```{r}
bikes |>
  summarize(mean_orig = mean(temp_actual), mean_centered = mean(temp_centered),
            sd_orig = sd(temp_actual), sd_centered = sd(temp_centered)) |>
  round(2) |> knitr::kable()
```

## Interpreting the centered model

```{r}
bikes_mlr_c <- lm(rides ~ temp_centered + wind_centered, data = bikes)
coef(bikes_mlr_c) |> round(3)
```

- $\widehat\beta_0$: the expected number of rides is about
  `r round(coef(bikes_mlr_c)[1])` on an *average* temperature and
  windspeed day -- much more interpretable than before.
- $\widehat\beta_j$'s **don't change** -- they still describe the
  expected change for a 1-unit increase in the original units of
  $x_j$, holding the other predictor constant.

## Standardizing predictor variables

*Idea:* 1 MPH of windspeed isn't comparable to 1°F of temperature, so
comparing raw slopes across predictors can be misleading. Standardize
each predictor to have mean 0 and SD 1:

```{r}
#| fig-width: 6
#| fig-height: 2.6
p1 <- gf_histogram(~temp_centered, data = bikes, col = "white", fill = "#003300")
p2 <- gf_histogram(~wind_centered, data = bikes, col = "white", fill = "#003300")
p1 + p2
```

```{r}
#| echo: true
bikes <- bikes |>
  mutate(
    temp_std = (temp_actual - mean(temp_actual, na.rm = TRUE)) / sd(temp_actual, na.rm = TRUE),
    wind_std = (windspeed - mean(windspeed, na.rm = TRUE)) / sd(windspeed, na.rm = TRUE)
  )
```

```{r}
bikes |>
  summarize(mean_orig = mean(temp_actual), mean_std = mean(temp_std),
            sd_orig = sd(temp_actual), sd_std = sd(temp_std)) |>
  round(2) |> knitr::kable()
```

## Interpreting the standardized model

```{r}
bikes_mlr_std <- lm(rides ~ temp_std + wind_std, data = bikes)
coef(bikes_mlr_std) |> round(3)
```

- $\widehat\beta_0$: expected rides on an average day is unchanged
  from the centered model (`r round(coef(bikes_mlr_std)[1])`).
- $\widehat\beta_j$: now the expected change in the response for a 1
  ___________________ increase in $x_j$, holding the other predictor
  constant.
  - We expect `r round(coef(bikes_mlr_std)[2])` *additional* rides
    for every 1 SD increase in temperature
    (`r round(sd(bikes$temp_actual),1)` degrees), holding windspeed
    constant.
  - We expect `r round(abs(coef(bikes_mlr_std)[3]))` *fewer* rides
    for every 1 SD increase in windspeed
    (`r round(sd(bikes$windspeed),1)` MPH), holding temperature
    constant.

::: {.callout-note}
Standardized coefficients put predictors measured in different units
on a common scale, so their magnitudes become directly comparable.
:::
