---
title: "Day 05: Evaluating SLR models"
author: "Name: ________________________________"
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
library(broom)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

city <- readr::read_csv("https://raw.githubusercontent.com/zief0002/modeling/main/data/riverview.csv")
city_lm <- lm(income ~ education, data = city)
```

## Data for today

We return to the Riverview data (income vs. education) from Day 4 --
a nice size for spotting patterns by eye, and we don't need to
subsample it.

```{r}
gf_point(income ~ education, data = city, xlab = "Education (years)",
         ylab = "Income ($1000s)", color = "#003300") |>
  gf_lm(color = "#f15a31")
```

```{r}
kable_coef(city_lm, terms = c("Intercept", "education"))
```

## Two definitions of a "good" model

1.  The model is appropriate, so our inferences (p-values, CIs) are
    ___________________________
2.  The x variable does a good job ___________________________ the y
    variable

Theoretical model: $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where
$\varepsilon_i \overset{\mathrm{iid}}{\sim} N(0, \sigma^2)$

Fitted model: $\widehat y_i = \widehat\beta_0 + \widehat\beta_1 x_i$,
so that $y_i = \widehat y_i + \widehat\varepsilon_i$

## Conditions for inference: LINE

For our p-values and confidence intervals to be trustworthy, the
theoretical model above has to be a reasonable description of the
data. Check:

-   **L**inear: ___________________________
-   **I**ndependent and identically distributed errors:
    ___________________________
-   **N**ormally distributed errors: ___________________________
-   **E**qual variance of the errors: ___________________________

::: {.callout-note}
These conditions are about the *errors*, not about the raw x or y
variables themselves.
:::

## Residuals

**Definition:** $e_i = \widehat\varepsilon_i = y_i - \widehat y_i$

**Properties:**

-   sum to zero $\Longrightarrow$ mean is ___
-   **uncorrelated** with $x$ and $\widehat y$
-   (should be) normally distributed
-   $SD(e_i) = \widehat\sigma \sqrt{1 - \dfrac1n - \dfrac{(x_i - \overline x)^2}{\sum (x_i - \overline x)^2}}$

## Residual plots

For SLR, plotting residuals against $x$ or against the fitted values
$\widehat y$ shows the same shape (since $\widehat y$ is a linear
function of $x$) -- either is a valid way to check linearity and
equal variance.

```{r}
#| fig-width: 8
#| fig-height: 3
p1 <- ggplot(augment(city_lm), aes(x = education, y = .resid)) +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linewidth = 1) +
  labs(x = "Education", y = "Residual")

p2 <- ggplot(augment(city_lm), aes(x = .fitted, y = .resid)) +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linewidth = 1) +
  labs(x = "Fitted values", y = "Residual")

p1 + p2
```

**A "good" residual plot** shows no pattern: residuals bounce randomly
around zero with roughly constant spread across the range of fitted
values.

```{r}
#| fig-height: 3
#| fig-width: 4.5
set.seed(72384)
x <- rnorm(100, mean = 10, sd = 3.5)
y <- 1 + 1 * x + rnorm(100, mean = 0, sd = 10)
good_mod <- lm(y ~ x)

gf_point(residuals(good_mod) ~ fitted(good_mod),
         ylab = "Residual", xlab = "Fitted values") |>
  gf_hline(yintercept = 0, color = "#f15a31", linetype = 2)
```

::: {.callout-tip}
## Group task

On the whiteboards, sketch a plot of $y$ vs. $x$ *and* the
corresponding residual plot that would indicate a violation of

1.  the linearity condition
2.  the equal variance condition
:::

## Assessing normality

Use a histogram or normal Q-Q plot of the residuals. A "good" Q-Q plot
hugs the diagonal reference line closely.

```{r}
#| fig-width: 8
#| fig-height: 3
p1 <- gf_histogram(~ .resid, data = augment(city_lm), color = "white",
                    bins = 10, xlab = "Residual", ylab = "Count")

p2 <- gf_qq(~ .resid, data = augment(city_lm),
            ylab = "Residual", xlab = "N(0,1) quantiles") |>
  gf_qqline()

p1 + p2
```

::: {.callout-note}
Normality is **harder to judge at small sample sizes** -- even true
random draws from a normal distribution can look lumpy or have a
wandering Q-Q plot when $n$ is small. Don't over-interpret minor
wiggles with only 30-50 points.
:::

## Assessing independence

Plot residuals against whatever variable might induce dependence
(time, location, subject, batch, ...). A clear pattern -- rather than
random bouncing around zero -- is a strong hint that the independence
condition is violated. For example, if employees were hired in waves
and residuals cycled up and down with the order they were added to
the dataset, that would be a red flag:

```{r}
#| fig-height: 3
set.seed(040726)
bad_indep <- tibble(
  index = 1:50,
  resid = 15 * sin(index / 5) + rnorm(50, sd = 5)
)

ggplot(bad_indep, aes(x = index, y = resid)) +
  geom_line(color = "#003300") +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linetype = 2) +
  labs(x = "Order collected", y = "Residual")
```

## What happens if the conditions aren't met?

-   **Linearity:** if nonlinear, everything breaks!
-   **Independence:** estimates are still unbiased, but the SEs are
    typically too **small** (we have less information than $n$
    independent points suggests)
-   **Normality:** estimates and SEs are still fine, but **prediction
    intervals** (which rely on individual points being normal around
    the mean) are wrong
-   **Equal variance:** estimates are still unbiased, but SEs are
    wrong -- and we don't know in which direction

## What do we do if assumptions are violated?

1.  Change our assumptions (hard -- needs more advanced methods)
2.  Transform $y$, $x$, or both
3.  Change the type of inference (e.g. bootstrap)

# Is x a "good" predictor of y?

## Partitioning variability

```{r}
#| fig-width: 9
#| fig-height: 3
slr_aug <- augment(city_lm) |> mutate(mean.y = mean(income))

p1 <- ggplot(slr_aug, aes(x = education, y = income)) +
  geom_point(color = "#003300") +
  geom_segment(aes(xend = education, yend = mean.y), color = "gray70", alpha = 0.6) +
  geom_hline(yintercept = mean(slr_aug$income), color = "#f15a31") +
  labs(title = "SS Total")

p2 <- ggplot(slr_aug, aes(x = education, y = .fitted)) +
  geom_point(color = "#003300") +
  geom_segment(aes(xend = education, yend = mean.y), color = "gray70", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "#f15a31") +
  geom_hline(yintercept = mean(slr_aug$income), color = "#f15a31", linetype = "dashed") +
  labs(title = "SS Regression", y = "income")

p3 <- ggplot(slr_aug, aes(x = education, y = income)) +
  geom_point(color = "#003300") +
  geom_segment(aes(xend = education, yend = .fitted), color = "#003069", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "#f15a31") +
  labs(title = "SS Error")

p1 + p2 + p3
```

Each gray/blue segment is one squared deviation in the sum -- same $x_i$'s, three different "distances": how far each point is from $\overline y$ (Total), how far the model's prediction $\widehat y_i$ is from $\overline y$ (Regression -- the part $x$ explains), and how far each point is from its own prediction $\widehat y_i$ (Error -- the residuals).

$$\underbrace{\sum (y_i - \overline y)^2}_{SSTO} = \underbrace{\sum (\widehat y_i - \overline y)^2}_{SSR} + \underbrace{\sum e_i^2}_{SSE}$$

$$\text{total variability} = \text{variability of } \widehat Y\text{'s} \; + \; \text{variability of residuals (}\underline{\hspace{2cm}}\text{)}$$

$$R^2 = \frac{SSR}{SSTO} = \frac{\text{variability of } \widehat Y}{\text{variability of } Y}$$

::: {.callout-note}
## $R^2$

$R^2$ gives the proportion of total variability in $Y$ explained by
the SLR model.
:::

For the Riverview data, $R^2 =$ `r round(summary(city_lm)$r.squared, 3)`
-- about `r round(summary(city_lm)$r.squared, 3) * 100`% of the
variability in income is explained by education.

## Properties of $R^2$

-   Between ___ and ___
-   Big when $\sigma$ is small relative to the overall variability of
    $y$ (i.e. a big proportion of variability explained by the model)
-   For SLR, $R^2 =$ ___ (where $r$ is the correlation coefficient
    between $x$ and $y$)

::: {.callout-note}
## Quick notation reference

| Symbol | Meaning |
|:---|:---|
| $r$ | sample correlation coefficient |
| $\rho$ | population correlation coefficient |
| $R^2$ | coefficient of determination |
| $p$ | a population proportion |
| $\widehat p$ | a sample proportion |
| p-value | $P(\text{test statistic} \mid H_0)$ |
| $P(A)$ | probability of event $A$ |
:::

## $R^2$ dangers: high $R^2$ doesn't mean "good fit"

```{r}
#| fig-width: 8
#| fig-height: 3.2
set.seed(040726)
fake_data <- tibble(
  x = c(runif(50, 0, 200), 400),
  y = c(.5 * x[1:50]^2 + rnorm(50, 0, 50), 30000)
)
fake_lm <- lm(y ~ x, data = fake_data)

p1 <- ggplot(fake_data, aes(x = x, y = y)) +
  geom_point(color = "#003300") +
  geom_smooth(method = "lm", se = FALSE, color = "#f15a31") +
  labs(title = paste0("R² = ", round(summary(fake_lm)$r.squared, 3)))

p2 <- ggplot(augment(fake_lm), aes(x = .fitted, y = .resid)) +
  geom_point(color = "#003300") +
  geom_hline(yintercept = 0, color = "#f15a31", linetype = 2) +
  labs(x = "Fitted values", y = "Residual")

p1 + p2
```

A single high-leverage point drove $R^2$ up to
`r round(summary(fake_lm)$r.squared, 3)` even though the residual plot
shows an obvious curve -- the linear model is a bad fit! **Always check
the residual plot, not just $R^2$.**

::: {.callout-tip}
## Your turn

Would you trust this model just based on its $R^2$? What does the
residual plot tell you that $R^2$ alone can't?
:::
