---
title: "Day 04: Inference for Prediction"
author: "Name: ________________________________"
---

```{r}
#| include: false
library(tidyverse)
library(bayesrules)
library(ggformula)
library(rsample)
library(broom)
library(ggdist)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
source("notes-helpers.R")

bikes_lm <- lm(rides ~ temp_actual, data = bikes)
```

## Last time, and today

Last time: the sampling distribution for $\beta_i$'s, confidence
intervals and hypothesis tests for $\beta_i$'s, both by hand and in R.

**Today:** inference for *predictions*.

## Two types of predictions

1. Predicting the **mean response** at a specific value of $x$ --
   e.g. the *average* number of riders on a 62 degree day.
2. Predicting the response for a **specific future observation** --
   e.g. the number of riders *tomorrow* (forecast to be 62 degrees).

::: {.callout-tip}
Think of two additional real-world examples of each type of
prediction.
:::

## Inference for prediction

Best estimate: $\widehat y_0 = \widehat\beta_0 + \widehat\beta_1 x_0$

General interval formula:
$$\text{estimate} \pm t^* \times SE$$

- $\widehat y_0$ is our estimate
- use a $t$-distribution with $df = n-2$ to find $t^*$

We'll need a *different* SE depending on whether we're building:

- a confidence interval for the mean response $\widehat\mu(Y \mid X_0)$, or
- a prediction interval for an individual response $Y \mid X_0$

## Standard errors

$$SE\big(\widehat\mu(Y\mid X)\big) = \widehat\sigma \sqrt{\frac{1}{n} + \frac{(x_0-\overline x)^2}{\sum_{i=1}^n (x_i-\overline x)^2}}$$

$$SE(\widehat y) = \widehat\sigma \sqrt{1 + \frac{1}{n} + \frac{(x_0-\overline x)^2}{\sum_{i=1}^n (x_i-\overline x)^2}}$$

Comparing the two formulas: the prediction interval SE has an extra
"$+1$" under the square root, so it will always be
___________________ (wider/narrower) than the confidence interval SE.

As $x_0$ moves farther from $\overline x$, both standard errors
___________________ (grow/shrink), because
$(x_0-\overline x)^2$ grows.

## Where the extra variability comes from

$$SE(\widehat y) = \sqrt{\underbrace{\widehat\sigma^2}_{\text{variability in } \varepsilon_i\text{'s}} + \underbrace{\frac{\widehat\sigma^2}{n} + \frac{\widehat\sigma^2(x_0-\overline x)^2}{\sum_{i=1}^n (x_i-\overline x)^2}}_{\text{variability in } \widehat\beta_i\text{'s}}}$$

The first term is the leftover noise around the line at any $x$; the
rest comes from not knowing the true line exactly.

```{r}
#| fig-width: 7
#| fig-height: 2.8
set.seed(1234)
boots <- bootstraps(bikes, times = 20, apparent = TRUE)
fit_lm_on_bootstrap <- function(split) lm(rides ~ temp_actual, data = analysis(split))
boot_models <- boots |>
  mutate(model = map(splits, fit_lm_on_bootstrap), coef_info = map(model, tidy))
boot_coefs <- boot_models |>
  unnest(coef_info) |>
  select(-std.error, -statistic, -p.value) |>
  pivot_wider(names_from = term, values_from = estimate)

bike_dsn <- tibble(
  temp_actual = c(50, 60, 70, 80),
  yhat = predict(bikes_lm, data.frame(temp_actual = c(50, 60, 70, 80))),
  resid_sd = sigma(bikes_lm)
)

p_beta <- ggplot(boot_coefs) +
  geom_abline(aes(slope = temp_actual, intercept = `(Intercept)`),
              color = "#0072B2", linewidth = 0.5, alpha = 0.4) +
  geom_point(data = bikes, aes(temp_actual, rides), color = "grey60") +
  labs(x = "Temperature", y = "No. rides", title = "Variability in the fitted line")

p_eps <- ggplot(data = bikes, aes(x = temp_actual, y = rides)) +
  geom_point(alpha = 0.4, size = 1.2, color = "#003300") +
  geom_lm(color = "#f15a31") +
  stat_slab(aes(x = temp_actual, ydist = "norm", arg1 = yhat, arg2 = resid_sd),
            data = bike_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  labs(x = "Temperature", y = "No. rides", title = "Variability around the line")

p_beta + p_eps
```

## R: point estimate

`predict()` quickly calculates $\widehat y$ for a given $x$ (or a
vector of $x$'s):

```{r}
#| echo: true
predict(bikes_lm, newdata = data.frame(temp_actual = 52))
predict(bikes_lm, newdata = data.frame(temp_actual = c(52, 62, 72, 82)))
```

## R: intervals

The `interval` argument specifies which type of interval you want:

```{r}
#| echo: true
predict(bikes_lm, newdata = data.frame(temp_actual = 52), interval = "confidence")
predict(bikes_lm, newdata = data.frame(temp_actual = 52), interval = "prediction")
```

Notice the prediction interval is ___________________ than the
confidence interval, exactly as the SE formulas predicted.

## R: confidence level and SEs

The `level` argument sets the confidence level (default 0.95):

```{r}
#| echo: true
predict(bikes_lm, newdata = data.frame(temp_actual = 52), interval = "confidence", level = .95)
predict(bikes_lm, newdata = data.frame(temp_actual = 52), interval = "prediction", level = .99)
```

Adding `se.fit = TRUE` returns the standard error needed for
"by hand" calculations:

```{r}
#| echo: true
predict(bikes_lm, newdata = data.frame(temp_actual = 52), se.fit = TRUE)
```

::: {.callout-tip}
## Try it

Using `$fit` and `$se.fit` above along with the appropriate $t^*$,
reconstruct the 95% confidence interval for the mean response at
`temp_actual = 52` by hand, and check it against the `predict()`
output.
:::
