---
title: "Stat230 Homework XX"
author: "Your Name"
format:
  pdf: 
    colorlinks: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(patchwork)
library(broom)
library(ggformula)
```

To submit your assignment, 

1. render your file to pdf
2. download the pdf to your computer (if using maize)
3. submit your pdf to gradescope in the appropriate submission space
    + Make sure to mark your pages through the gradescope system!


*Guidelines* 

  + Make sure your PDF is "clean": 
      - `echo: TRUE`, `warning: FALSE`, `message: FALSE`
      - Do not print long pages of output. Printing tibbles is OK
      - All questions should be answered with sentences written in text
  + Make sure your .qmd is "clean":
      - remove unnecessary code 

# Problem 1 

(a) Your written answer hee

```{r}
# your code here
```

(b) Your written answer here

```{r}
# your code here
```

(c) Your code here

```{r}
# your code here
```

# Problem 2

(a) Your written answer hee

```{r}
# your code here
```

(b) Your written answer here

```{r}
# your code here
```

(c) Your code here

```{r}
# your code here
```


# Problem 3

(a) Your written answer hee

```{r}
# your code here
```

(b) Your written answer here

```{r}
# your code here
```

(c) Your code here

```{r}
# your code here
```




