---
title: "Exam01 Practice Questions"
format:
    html: default
    pdf: default
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(ggformula)
theme_set(theme_minimal())
```

Practice questions for each skill on the [Exam01 Review](exam01-review.qmd) sheet. 

## 1. Express a SLR model using proper notation

*Example question* In the Day 2 activity, we explored the relationship between `audience` score and `critics` score for a sample of movies. Here is the scatterplot we looked at: 

```{r}
#| fig-width: 4
#| fig-height: 3

movies <- read.csv("https://aloy.github.io/stat230-materials/data/movie_scores.csv")
gf_point(audience ~ critics, data = movies) |>
    gf_lm()
```

(a) What is the response variable? What is the explanatory variable?

(b) Write out the *theoretical* SLR model for this relationship. Define $Y_i$ and $x_i$ in context, include all parameters, and state the assumption placed on the error term.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$, where $Y_i$ is the audience score for movie $i$, $x_i$ is the critics score for movie $i$, $\beta_0$ and $\beta_1$ are the (unknown) population intercept and slope, and $\varepsilon_i \overset{iid}{\sim} N(0, \sigma^2)$.

:::
:::

*Example question* The Day 3 activity fit a model relating leaf width (`Width`, in mm) to `Year` for a sample of 252 leaves, giving the coefficient estimates below.

| term | estimate |
|---|---|
| (Intercept) | 37.723 |
| Year | -0.018 |

Write out the *fitted* SLR model using proper notation.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$\widehat{\text{Width}}_i = 37.723 - 0.018 \times \text{Year}_i$

Note there's no error term in the fitted model -- $\widehat{Y}_i$ is a single predicted value, not a random variable, so there's nothing left to place a distributional assumption on.

:::
:::

## 2. Interpret the coefficients and $R^2$ of a model in context

```{r}
LeafWidth <- read_csv("https://aluby.github.io/stat230/data/LeafWidth.csv")
leaf_mod <- lm(Width ~ Year, data = LeafWidth)
```

*Example question* Using the fitted model ($\widehat{\text{Width}}_i = 37.723 - 0.018 \times \text{Year}_i$), interpret the slope and intercept in context.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

**Slope:** For each additional year, the model predicts an average decrease of 0.018 mm in leaf width.

**Intercept:** In year 0, the model predicts an average leaf width of 37.723 mm. This is an extreme extrapolation -- the data only cover a narrow, recent range of years -- so this interpretation isn't meaningful in context.

:::
:::


*Example question* For the `LeafWidth` model above, predicting `Width` from `Year` has $R^2 = 0.06$. Which of the following is the best interpretation of this value?



(a) For each additional year, leaf width increases by 0.6mm on average.
(b) About 6% of leafs have their width predicted exactly correctly by this model.
(c) About 6% of the variability in width among these leaves is explained by year.
(d) The correlation between width and year is 0.6.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

**(c)** -- $R^2$ is the proportion of variability in the response explained by the model. (Watch out for (d): $r = \sqrt{R^2}$ -- $R^2$ and $r$ are related but not equal.)

:::
:::

## 3. Check necessary conditions for a regression model and discuss the implications of a model deficiency

*Example question* For each statement below, circle TRUE or FALSE.

(a) TRUE or FALSE: We can check the linearity condition using a scatterplot of the response against the explanatory variable, before ever fitting a model.
(b) TRUE or FALSE: A normal Q-Q plot that curves away from the diagonal line in both tails suggests the residuals are more heavily-tailed than a Normal distribution.
(c) TRUE or FALSE: A residuals-vs-fitted-values plot that fans out (wider spread for larger fitted values) suggests a violation of the equal variance condition.
(d) TRUE or FALSE: If the independence condition is violated, we can typically detect it just by looking at a residuals-vs-fitted plot.
(e) TRUE or FALSE: If the linearity condition is badly violated, the estimated slope may be misleading, even if the p-value for the slope is small.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) TRUE (b) TRUE (c) TRUE (d) FALSE -- independence is usually about *how the data were collected* (e.g., time series, clustering), and often can't be verified from a residuals-vs-fitted plot; a residuals-vs-index plot can sometimes reveal it if the data have a meaningful order. (e) TRUE

:::
:::

*Example question* Suppose you fit a model predicting `income` from `education` for the Riverview data, and the residuals-vs-fitted plot shows a clear curved (U-shaped) pattern. Which of the following can you still trust about this model? Select all that apply, and explain your reasoning.

(a) The point estimate of the slope
(b) The p-value for the slope
(c) The standard error of the slope
(d) Predictions made from the model

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

None of these should be trusted. A curved residual pattern means the *linearity* condition is violated -- a straight line is the wrong shape for this relationship. That makes the point estimate itself misleading, not just the inference built on top of it, so the p-value, standard error, and any predictions are all compromised too.

:::
:::

## 4. Perform a hypothesis test to draw inference about a single regression coefficient

*Example question* The LeafWidth model gives the following output:

```{r}
broom::tidy(leaf_mod) |>
    knitr::kable(digits = 3)
```

On $n = 252$ data points. Conduct a hypothesis test for $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \ne 0$. State the test statistic, and describe (without necessarily calculating exactly) how the p-value could be found.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

$t = \dfrac{-0.018 - 0}{0.004} \approx -4.0$

The p-value is the two-sided tail area beyond $t = \pm 4.0$ under a $t$-distribution with $df = 250$: $P(T \le -4.0) + P(T \ge 4.0)$. Since $t \approx -4.0$ is far in the tail, this p-value will be very small (R reports it as $7.43\times 10^{-5}$), giving strong evidence that $\beta_1 \ne 0$.

:::
:::

*Example question* Before collecting the NHANES sitting-time data, a researcher hypothesizes that *more* sitting time is associated with a *higher* depression risk score -- not just a difference in either direction. The fitted model (`dep_score ~ sedentary_hours`) gives a slope of 0.0314 (SE = 0.0116) for `sedentary_hours`, with $df = 657$.

(a) State $H_0$ and $H_A$ in symbols, matching the researcher's directional hypothesis.
(b) Is the p-value R reports by default (from `summary()`) the correct one to use here? If not, explain how you would find the correct p-value. 

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 > 0$

(b) No -- R's default p-value is two-sided, but the researcher's hypothesis is one-sided. Since the test statistic ($t = 0.0314/0.0116 \approx 2.71$) is positive, in the direction of $H_A$, the correct one-sided p-value is *half* of R's reported two-sided p-value.

:::
:::

## 5. Construct and interpret confidence intervals for regression coefficients

*Example question* Recall Movies model: `lm(audience ~ critics)` gives a slope estimate of 0.519 (SE = 0.0345), with $df = 144$. 

(a) Using the `qt()` output below, report the correct $t^*$ value to use rounded to 3 decimal places

```{r}
qt(.01, df = 144, lower.tail = FALSE)
qt(.025, df = 144, lower.tail = FALSE)
qt(.05, df = 144, lower.tail = FALSE)
qt(.1, df = 144, lower.tail = FALSE)
```

(b) Construct and report a 90% CI for the slope, showing the formula with numbers plugged in.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution
(a) You should use `qt(.05, df = 144)`: 1.656

(b)

$0.519 \pm 1.656 \times 0.0345 \;\Longrightarrow\; (0.462, \; 0.576)$

:::
:::

*Example question* Recall Riverview model using *standardized* education: `income ~ scale(education)` gives a slope estimate of 11.567 (SE = 1.613), with $df = 30$. For a 95% confidence interval, use a critical value is $t^* = 2. 

It may also be helpful know that the `income` has a mean of 53.7 and a SD of 14.5. `education` has a mean of 16 and a SD of 4.36. 

(a) Construct a 95% CI for this slope.
(b) Interpret the interval in context.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) $11.567 \pm 2 \times 1.613 \;\Longrightarrow\; (8.83, \; 14.8)$

(b) We are 95% confident that for each one-SD increase in education (about 4.36 years), average income increases by between \$8,829 and \$14,305 (8.829 and 14.305 thousand dollars).

:::
:::

## 6. Use a regression model to make appropriate predictions and communicate the uncertainty of these predictions through an interval

*Example question* Using the movies model, we predict the audience score for a movie with a critics score of 75. The point estimate is 71.2, and R produces two 90% intervals: (50.4, 92.1) and (69.3, 73.1).

(a) Which interval is the confidence interval, and which is the prediction interval? How can you tell?
(b) If we wanted to predict the audience score for *Kingsman: The Secret Service*, which has a critics score of 75, which interval should we use?
(c) For the type of interval you did *not* choose in (b), what is the correct interpretation?

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) (69.3, 73.1) is the confidence interval -- it's narrower, since it only accounts for uncertainty in the *average* audience score at critics = 75. (50.4, 92.1) is the prediction interval -- it's wider, since it also accounts for the extra variability of one individual movie's score around that average.

(b) The prediction interval, (50.4, 92.1), since we're predicting a single new movie's score, not an average.

(c) The other inerval is the confidence interval. This gives the interval for the *average* audience score for all movies with a critics score of 75.
:::
:::

*Example question* A researcher wants to use the NHANES `dep_score ~ sedentary_hours` model to make a claim about people who sit 6 hours per day. Explain which interval (confidence or prediction) is appropriate for each scenario below, and briefly say why.

(a) "I want a range for the depression risk score of one specific person who sits 6 hours a day."
(b) "I want a range for the *average* depression risk score among everyone who sits 6 hours a day."

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) A prediction interval -- this is about one individual, whose depression risk score could be far from the average even if we knew the average perfectly.

(b) A confidence interval -- this is about a population average, which only requires accounting for uncertainty in *where the regression line is*, not individual-level variability. The confidence interval will be substantially narrower than the prediction interval for the same $x$ value.

:::
:::

## 7. Express an SLR model when transformations have been used on the response and/or the explanatory variables

*Example question* Instead of the raw PHQ-9 depression score, we actually used `dep_score`, defined as $\log(\text{PHQ-9} + 1)$. The fitted model is `dep_score ~ sedentary_hours`, with intercept 0.904 and slope 0.0314.

(a) Write out the fitted SLR equation, being careful to write it in terms of `dep_score`, not the raw PHQ-9 score.
(b) A student concludes: "for every one-hour increase in sitting time, the model predicts an average increase of 0.0314 points in PHQ-9 score." Explain in one sentence why this is incorrect, then give a corrected interpretation.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) $\widehat{\text{dep\_score}}_i = 0.904 + 0.0314 \times \text{sedentary\_hours}_i$, where $\text{dep\_score}_i = \log(\text{PHQ-9}_i + 1)$.

(b) The model was fit on the *log-transformed* PHQ-9 score, not the raw PHQ-9 score, so "0.0314" describes a change in $\log(\text{PHQ-9}+1)$, not a change in PHQ-9 itself -- the usual slope interpretation doesn't carry over once you've transformed the response. Corrected: "for every one-hour increase in sitting time, the model predicts an average increase of 0.0314 in depression risk score (the log-transformed PHQ-9 measure)."

:::
:::

*Example question* Suppose instead of transforming the response, we wanted to write out the general *theoretical* model for a log-transformed response, where $Y_i$ is the raw response and $x_i$ is the explanatory variable.

(a) Write the theoretical model using proper notation, and state the assumption on the error term.
(b) Give one reason (from what we discussed in class) why we might choose to log-transform a response variable before fitting a model.

::: {.content-visible when-format="html"}
::: {.callout-tip collapse="true"}
### Solution

(a) $\log(Y_i) = \beta_0 + \beta_1 x_i + \varepsilon_i$, where $\varepsilon_i \overset{iid}{\sim} N(0, \sigma^2)$. (Note: $\varepsilon_i$'s Normal/equal-variance assumptions now apply on the log scale, not the original scale of $Y$.)

(b) A log transformation can help address a right-skewed response and/or non-constant variance in the residuals (e.g., PHQ-9 scores were heavily right-skewed, with many zeros and a long tail of high scores) -- both violations of the SLR conditions on the original scale.

:::
:::
