---
title: "r-functions-reference"
format: 
  html:
    fig-width: 4
    fig-height: 2
---

```{r}
#| include: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

*Note:* This is a bare-bones functions reference for R code in Stat230. All examples use the `bikes` dataset from class, which lives in the {bayesrules} package. Most plotting functions use the {ggformula} package.  

```{r}
library(ggformula)
library(bayesrules)
library(dplyr)
library(ggplot2)
```

# Viewing data

Prints the first few rows: 

```{r}
head(bikes)
```

Get a quick overview of a data set (from {dplyr} package)

```{r}
library(dplyr)
glimpse(bikes)
```

# EDA

## Graphs

### Histogram

I like to specify `col="white"` to draw white lines between the bars. You can also customize the color of the bars with `fill`

```{r}
gf_histogram(~temp_actual, data = bikes, col = "white", fill = "darkgreen")
```

### Boxplot

One quantitative variable:

```{r}
gf_boxplot(~temp_actual, data = bikes, fill = "darkgreen")
```

One quantitative variable separated and colored by a categorical variable:

```{r}
gf_boxplot(temp_actual ~ weekend, data = bikes, fill = ~weekend)
```

### Scatterplot

```{r}
gf_point(rides ~ temp_actual, data = bikes)
```

Colored by a 3rd variable: 

```{r}
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes)
```

### QQ plot

```{r}
gf_qq(~temp_actual, data = bikes) |>
  gf_qqline()
```

## Numeric Summaries

Syntax is `data$variable_name`. 

Mean: 

```{r}
mean(bikes$temp_actual)
```

Standard deviation:

```{r}
sd(bikes$temp_actual)
```

Median: 

```{r}
median(bikes$temp_actual)
```

Variance

```{r}
var(bikes$temp_actual)
```

Correlation (2 quantitative variables)

```{r}
cor(bikes$temp_actual, bikes$temp_feel)
```

# SLR 

## Fitting the model

To get the slope and intercept, use `lm()`

```{r}
lm(rides ~ temp_actual, data = bikes)
```

To get the summary, first save the `lm` with `<-` and then call `summary`

```{r}
bikes_slr <- lm(rides ~ temp_actual, data = bikes)
summary(bikes_slr)
```

If you want to access specific elements of the summary (like $R^2$ or $\sigma$) you can use: 

```{r}
summary(bikes_slr)$sigma
summary(bikes_slr)$r.squared
```

## Plotting

Overlay a regression line with a pipe (`|>`) and `gf_lm()`

```{r}
gf_point(rides ~ temp_actual, data = bikes) |>
  gf_lm()
```

You can also specify a prediction or confidence interval in `gf_lm`:

```{r}
gf_point(rides ~ temp_actual, data = bikes) |>
  gf_lm(interval = "confidence", level = .9)
```

```{r}
gf_point(rides ~ temp_actual, data = bikes) |>
  gf_lm(interval = "confidence", level = .9) |>
  gf_lm(interval = "prediction", level = .9, fill = "goldenrod")
```

## Predictions

Use `predict()` to get a point estimate for $\hat{y}_i$, a confidence interval or prediction interval. 

```{r}
predict(bikes_slr, newdata = data.frame(temp_actual = 75))
```

```{r}
predict(bikes_slr, newdata = data.frame(temp_actual = 75),
        interval = "confidence", 
        level = .9)
```

```{r}
predict(bikes_slr, newdata = data.frame(temp_actual = 75),
        interval = "prediction", 
        level = .9)
```

## Inference

Confidence intervals:

```{r}
confint(bikes_slr)
```

Nice version of regression table:

```{r}
library(broom)
tidy(bikes_slr)
```

## Conditions

Use `augment` to create an augmented dataset with residuals and fitted values for all data points:

```{r}
bikes_slr_aug <- augment(bikes_slr)
head(bikes_slr_aug)
```

You can then make a histogram or scatterplot with the usual {ggformula} tools: 

```{r}
gf_point(.resid ~ .fitted, data = bikes_slr_aug)
gf_histogram(~.resid , data = bikes_slr_aug, col = "white")
gf_qq(~.resid , data = bikes_slr_aug) |>
  gf_qqline()
```

Alternatively, you can make a panel of residual plots with the `resid_panel` shortcut function in {ggResidpanel} package: 

```{r}
library(ggResidpanel)
resid_panel(bikes_slr)
```

(If it takes a long time, try using standardized residuals instead)

```{r}
resid_panel(bikes_slr, type = "standardized")
```

# MLR


# Other R 

## Finding $t^*$

Corresponds to the 2.5th percentile of a $t_{498}$ distribution:

```{r}
qt(.025, df = 498)
```

