---
title: "R Functions reference for Stat230"
author: "Amanda Luby & Noah Lee"
format: 
  html:
    fig-width: 4
    fig-height: 2
---

```{r}
#| include: false
ggplot2::theme_set(ggplot2::theme_minimal())
```

*Note:* This is a bare-bones functions reference for R code in Stat230. All examples use the `bikes` dataset from class, which lives in the {bayesrules} package. Most plotting functions use the {ggformula} package.  

```{r}
library(ggformula)
library(bayesrules)
library(dplyr)
library(ggplot2)
```

# Viewing data

Prints the first few rows: 

```{r}
head(bikes)
```

Get a quick overview of a data set (from {dplyr} package)

```{r}
library(dplyr)
glimpse(bikes)
```

# EDA

## Graphs

### Histogram

I like to specify `col="white"` to draw white lines between the bars. You can also customize the color of the bars with `fill`

```{r}
gf_histogram(~temp_actual, data = bikes, col = "white", fill = "darkgreen")
```

### Boxplot

One quantitative variable:

```{r}
gf_boxplot(~temp_actual, data = bikes, fill = "darkgreen")
```

One quantitative variable separated and colored by a categorical variable:

```{r}
gf_boxplot(temp_actual ~ weekend, data = bikes, fill = ~weekend)
```

### Scatterplot

```{r}
gf_point(rides ~ temp_actual, data = bikes)
```

Colored by a 3rd variable: 

```{r}
gf_point(rides ~ temp_actual, color = ~weekend, data = bikes)
```

### QQ plot

```{r}
gf_qq(~temp_actual, data = bikes) |>
  gf_qqline()
```

## Numeric Summaries

Syntax is `data$variable_name`. 

Mean: 

```{r}
mean(bikes$temp_actual)
```

Standard deviation:

```{r}
sd(bikes$temp_actual)
```

Median: 

```{r}
median(bikes$temp_actual)
```

Variance

```{r}
var(bikes$temp_actual)
```

Correlation (2 quantitative variables)

```{r}
cor(bikes$temp_actual, bikes$temp_feel)
```

Correlation Matrix (only on numeric columns)

```{r}
bikes |>
  select(temp_actual, temp_feel, humidity, windspeed, rides) |>
  cor()
```

Pairs plot

```{r, message=FALSE, cache=TRUE}
library(GGally)
bikes |>
  select(temp_actual, temp_feel, humidity, windspeed, rides) |>
  ggpairs()
```

# SLR 

## Fitting the model

To get the slope and intercept, use `lm()`

```{r}
lm(rides ~ temp_actual, data = bikes)
```

To get the summary, first save the `lm` with `<-` and then call `summary`

```{r}
bikes_slr <- lm(rides ~ temp_actual, data = bikes)
summary(bikes_slr)
```

If you want to access specific elements of the summary (like $R^2$ or $\sigma$) you can use: 

```{r}
summary(bikes_slr)$sigma
summary(bikes_slr)$r.squared
```

## Plotting

Overlay a regression line with a pipe (`|>`) and `gf_lm()`

```{r}
gf_point(rides ~ temp_actual, data = bikes) |>
  gf_lm()
```

You can also specify a prediction or confidence interval in `gf_lm`:

```{r}
gf_point(rides ~ temp_actual, data = bikes) |>
  gf_lm(interval = "confidence", level = .9)
```

```{r}
gf_point(rides ~ temp_actual, data = bikes) |>
  gf_lm(interval = "confidence", level = .9) |>
  gf_lm(interval = "prediction", level = .9, fill = "goldenrod")
```

## Predictions

Use `predict()` to get a point estimate for $\hat{y}_i$, a confidence interval or prediction interval. 

```{r}
predict(bikes_slr, newdata = data.frame(temp_actual = 75))
```

```{r}
predict(bikes_slr, newdata = data.frame(temp_actual = 75),
        interval = "confidence", 
        level = .9)
```

```{r}
predict(bikes_slr, newdata = data.frame(temp_actual = 75),
        interval = "prediction", 
        level = .9)
```

## Inference

Confidence intervals:

```{r}
confint(bikes_slr)
```

Nice version of regression table:

```{r}
library(broom)
tidy(bikes_slr)
```

## Conditions

Use `augment` to create an augmented dataset with residuals and fitted values for all data points:

```{r}
bikes_slr_aug <- augment(bikes_slr)
head(bikes_slr_aug)
```

You can then make a histogram or scatterplot with the usual {ggformula} tools: 

```{r}
gf_point(.resid ~ .fitted, data = bikes_slr_aug)
gf_histogram(~.resid , data = bikes_slr_aug, col = "white")
gf_qq(~.resid , data = bikes_slr_aug) |>
  gf_qqline()
```

Alternatively, you can make a panel of residual plots with the `resid_panel` shortcut function in {ggResidpanel} package: 

```{r}
library(ggResidpanel)
resid_panel(bikes_slr)
```

(If it takes a long time, try using standardized residuals instead)

```{r}
resid_panel(bikes_slr, type = "standardized")
```

# MLR

## F-test

Sequential anova:
```{r}
bikes_mlr_full <- lm(rides ~ temp_actual + windspeed + weekend, data = bikes)
anova(bikes_mlr_full)
```
Comparing two models:
```{r}
bikes_mlr_red <- lm(rides ~ temp_actual, data = bikes)
anova(bikes_mlr_full, bikes_mlr_red)
```


# Logistic Regression

```{r}
#| include: false
diabetes <- read.csv("https://raw.githubusercontent.com/plotly/datasets/master/diabetes.csv")
diabetes <- diabetes %>% 
  select(-DiabetesPedigreeFunction, -Insulin) %>%
  filter(
    Glucose>0,
    BloodPressure>0,
    SkinThickness>0,
    BMI>0.0,
    Age>0
  )  %>%
  mutate(
    pregnant = ifelse(Pregnancies > 2, 1, 0),
    obese = ifelse(BMI > 30.0, 1, 0)
  ) %>%
  select(-Pregnancies)
```

## Logistic EDA

### Plotting against binned predictor

Plotting Log Odds Against Binned Predictor

```{r}
diabetes %>%
  mutate(bin = ntile(Glucose, 20)) %>%
  group_by(bin) %>%
  summarize(
    mean_x = mean(Glucose),
    prob = mean(Outcome),
    odds = prob/(1-prob),
    logodds = log(odds)
  ) %>%
  ggplot(aes(x=mean_x, y = logodds)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, linetype="dashed")
```

### Bar graph

Looking at one categorical variable on the response:

```{r, warning=FALSE}
gf_bar(~factor(Outcome), fill = ~factor(obese), data=diabetes) +
  labs(title = "Diabetes Count by Obese and Pregnant", fill = "Obese") +
  coord_flip()
```

Make it a proportion:

```{r, warning=FALSE}
gf_bar(~factor(Outcome), fill = ~factor(obese), position="fill", data=diabetes) +
  labs(title = "Diabetes Count by Obese and Pregnant", fill = "Obese") +
  coord_flip()
```

Interaction between categorical variables

```{r, warning=FALSE}
gf_bar(~factor(Outcome), fill = ~factor(obese), position="fill", data=diabetes) +
  facet_wrap(~factor(pregnant), 
           labeller = labeller(`factor(pregnant)` = c("0" = "Not Pregnant", "1" = "Pregnant"))) +
  labs(title = "Diabetes Count by Obese and Pregnant", fill = "Obese") +
  coord_flip()
```

Boxplot to compare quantitative variable to the response:

```{r}
gf_boxplot(BMI ~ factor(Outcome), fill = ~factor(obese), data = diabetes)
```

## Fitting the model (binary data)

```{r}
diabetes_glm <- glm(Outcome~BMI+pregnant*obese, family=binomial, data=diabetes)
```


## Residual Checks

### Pearson Residuals

```{r}
diabetes_aug_p <- augment(diabetes_glm, 
                          type.predict= "link", 
                          type.residuals = "pearson") |>
  mutate(.resp.resid = resid(diabetes_glm, type= "response"))
```

Or 

```{r}
resid_panel(diabetes_glm, 
            type = "pearson", 
            plots = c("resid", "qq"), 
            smoother = TRUE)
resid_xpanel(diabetes_glm, 
             type = "pearson", 
             smoother = TRUE)
```


### Deviance Residuals

```{r}
diabetes_aug_d <- augment(diabetes_glm, 
                          type.predict= "link", 
                          type.residuals = "pearson") |>
  mutate(.resp.resid = resid(diabetes_glm, type= "response"))
```

Or

```{r}
resid_panel(diabetes_glm, 
            type = "deviance", 
            plots = c("resid", "qq"), 
            smoother = TRUE)
resid_xpanel(diabetes_glm, 
             type = "deviance", 
             smoother = TRUE)
```

### Binned Residuals

```{r}
diabetes_aug_p %>%
  mutate(bin = ntile(.fitted, 20)) %>%
  group_by(bin) %>%
  summarize(
    mean_fit = mean(.fitted),
    mean_resid = mean(.resid)
  ) %>% 
  ggplot(aes(x=mean_fit, y=mean_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dotted")
```


# Case Study tips

Compact coefficient table: 

```{r}
tidy(bikes_slr)
```

Compact coefficient table, formatted (k)nicely: 

```{r}
tidy(bikes_slr) |>
  knitr::kable(digits = 2)
```

Coefficient graph: 

```{r}
broom::tidy(bikes_slr, conf.int = TRUE, conf.level = .9) |>
  gf_pointrange(estimate + conf.low + conf.high ~ term) |>
  gf_hline(yintercept = 0, linetype = "dashed", col = "gray") +
  coord_flip()
```


For other case-study specific tips: 

  - [Case Study 2](https://aluby.github.io/stat230/activities/18-case-studies.html)
  - [Case Study 3](https://aluby.github.io/stat230/activities/24-case-studies.html)

# Other R 

## Finding $t^*$

Corresponds to the 2.5th percentile of a $t_{498}$ distribution:

```{r}
qt(.025, df = 498)
```

