---
title: "Day 12 notes: instructor answer key"
subtitle: "Filled-in version of notes/notes12.qmd (not for students)"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
library(broom)
theme_set(theme_minimal(base_size = 12))
source(here::here("notes", "notes-helpers.R"))

fev <- read.table("https://aloy.github.io/stat230-materials/data/FEVdataAge10to19.txt",
                   sep = "\t", header = TRUE) |>
  mutate(Smoke = as.logical(Smoke))

fev_lm <- lm(FEV ~ Age * Smoke, data = fev)
V <- vcov(fev_lm)
b <- coef(fev_lm)
df_resid <- fev_lm$df.residual

r3 <- function(x) formatC(x, format = "f", digits = 3)
r4 <- function(x) formatC(x, format = "f", digits = 4)

# Example 1: smoker slope, gamma = b1 + b3
g1 <- b[2] + b[4]
se1_naive <- sqrt(V[2,2] + V[4,4])
se1 <- sqrt(V[2,2] + V[4,4] + 2 * V[2,4])
t90 <- qt(0.95, df_resid)
t1 <- g1 / se1
p1 <- 2 * pt(-abs(t1), df_resid)

# Your turn: difference at age 15, gamma = b2 + 15 * b3
cc <- c(0, 0, 1, 15)
g2 <- sum(cc * b)
se2 <- sqrt(drop(t(cc) %*% V %*% cc))
t95 <- qt(0.975, df_resid)
t2 <- g2 / se2
p2 <- 2 * pt(-abs(t2), df_resid)
```

<style>
.ans { color: #b00020; font-weight: 600; }
</style>

Answers are in <span class="ans">red</span>. Every number is computed from the model, so they match the notes' R output.

## FEV data: the model

$$\mu\{FEV \mid Age, Smoke\} = \beta_0 + \beta_1\, Age + \beta_2\, Smoke + \beta_3\, Age \times Smoke$$

```{r}
kable_coef(fev_lm, terms = c("Intercept", "Age", "SmokeTRUE", "Age:SmokeTRUE"))
```

**Mean function for each group:**

Nonsmokers ($Smoke = 0$):
$\mu\{FEV \mid Age\} =$ <span class="ans">$\beta_0 + \beta_1\, Age$</span>
$\;\approx$ <span class="ans">`r r3(b[1])` + `r r3(b[2])` Age</span>

Smokers ($Smoke = 1$):
$\mu\{FEV \mid Age\} =$ <span class="ans">$(\beta_0 + \beta_2) + (\beta_1 + \beta_3)\, Age$</span>
$\;\approx$ <span class="ans">`r r3(b[1] + b[3])` + `r r3(b[2] + b[4])` Age</span>

## Warm up: what's missing from the table?

The slope for smokers is $\beta_1 + \beta_3$.

**Why can't we just add $SE(\widehat\beta_1) + SE(\widehat\beta_3)$?**

<span class="ans">Two reasons. (1) Even for independent estimates, **variances** add, not standard deviations: $\sqrt{a^2 + b^2} < a + b$, so adding SEs overstates the uncertainty. (2) Here the estimates are not independent. They come from the same fitted model and are negatively correlated (correlation of about `r r3(V[2,4] / sqrt(V[2,2] * V[4,4]))`), which pulls the true SE of the sum down even further.</span>

**Why can't we use the p-value for `Age:SmokeTRUE` to test whether the smoker slope is zero?**

<span class="ans">That row tests $H_0: \beta_3 = 0$, i.e. whether the smoker slope **differs from the nonsmoker slope**. The smoker slope being zero is $H_0: \beta_1 + \beta_3 = 0$, a different hypothesis about a different quantity.</span>

## Standard errors for linear combinations

A **linear combination** of coefficients has the form $\gamma = c_1 \beta_i + c_2 \beta_j$.

**Estimate:** $\widehat\gamma =$ <span class="ans">$c_1 \widehat\beta_i + c_2 \widehat\beta_j$</span>

**Standard error:**

$$SE(\widehat\gamma) = \sqrt{c_1^2 \{SE(\widehat\beta_i)\}^2 + c_2^2 \{SE(\widehat\beta_j)\}^2 + 2 c_1 c_2 \, Cov(\widehat\beta_i, \widehat\beta_j)}$$

## The covariance matrix of the coefficients

```{r}
#| echo: true
vcov(fev_lm)
```

Check: $SE(\widehat\beta_1) =$ <span class="ans">`r r3(sqrt(V[2,2]))`</span>
(from the coefficient table), so
$\{SE(\widehat\beta_1)\}^2 =$ <span class="ans">`r format(V[2,2], digits = 4)`</span>,
which matches the (Age, Age) entry of the matrix.

## Example 1: CI for the smoker slope

**Target:** $\gamma = \beta_1 + \beta_3$. What are $c_1$ and $c_3$?
<span class="ans">$c_1 = 1$, $c_3 = 1$</span>

**Estimate:**
<span class="ans">$\widehat\gamma = \widehat\beta_1 + \widehat\beta_3 = `r r3(b[2])` + (`r r3(b[4])`) = `r r4(g1)`$</span>

**Standard error:**

<span class="ans">
$$SE(\widehat\gamma) = \sqrt{(1)^2(`r format(V[2,2], digits = 4)`) + (1)^2(`r format(V[4,4], digits = 4)`) + 2(1)(1)(`r format(V[2,4], digits = 4)`)} = \sqrt{`r format(se1^2, digits = 4)`} = `r r4(se1)`$$
</span>

**90% CI** ($df = 341$, $t^* \approx 1.649$):

<span class="ans">$`r r4(g1)` \pm `r r3(t90)`(`r r4(se1)`) = (`r r3(g1 - t90 * se1)`,\ `r r3(g1 + t90 * se1)`)$</span>

**Interpretation:**

<span class="ans">We are 90% confident that, among smokers aged 10-19, each additional year of age is associated with an increase in mean FEV between about `r r3(g1 - t90 * se1)` and `r r3(g1 + t90 * se1)` L/s.</span>

::: {.callout-note}
## What if we ignored the covariance?

Using only $\sqrt{\{SE(\widehat\beta_1)\}^2 + \{SE(\widehat\beta_3)\}^2}$ would
give `r r3(se1_naive)`. The correct SE is `r r3(se1)`. Here the covariance is
**negative**, so ignoring it makes the interval about `r round(100 * (se1_naive / se1 - 1))`% too wide.
:::

## Testing a linear combination

Same four steps as a test for a single coefficient. Only the estimate and the SE change.

**Example:** Is the smoker slope different from zero?

<span class="ans">
$H_0: \beta_1 + \beta_3 = 0$ vs. $H_a: \beta_1 + \beta_3 \ne 0$

$t = \dfrac{`r r4(g1)` - 0}{`r r4(se1)`} = `r round(t1, 2)`$ on $df = 341$

p-value $= 2\,P(t_{341} > |`r round(t1, 2)`|) \approx `r r3(p1)`$

Using $\alpha = 0.05$ we fail to reject $H_0$: there is only weak evidence that mean FEV changes with age among smokers. (This agrees with the 90% CI barely excluding 0. A 95% CI would include 0.)
</span>

## Your turn: smokers vs. nonsmokers at age 15

**1.** Mean function at $Age = 15$, then subtract (smoker $-$ nonsmoker):

<span class="ans">
Smokers: $(\beta_0 + \beta_2) + (\beta_1 + \beta_3)(15)$

Nonsmokers: $\beta_0 + \beta_1(15)$

$\gamma = (\beta_0 + \beta_2) + 15\beta_1 + 15\beta_3 - \beta_0 - 15\beta_1 = \beta_2 + 15\beta_3$
</span>

**2.** <span class="ans">$c_2 = 1$ $\qquad c_3 = 15$</span>
(and $c_0 = c_1 = 0$, since $\beta_0$ and $\beta_1$ cancel).

**3.** Standard error:

<span class="ans">
$$SE(\widehat\gamma) = \sqrt{(1)^2 \{SE(\widehat\beta_2)\}^2 + (15)^2 \{SE(\widehat\beta_3)\}^2 + 2(1)(15)\,Cov(\widehat\beta_2, \widehat\beta_3)}$$
$$= \sqrt{`r format(V[3,3], digits = 4)` + 225(`r format(V[4,4], digits = 4)`) + 30(`r format(V[3,4], digits = 4)`)} = \sqrt{`r format(se2^2, digits = 4)`} = `r r3(se2)`$$
</span>

Common mistake: forgetting to square the 15 on the variance term while using it unsquared on the covariance term (the covariance term has $c_2 c_3 = 15$).

**4.** 95% CI ($df = 341$, $t^* \approx 1.967$):

<span class="ans">$`r r3(g2)` \pm `r r3(t95)`(`r r3(se2)`) = (`r r3(g2 - t95 * se2)`,\ `r r3(g2 + t95 * se2)`)$, where $\widehat\gamma = `r r3(b[3])` + 15(`r r3(b[4])`) = `r r3(g2)`$</span>

**5.** Test $H_0: \gamma = 0$ vs. $H_a: \gamma \ne 0$:

<span class="ans">$t = \dfrac{`r r3(g2)`}{`r r3(se2)`} = `r round(t2, 2)`$ on 341 df, p-value $\approx `r format(p2, digits = 2)`$. Reject $H_0$ at the 5% level.</span>

**6.** Interpretation:

<span class="ans">At age 15, smokers' mean FEV is estimated to be about `r r3(abs(g2))` L/s lower than nonsmokers' mean FEV. We are 95% confident the true difference is between `r r3(abs(g2 + t95 * se2))` and `r r3(abs(g2 - t95 * se2))` L/s lower. The difference is statistically significant (p $\approx$ `r format(p2, digits = 2)`).</span>

## Doing it in R

```{r}
#| echo: true
cc <- c(0, 0, 1, 15)   # coefficients are (Intercept), Age, SmokeTRUE, Age:SmokeTRUE

est <- sum(cc * coef(fev_lm))
se <- sqrt(drop(t(cc) %*% vcov(fev_lm) %*% cc))

c(estimate = est, se = se)
```

```{r}
#| echo: true
tstar <- qt(0.975, df = fev_lm$df.residual)
est + c(-1, 1) * tstar * se          # 95% CI

2 * pt(-abs(est / se), df = fev_lm$df.residual)   # two-sided p-value
```

These match the hand calculation above.

## A shortcut: re-center the model

```{r}
#| echo: true
fev <- mutate(fev, Age15 = Age - 15)
lm(FEV ~ Age15 * Smoke, data = fev) |> tidy()
```

The `SmokeTRUE` row (estimate `r r3(g2)`, SE `r r3(se2)`, p `r format(p2, digits = 2)`) is exactly the age-15 difference from Your turn.

## Summary

| To do inference for... | Use | Notes |
|:--|:--|:--|
| a single coefficient $\beta_j$ | $t$-test / CI from the coefficient table | Day 11 |
| a linear combination $c_1\beta_i + c_2\beta_j$ | $t$-test / CI with $SE = \sqrt{\mathbf{c}^T \mathbf{V} \mathbf{c}}$ | today |
| several coefficients at once ($\beta_2 = \beta_3 = 0$) | extra sums of squares $F$-test | Day 11 |
